# Distance-based methods {#sec-distance-methods}

```{r}
#| include: false
#| cache: false
source("before-each-chapter.R")
```

Distance-based anomaly detection methods are based on the distances between observations. The pairwise distances between observations are computed, and anomalies are identified as those observations that are far from other observations.

Before we discuss some of the methods that fall into this class of algorithms, we will first discuss some of the different ways of measuring distances between observations.

## Multivariate pairwise distances {#sec-distances}

Suppose our observations are denoted by $\bm{y}_1,\dots,\bm{y}_n$. In univariate data, the distance between any two observations is simply the absolute difference between the points. But in multivariate data, there are many different ways of measuring distances between observations, even when the data are all numerical.

### Euclidean distance

Perhaps the most obvious way to define the distance between any two points $\bm{y}_i$ and $\bm{y}_j$ is to measure the length of the straight line between the points. This is commonly known as Euclidean distance, named after the famous Greek mathematician Euclid, and can be defined as
$$\|\bm{y}_i - \bm{y}_j\|_2 = \sqrt{\sum_{k=1}^m (y_{ik}-y_{jk})^2}.$$
This is also known as the $L_2$ distance. For $m=1$ or $m=2$, this is the physical distance between the points when plotted on a strip plot or a scatterplot (provided there is no jittering used).

The `dist()` function will return a distance matrix containing all pairwise distances between observations. The argument `method` specifies the type of distance to compute, with the default being Euclidean distance.

Here is an example using only the first five observations of the `old_faithful` data set (omitting the time stamp). Because the distances are symmetric, only the lower triangle of the matrix is computed.

```{r}
#| label: dist-1
#| code-fold: false
head(oldfaithful, 5)
of <- head(oldfaithful, 5) |> select(duration, waiting)
of |> dist()
```

### Manhattan distance

The Manhattan distance (or absolute distance) between two observations is given by
$$\|\bm{y}_i - \bm{y}_j\|_1 = \sum_{k=1}^m |y_{ik}-y_{jk}|.$$
This is also known as the $L_1$ distance. It is called the Manhattan distance as it gives the shortest path between the corners of city blocks (denoted as points on a grid) when those blocks are rectangular, as they mostly are in Manhattan. For the same reason, it is also sometimes called the "taxicab" distance or the "city block" distance.

Let's compute the Manhattan distances for the same five observations from the `oldfaithful` data set.

```{r}
#| label: dist-2
#| code-fold: false
of |> dist(method = "manhattan")
```

### Minkowski distance

This generalises the Manhattan and Euclidean distances to use powers of $p$ to define the $L_p$ distance:
$$\|\bm{y}_i - \bm{y}_j\|_p = \left(\sum_{k=1}^m (y_{ik}-y_{jk})^p\right)^{1/p}.$$
It is named after the German mathematician, Hermann Minkowski. When $p=1$ this is the Manhattan distance, and when $p=2$ this is the Euclidean distance.

Here is the Minkowski distance with $p = 3$ for the same five observations from the `oldfaithful` data set.

```{r}
#| label: dist-3
#| code-fold: false
of |> dist(method = "minkowski", p = 3)
```

### Chebyshev Distance

The maximum distance between any components of $\bm{y}_i$ and $\bm{y}_j$ is known as the Chebyshev distance (after the Russian mathematician, Pafnuty Chebyshev), given by
$$\|\bm{y}_i - \bm{y}_j\|_{\text{max}} = \max_k |y_{ik}-y_{jk}|.$$

```{r}
#| label: dist-4
#| code-fold: false
of |> dist(method = "maximum")
```

This is also known as chessboard distance, as it is the number of moves a king would have to travel on a chessboard to move between two squares. The Chebyshev distance is equivalent to the Minkowski distance with $p=\infty$.

Notice that all four of the distances introduced so far are equal to the absolute difference between observations when $m=1$. The next distance does not have this property.

### Mahalanobis distance

When the variables have different scales, the variables with the largest ranges will dominate the distance measures. In the Old Faithful example, durations are much longer than waiting times, and so the duration variable is dominating the calculation of distances. Consequently, it is often preferable to scale the data before computing distances.

Suppose we scale the data using the robust multivariate approach described in @sec-scaling, so that the scaled data are given by
$\bm{z}_i = \bm{U} (\bm{y}_i - \bm{m})$, where $\bm{m}$ is the pointwise median, and $\bm{U}^\intercal\bm{U} = \bm{S}_{\text{OGK}}^{-1}$ is the Cholesky decomposition of the inverse of $\bm{S}_{\text{OGK}}$, a robust estimate of the covariance matrix of the data. Then the Euclidean distance between $\bm{z}_i$ and $\bm{z}_j$ is given by
$$
  \| \bm{z}_i - \bm{z}_j\|_2 = \sqrt{(\bm{y}_i - \bm{y}_j)^\intercal \bm{U}^\intercal \bm{U} (\bm{y}_i - \bm{y}_j)} = \sqrt{(\bm{y}_i - \bm{y}_j)^\intercal \bm{S}_{\text{OGK}}^{-1} (\bm{y}_i - \bm{y}_j)}
$$
This is known as the Mahalanobis distance, named after the Indian statistician Prasanta Chandra Mahalanobis (although he wasn't using a robust measure of covariance). It is a multivariate generalisation of the number of standard deviations between any two observations.

Mahalanobis distance is not an option provided by the `dist()` function, so we will need to compute the scaled data first.

```{r}
#| label: zscores
#| code-fold: false
z <- oldfaithful |>
  select(duration, waiting) |>
  mvscale()
z |>
  head(5) |>
  dist()
```

For more information about pairwise distances, see @Borg2005.

## Nearest neighbours

If there are $n$ observations, then there are $n(n-1)/2$ pairwise distances to compute, so this is an $O(n^2)$ operation which can take a long time for large $n$.

Some algorithms only compute the pairwise distances of the $k$ nearest observations, although finding those observations requires some additional distances to be computed. For some types of distances, efficient solutions are available using kd trees [@Bentley1975;@Arya1998] that find the $k$ nearest neighbours to each observation in $O(n\log(n))$ time.

The calculation of $k$ nearest neighbours is useful for more than anomaly detection problems. It is also the basis of a popular classification method due to the Berkeley statisticians Evelyn Fix and Joe Hodges [@knn] which is often known as the "kNN algorithm".

Suppose we use the Old Faithful data to find eruptions that are neighbours in the (duration, waiting) space. The `dbscan` package uses kd trees to quickly identify the $k$ nearest observations to each eruption. We will use the scaled data computed in the last section to find the 5 nearest neighbours to each eruption.

```{r}
#| label: knn
#| code-fold: false
#| dependson: zscores
# Find 5 nearest neighbours to each eruption
knn <- dbscan::kNN(z, k = 5)
# First eruption in the data set
oldfaithful[1, ]
# Five closest observations
oldfaithful[knn$id[1, ], ]
```

For large data sets, approximations are available which speed up the computation even more, but are less accurate in finding the $k$ nearest neighbours. The `approx` argument specifies a distance tolerance which makes the process faster for large data sets, although the neighbours returned may not be the exact nearest neighbours.

```{r}
#| label: ann
#| code-fold: false
#| dependson: zscores
# Find 5 approximate nearest neighbours to each eruption
kann <- dbscan::kNN(z, k = 5, approx = 2)
# Five closest observations
oldfaithful[kann$id[1, ], ]
```

In this case, four of the same five observations have been found, with the last being different.

## Local outlier factors {#sec-lof}

A popular way of using $k$-nearest neighbours for anomaly detection is via local outlier factors [@lof]. This is designed to find points in areas with few surrounding observations.

Suppose we write the distance between observations $\bm{y}_i$ and $\bm{y}_j$ as $\|\bm{y}_i - \bm{y}_j\|$, and let $d_{i,k}$ be the distance between observation $\bm{y}_i$ and its $k$th nearest neighbour.

Let $N_{i,k}$ be the set of $k$ nearest neighbours within $d_{i,k}$ of $\bm{y}_i$. (If there are multiple observations all exactly $d_{i,k}$ from $\bm{y}_i$, then $N_{i,k}$ may contain more than $k$ observations, but we will ignore that issue here.)

Note that an observation $\bm{y}_j$ may be within $N_{i,k}$ while $\bm{y}_i$ does not fall within $N_{j,k}$, as shown in the diagram below.

```{r}
#| label: fig-njk
#| fig-cap: "A synthetic data set of 11 observations. Nearest neighbourhoods containing five observations each are shown for $\\bm{y}_1$ (in orange) and $\\bm{y}_2$ (in blue). The neighbourhood of $\\bm{y}_1$ contains $\\bm{y}_2$ in its nearest five observations, but $\\bm{y}_2$ does not include $\\bm{y}_1$ in its nearest five observations."
#| warning: false
#| echo: false
set.seed(2)
df <- tibble(
  x = c(rnorm(5), rnorm(6, 5, 1)),
  y = c(rnorm(5), rnorm(6, 5, 1))
 ) |>
  mutate(d = dbscan::kNNdist(pick(x, y), 5))
chosen <- c(3, 11)
df |>
  ggplot() +
  ggforce::geom_circle(
    aes(x0 = x, y0 = y, r = d, col = c("N15", "N25"), fill = c("N15", "N25")),
    alpha = 0.2,
    data = df[chosen, ]
  ) +
  geom_point(aes(x = x, y = y), size = 3) +
  annotate(
    "text",
    x = df$x[chosen] + 0.3,
    y = df$y[chosen] + 0.3,
    label = latex2exp::TeX(
      paste0("$y_", 1:2, "$"),
      bold = TRUE,
      output = "character"
    ),
    parse = TRUE,
    col = discrete_colours[1:2]
  ) +
  annotate(
    "text",
    x = c(-0.5, 3.6),
    y = c(4.2, 6),
    label = latex2exp::TeX(paste0("$N_{", 1:2, ",5}$"), output = "character"),
    parse = TRUE,
    col = discrete_colours[1:2]
  ) +
  geom_segment(
    aes(x = x, y = y, xend = c(7.75, 8.75), yend = y),
    arrow = arrow(length = unit(0.02, "npc")),
    data = df[chosen, ],
    col = discrete_colours[1:2]
  ) +
  annotate(
    "text",
    x = c(4.7, 7),
    y = c(-0.8, 4.1),
    label = latex2exp::TeX(paste0("$d_{", 1:2, ",5}$"), output = "character"),
    parse = TRUE,
    col = discrete_colours[1:2]
  ) +
  coord_fixed(xlim = c(-4, 8.5), ylim = range(df$y)) +
  guides(col = "none", fill = "none") +
  theme_void()
```

Let $r_k(i,j) = \max(d_{j,k}, \|\bm{y}_i-\bm{y}_j\|)$ be the "reachability" of $\bm{y}_i$ from $\bm{y}_j$. Thus, $r_k(i,j)$ is the distance between observations $\bm{y}_i$ and $\bm{y}_j$ when they are far from each other, but is equal to $d_{j,k}$ if $\bm{y}_i$ is one of the $k$ nearest neighbours of $\bm{y}_j$. The reachability is a truncated variant of the usual Euclidean distance $\|\bm{y}_i-\bm{y}_j\|$ so that it is not less than $d_{j,k}$.

The average reachability of $\bm{y}_i$ *from* its nearest neighbours is given by
$$
  \bar{r}_{i,k} = k^{-1} \sum_{\bm{y}_j \in N_{i,k}} r_k(i,j)
$$
This is not the same as the average reachability of the neighbours *from* $\bm{y}_i$ which, by definition, would be $d_{i,k}$. An observation will have high average reachability if it is far from its neighbours, whereas it will have low average reachability if it has many close neighbours.

Then the local outlier factor for observation $\bm{y}_i$ is given by
$$
\ell_{i,k} = \frac{\bar{r}_{i,k}}{k}\sum_{\bm{y}_j \in N_{i,k}} \bar{r}^{-1}_{j,k}.
$$
Thus, it is the product of $\bar{r}_{i,k}$ and the average value of $\bar{r}^{-1}_{j,k}$ for points within its neighbourhood.

This provides a relative measure of the density of each observation compared to its neighbours. An observation is regarded as an anomaly if it is in a low-density region (far from its neighbours) while its neighbours are in higher density regions (with many neighbours nearby). A value of $\ell_{i,k}$ much greater than 1 shows that the observation has much larger average reachability compared to its neighbours, and so is regarded as an anomaly.

One problem with this definition is that $\ell_{i,k}$ can be infinite. If the data set contains at least $k+1$ identical observations, then the average reachability $\bar{r}_{j,k}$ will be 0 if $\bm{y}_j$ is one of the group of identical observations. Consequently, $\ell_{i,k} = \infty$ if $\bm{y}_i$ is a neighbour to one of the group of identical observations. In this book we use a slightly different definition from that used elsewhere and replace these infinite values with 1s (indicating they have similar reachability to their identical neighbours).

Another problem is that it can be difficult to determine an appropriate value for $k$.

### Example: Old Faithful data

Let's apply the idea to the `duration` variable in the Old Faithful data set. We will use the `lof_scores()` function to compute the local outlier factors for each eruption, with $k=100$ nearest neighbours. The value of $k=100$ has been chosen after experimenting with various values.

```{r}
#| label: ofscores
of_scores <- oldfaithful |>
  mutate(lof = lof_scores(duration, k = 100))
of_scores |> arrange(desc(lof))
```

Here the three shortest eruptions (which we have previously identified as genuine anomalies) are those with the highest local outlier factors. Next there is a group of eruptions with durations above 300 seconds.

```{r}
#| label: fig-ofscores
#| fig-cap: Local outlier factors for each eruption duration plotted against the duration.
#| dependson: ofscores
of_scores |>
  ggplot(aes(x = duration, y = lof)) +
  geom_point() +
  labs(x = "Duration (seconds)", y = "Local outlier factor")
```

@fig-ofscores shows a scatterplot of the local outlier factors for each eruption plotted against the eruption duration. The three short eruptions clearly stand out. The bimodal distribution of eruption durations is also evident here, with the peak in local outlier factors around 150 seconds, indicating a low density of eruptions at that duration.

## Detect deviating cells algorithm

So far, we have treated each multivariate observation as a single point in the multivariate space, and have computed distances between those points. Under this perspective, the observations are the rows of the data matrix, and the columns are the variables. So an anomaly is a row that is far from other rows. That may be because of the values in one or more columns, or the unusual combination of values across columns. Almost all robust multivariate methods require at least half the observations to be non-anomalous, so that the main body of the data can be identified. But sometimes many rows (observations) each have a few contaminated cell values, and it is not possible to identify "normal" rows. In that case, it may be possible to identify the anomalous cells instead of the anomalous rows. This is the idea behind the "detect deviating cells" (DDC) algorithm of @Rousseeuw2016-wm.

Suppose a fraction $\varepsilon$ of cells are contaminated, and that the contaminated cells are randomly distributed across the data matrix. Then, on average, each row will have $\varepsilon m$ contaminated cells. The expected fraction of contaminated rows is $1-(1-\varepsilon)^m$, which is close to 1 when $m\varepsilon$ is large.

Rather than treating each row as a point, the DDC algorithm treats each cell in the data matrix as a point, and detects anomalous cells. It does this by checking whether each cell value is consistent with what would be predicted from the other variables in the same row (observation), following these steps.

1. Non-numerical columns, and discrete columns taking a few distinct values, are omitted.
2. Optionally, numerical columns can be transformed to be closer to normality, for example by applying a log transformation to positive-valued columns.
3. All numerical columns are standardised using robust location and scale estimates.
4. Cell values that are outliers in their column are flagged.
5. Robust pairwise correlations are computed between all pairs of columns. Only pairs of columns whose absolute correlation exceeds a threshold (0.5 by default) are considered "connected"; for each such pair, a robust bivariate slope is also estimated.
6. Each cell value is predicted using a weighted combination of connected columns, ignoring flagged cells. The weights are based on the absolute correlations between the columns.
7. Cells for which the observed value differs substantially from the predicted value are flagged as anomalous.

Entire rows that contain an unusual accumulation of large cell residuals can additionally be flagged as rowwise outliers.

This method is implemented in the `cellWise::DDC()` function. Here we apply it to the French male mortality data, with the data matrix containing the mortality rates for each year (rows) and age (columns).

```{r}
#| label: ddc-mx
#| output: false
mx_wide <- fr_mortality |>
  filter(Sex == "Male") |>
  select(-Sex) |>
  tidyr::pivot_wider(names_from = Age, values_from = Mortality) |>
  arrange(Year)
mx_ddc <- cellWise::DDC(mx_wide |> select(-Year))$stdResid |>
  as_tibble() |>
  mutate(Year = mx_wide$Year) |>
  tidyr::pivot_longer(-Year, names_to = "Age", values_to = "stdResid") |>
  mutate(
    Age = as.numeric(Age),
    anomaly = cut(
      stdResid,
      breaks = qnorm(c(0, 0.001, 0.01, 0.99, 0.999, 1))
    )
  )
```

The standardised residuals from the DDC algorithm are plotted as a heatmap in @fig-ddc-mx, with red indicating large positive residuals (higher than expected mortality), blue indicating large negative residuals (lower than expected mortality), and grey indicating small residuals. The matrix has been rotated anticlockwise in the plot, so that the years are on the horizontal axis and the ages are on the vertical axis, to be consistent with the custom of plotting time on the horizontal axis.

```{r}
#| label: fig-ddc-mx
#| fig-cap: "DDC algorithm applied to the French male mortality data. The standardised residuals from the DDC algorithm are plotted as a heatmap, with red indicating large positive residuals (higher than expected mortality), and blue indicating large negative residuals (lower than expected mortality)."
#| message: false
ggplot(mx_ddc) +
  geom_tile(aes(x = Year, y = Age, fill = anomaly)) +
  scale_fill_brewer(palette = "RdBu", direction = -1) +
  coord_cartesian(expand = FALSE) +
  guides(fill = "none")
```

The DDC algorithm identifies many anomalous cells in the data, particularly during years in which France was involved in war:

* 1853--1856: Crimean war
* 1870: Franco-Prussian war
* 1871: Repression of the ‘Commune de Paris’
* 1914--1918: World War I
* 1940--1944: World War II

Mortality rates are often also high in the year after a war, as soldiers return home with injuries and illnesses. The Spanish flu pandemic of 1918--1919 is evident in the plot, with higher than usual mortality rates for a wider age range than the wars. There was also unusually high mortality for males in their 20s in 1840--1843, which does not correspond to any known war [@french_mortality], but seems very similar to the pattern due to the Crimean War thirteen years later. The same pattern is not seen for females. Unusually high childhood mortality at around 10 years of age is seen until about 1870, probably driven by infectious diseases such as cholera, tuberculosis and measles. Finally, the large blue area at the upper right of the plot indicates unusually low mortality rates for older ages in recent years.


## BACON algorithm {#sec-bacon}

Francis Bacon wrote:

> "Whoever knows the ways of Nature will more easily notice her deviations; and, on the other hand, whoever knows her deviations will more accurately describe her ways." [@Bacon1620, Book II, Aphorism 29]

Inspired by this aphorism, @Billor2000 developed the BACON algorithm for multivariate outlier detection. The name also stands for "Blocked Adaptive Computationally efficient Outlier Nominator". The idea is to identify a clean subset of the data that is free of outliers, known as the "basic subset", and then use that subset to estimate the centre and spread of the data, and identify outliers as those observations that are far from that centre. At each iteration, every observation whose Mahalanobis distance from the current basic subset falls below a data-adaptive threshold is admitted simultaneously. This process typically converges in a few iterations regardless of sample size. The algorithm is implemented in the `robustX::BACON()` function, which is designed to handle both multivariate data and regression data. We will discuss the regression version in @sec-bacon-reg.

Given an $n \times p$ data matrix with observations $\bm{y}_1, \dots, \bm{y}_n$, let $B$ denote the current basic subset, and let $\bar{\bm{y}}_B$ and $\bm{S}_B$ be the mean vector and covariance matrix of the observations in $B$.

**Step 1: Initialise the basic subset**. Choose an initial $B$ of size $m = cp$, where $c$ is a small integer. Several initialisation strategies are available. The default in `BACON()` is to compute the Mahalanobis distances from the full-data mean $\bar{\bm{y}}$ and covariance $\bm{S}$, and select the $m$ observations with the smallest distances. An alternative initialisation strategy is to compute distances from the univariate medians rather than the full-data mean. This is more robust to outliers, but the nominated outliers can change if the data are rotated or rescaled. Another strategy is to select $m$ observations at random. Because the initial subset is not anchored to the data centre, this can be more effective when the data are multimodal, but results may vary across runs.

**Step 2: Compute discrepancies**. For each observation $\bm{y}_i$, compute the Mahalanobis distance from the basic subset:
$$
  d_i(\bar{\bm{y}}_B, \bm{S}_B) = \sqrt{(\bm{y}_i - \bar{\bm{y}}_B)^\intercal \bm{S}_B^{-1} (\bm{y}_i - \bar{\bm{y}}_B)}.
$$

**Step 3: Expand the basic subset**. Update $B$ by admitting every observation with
$$
  d_i(\bar{\bm{y}}_B, \bm{S}_B) < c_{npr}\sqrt{\chi^2_{p,1-\alpha}},
$$
where $\chi^2_{p,1-\alpha}$ is the $(1 - \alpha)$-quantile of the chi-squared distribution with $p$ degrees of freedom, and $c_{npr} = c_{np} + c_{hr}$ is a correction factor, where
\begin{align*}
  c_{np} &= 1 + \frac{p+1}{n-p} + \frac{2}{n-1-3p} \\
  c_{hr} &= \max\{0, (h-r)/(h+r)\},
\end{align*}
$h = \lfloor(n + p + 1)/2\rfloor$, and $r$ is the number of observations in the current basic subset $B$. The correction factor $c_{npr}$ inflates the threshold when the current basic subset of size $r$ is small relative to the target size $h$. This is because a small basic subset selected for being close to the data centre will have artificially compressed spread, so the Mahalanobis distances computed from it need to be inflated to something closer to what they would be from a full clean sample. As $r$ grows toward $h$, $c_{hr} \to 0$.

**Step 4: Iterate** Steps 2 and 3 until the size of $B$ no longer changes.

**Step 5: Nominate outliers.** All observations excluded from the final $B$ are nominated as outliers.

At each step the algorithm only needs to compare $n$ distances against a single threshold, an $O(n)$ operation. By adding multiple observations at each iteration, BACON eliminates the need for many iterations, making it practical for data sets of millions of observations.

A regression variant of BACON also exists, in which the Mahalanobis distances are replaced by scaled prediction errors from the current least-squares fit, and the chi-squared threshold is replaced by an appropriate quantile of the t-distribution. In R, the function `robustX::BACON()` handles both the multivariate and regression cases via the optional `y` argument.

### Example: Old Faithful data

We apply `robustX::BACON()` to the `oldfaithful` data. Because the Mahalanobis distance is used in the algorithm, there is no need to scale the data first. The default initialisation strategy is used, which selects the $m = cp = 4 \times 2 = 8$ observations with the smallest Mahalanobis distances from the full-data mean and covariance.

```{r}
#| label: bacon-of
#| code-fold: false
bacon <- oldfaithful |>
  select(duration, waiting) |>
  robustX::BACON()
```

The output shows the number of observations in each basic subset over the iterations. The algorithm converges in `r english::words(bacon$steps)`, with the final basic subset containing `r sum(bacon$subset)` observations. The remaining `r NROW(oldfaithful) - sum(bacon$subset)` observations are nominated as outliers. The result contains a Mahalanobis distance in `$dis` for each observation, and a logical vector `$subset` that is `TRUE` for observations belonging to the final basic subset (the non-outliers). We attach these to the data and examine the nominated outliers.

```{r}
#| label: bacon-results
#| code-fold: false
of_bacon <- oldfaithful |>
  mutate(
    distance = bacon$dis,
    outlier = !bacon$subset
  )
of_bacon |>
  filter(outlier) |>
  arrange(desc(distance))
```

As expected, the nominated outliers include the very short eruptions that we have identified as genuine anomalies throughout this chapter: their Mahalanobis distances from the robust centre of the data are substantially larger than those of other eruptions. However, many other observations have also been nominated as outliers.

```{r}
#| label: fig-bacon-of
#| fig-cap: "BACON outliers for the Old Faithful eruptions. All observations in the lower cluster have been identified as outliers, while all observations in the upper cluster have been identified as non-outliers. The three short eruptions are among the outliers, but so are many other observations."
#| dependson: bacon-results
of_bacon |>
  ggplot(aes(x = duration, y = waiting, colour = outlier)) +
  geom_point() +
  labs(x = "Duration", y = "Waiting")
```

@fig-bacon-of displays which outliers are identified by the BACON algorithm. It has identified the cluster of longer durations and waiting times as non-outliers, and all the shorter durations and small waiting times as outliers. Clearly, this is not a good result, as the shorter durations are not anomalous, but rather represent a second cluster of eruptions. By starting with a very small central initial basic subset, BACON has been misled by the bimodality of the data, and has identified one of the modes as outliers.

For comparison, if we start with a randomly selected initial basic subset, then the algorithm usually converges to a different (and more reasonable) solution, with far fewer outliers, because its initial subset will probably include observations from each cluster.

```{r}
#| label: fig-bacon-v2
#| code-fold: false
#| fig-cap: "BACON outliers for the Old Faithful eruptions, starting with a random initial basic subset."
set.seed(1)
bacon2 <- oldfaithful |>
  select(duration, waiting) |>
  robustX::BACON(init.sel = "random")
oldfaithful |>
  mutate(Outlier = !bacon2$subset) |>
  ggplot(aes(x = duration, y = waiting, colour = Outlier)) +
  geom_point() +
  labs(x = "Duration", y = "Waiting")
```


## BACON algorithm for linear regression {#sec-bacon-reg}

Suppose we observe a response $y_i$ and a vector of $\bm{x}_i$, and assume that the non-outlying observations follow the model
$$
  y_i = \bm{x}_i^\intercal \bm{\beta} + \varepsilon_i,
$$
where the errors $\{\varepsilon_1,\dots,\varepsilon_n\}$ are iid with mean zero and common variance $\sigma^2$. Here $\bm{x}_i$ may include a leading 1 if an intercept is included in the model. Let $p$ be the length of $\bm{x}_i$, including the intercept if it is present.

The model can be written in matrix form as $\bm{y} = \bm{X}\bm{\beta} + \bm{\varepsilon}$, where $\bm{y}$ is the $n$-vector of responses, $\bm{X}$ is the $n\times p$ matrix of predictors, and $\bm{\varepsilon}$ is the $n$-vector of errors. Provided $\bm{X}$ is of rank $p$, the model can be estimated using ordinary least squares regression [@seberlee2003], resulting in the estimate
$$
  \hat{\bm{\beta}} = (\bm{X}^\intercal\bm{X})^{-1}\bm{X}^\intercal\bm{y}.
$$
The fitted values (i.e., predicted values for the training data) are given by
$$
  \hat{\bm{y}} = \bm{X}\hat{\bm{\beta}} = \bm{H}\bm{y},
$$
where $\bm{H} = \bm{X}(\bm{X}^\intercal\bm{X})^{-1}\bm{X}^\intercal$ is known as the "hat"-matrix because it creates the "y-hat" values $\hat{\bm{y}}$ from the data $\bm{y}$.

The diagonals of $\bm{H}$, given by $h_1,\dots,h_n$, take values between 0 and 1. These are known as the "leverage" values [@faraway2004linear, p. 69], and measure how much each observation influences the corresponding fitted value. High leverage values (close to 1) correspond to observations that have a large influence on the estimated coefficients, and so leaving those observations out will lead to very different values for the fitted values and residuals. On the other hand, small leverage values (close to 0) correspond to observations that have little influence on the estimated coefficients, and so leaving those observations out will lead to similar values for the fitted values and residuals. Often high leverage values are those corresponding to anomalies in the predictor space.

The residuals from the model are the differences between what was observed and what was predicted, given by
$$
  \bm{e} = \bm{y} - \hat{\bm{y}} = (\bm{I} - \bm{H})\bm{y}.
$$ {#eq-residual-distribution}
Often, we need standardised residuals, which are obtained by dividing each residual by its estimated standard deviation, giving
$$
  r_i = \frac{e_i}{\hat{\sigma}\sqrt{1-h_i}}, \qquad i = 1,\dots, n,
$$
where
$$
\hat\sigma^2 = \frac{1}{n-p}\sum_{i=1}^n e_i^2
$$ {#eq-residual-variance}
is the estimated residual variance.

A linear model can be estimated in R using the `stats::lm()` function. The `broom::augment()` function will compute the residuals (named `.resid`), the standardised residuals (named `.std.resid`), and the leverage values (named `.hat`).

In the regression version of the BACON algorithm, the goal is to identify observations that are inconsistent with the regression model. This is more complicated than the multivariate case discussed in @sec-bacon, because an observation can be anomalous in terms of its predictor values, or in terms of its response value given its predictor values, or both. Also, high leverage values can lead to residual anomalies being masked. Masking occurs when the presence of multiple outliers distorts the fitted model enough that the outliers appear to have small residuals. The regression BACON algorithm therefore begins by using the multivariate BACON algorithm on the predictors, and then uses scaled prediction errors to decide which observations are compatible with the regression model.

Let $B$ be the current basic subset, with $r$ observations, and let $\bm{X}_B$ and $\bm{y}_B$ denote the corresponding predictor matrix and response vector. Let $\hat{\bm{\beta}}_B$ be the least-squares estimate computed using only the observations in $B$, and let
$$
  \hat{\sigma}_B^2 = (r-p)^{-1}\sum_{i\in B}(y_i - \bm{x}_i^\intercal\hat{\bm{\beta}}_B)^2.
$$
We also define
$$
  h_i(B) = \bm{x}_i^\intercal(\bm{X}_B^\intercal\bm{X}_B)^{-1}\bm{x}_i
$$
as the leverage of observation $i$ with respect to the basic subset $B$. The regression discrepancy for observation $i$ is the absolute scaled residual or prediction error
$$
  |t_i(B)| =
  \begin{cases}
    \displaystyle
    \left|\frac{y_i - \bm{x}_i^\intercal\hat{\bm{\beta}}_B}{\hat{\sigma}_B\sqrt{1-h_i(B)}}\right|, & i \in B, \\
    \displaystyle
    \left|\frac{y_i - \bm{x}_i^\intercal\hat{\bm{\beta}}_B}{\hat{\sigma}_B\sqrt{1+h_i(B)}}\right|, & i \notin B.
  \end{cases}
$$
For observations already in the basic subset this is the usual standardised residual. For observations outside the basic subset it is a scaled prediction error. The sign difference arises because $1-h_i(B)$ is the variance deflation for an in-sample residual, while $1+h_i(B)$ is the corresponding variance inflation for an out-of-sample prediction error; an observation not used to fit the model contributes additional uncertainty beyond that of the residual variance alone.

The algorithm then proceeds as follows.

**Step 1: Initialise the basic subset.** Apply the multivariate BACON algorithm (@sec-bacon) to the predictor variables, ignoring the response. This identifies observations with central predictor values and protects the first regression fit from high-leverage points. These observations are then used to construct a full-rank initial regression subset, which is grown using the smallest scaled prediction errors until it reaches approximately $m=cp$, where $c$ is usually 4 or 5. Larger values of $c$ may be needed if the predictor variables have many repeated values, as then the resulting initial predictor matrix $\bm{X}_B$ may be of low rank.

**Step 2: Fit the regression model.** Fit least squares to the current basic subset $B$, obtaining $\hat{\bm{\beta}}_B$ and $\hat{\sigma}_B$.

**Step 3: Compute discrepancies.** Compute $|t_i(B)|$ for every observation, using the formula above.

**Step 4: Expand the basic subset.** Update $B$ by admitting every observation with
$$
  |t_i(B)| < \text{t}_{r-p,\,1-\alpha/(2(r+1))},
$$
where $\text{t}_{r-p,\,1-\alpha/(2(r+1))}$ is the $1-\alpha/(2(r+1))$ quantile of a t-distribution with $r-p$ degrees of freedom. Equivalently, this is the upper $\alpha/(2(r+1))$ tail quantile, as computed in R by `qt(alpha / (2 * (r + 1)), r - p, lower.tail = FALSE)`.

**Step 5: Iterate** Steps 2--4 until the size of $B$ no longer changes.

**Step 6: Nominate outliers.** Observations excluded from the final basic subset are nominated as regression outliers.

As in the multivariate version, the algorithm grows the basic subset in blocks, rather than adding one observation at a time. It gives a robust regression fit at relatively low computational cost. The final values $|t_i(B)|$ provide anomaly scores: large values correspond to observations whose responses are unusually far from what would be predicted by the regression relationship fitted to the final basic subset. This contrasts with the ordinary residuals shown in @fig-shiraz-residuals, which are measured relative to a line fitted to all observations and may therefore be compressed near outliers that have distorted the fit.

### Example: Shiraz reviews

Consider the wine reviews of Shiraz (aka Syrah), plotted in @fig-shiraz. We can fit a linear regression model to these data, using the points awarded to each wine as a predictor, and the log prices as the response variable.

```{r}
#| label: shiraz-data
#| eval: false
#| code-fold: false
wine_reviews <- fetch_wine_reviews()
```

```{r}
#| label: shiraz-regression
#| code-fold: false
shiraz <- wine_reviews |>
  filter(
    variety %in% c("Shiraz", "Syrah"),
    !is.na(price),
    !is.na(points),
    price > 0
  )
fit_wine <- lm(log(price) ~ points, data = shiraz)
summary(fit_wine)
```

The fitted model can be written as
$$
  \log(\text{Price})  = `r sprintf("%.3f",coef(fit_wine)[1])` + `r sprintf("%.3f",coef(fit_wine)[2])` \times \text{Points} + \varepsilon
$$
where $\varepsilon$ is a random error term with mean zero and variance $`r sprintf("%.3f",sigma(fit_wine))`^2$. The fitted regression line is shown in @fig-shiraz-regression.

```{r}
#| label: fig-shiraz-regression
#| warning: false
#| fig-cap: Log price of Shiraz as a function of points. The points are horizontally jitted to reduce overplotting.
wine_aug <- fit_wine |>
  broom::augment(data = shiraz)
wine_aug |>
  ggplot(aes(y = price, x = points)) +
  geom_jitter(height = 0, width = 0.1, alpha = 0.5, colour = "#999999") +
  geom_line(aes(y = exp(.fitted))) +
  scale_y_log10()
```

The scaled residuals from this model are shown in @fig-shiraz-residuals. Wines with large positive residuals are more expensive than expected for their points score, while wines with large negative residuals are cheaper than expected. However, these residuals are computed from a least-squares line fitted to all observations, including any anomalously expensive or cheap wines.

```{r}
#| label: fig-shiraz-residuals
#| fig-cap: Standardised residuals from a least-squares regression of log price on points for Shiraz wines.
wine_aug |>
  ggplot(aes(x = points, y = .std.resid)) +
  geom_jitter(height = 0, width = 0.1, alpha = 0.5, colour = "#999999") +
  geom_hline(yintercept = 0) +
  labs(x = "Points", y = "Standardised residuals")
```

```{r}
#| label: shiraz-points
#| echo: false
shiraz_points <- shiraz |> count(points) |> arrange(desc(n))
```

We now apply regression BACON, using `points` as the predictor and `log(price)` as the response. The function first applies multivariate BACON to the predictor variable, and then applies the regression version of the algorithm using scaled prediction errors. Because the predictor has many repeated values (taking only integer values between 80 and 100), we need to specify the initial set to be larger than the most common points score, which is `r shiraz_points$points[1]` with `r shiraz_points$n[1]` observations. So we set the initial size to $m=600$ observations to ensure the initial basic subset does not contain only observations with a single points score. The BACON algorithm then identifies outliers as those with large scaled prediction errors from the fitted regression line based on the final basic subset.

```{r}
#| label: shiraz-bacon
#| code-fold: false
bacon_alpha <- 0.05
bacon_wine <- robustX::BACON(
  x = shiraz$points,
  y = log(shiraz$price),
  m = 600,
  alpha = bacon_alpha,
  verbose = FALSE
)
wine_bacon <- wine_aug |>
  mutate(
    bacon_distance = bacon_wine$tis,
    bacon_outlier = !bacon_wine$subset,
    bacon_x_outlier = !bacon_wine$mv.subset
  )
tibble(
  observations = nrow(wine_bacon),
  predictor_outliers = sum(wine_bacon$bacon_x_outlier),
  regression_outliers = sum(wine_bacon$bacon_outlier),
  mv_steps = bacon_wine$steps["mv"],
  regression_steps = bacon_wine$steps["lm"]
)
```

No predictor outliers are identified by the multivariate BACON algorithm, as all observations have points scores that are close to those of many other observations. However, `r english::words(sum(wine_bacon$bacon_outlier))` regression outliers are identified by the regression BACON algorithm, as their prices are inconsistent with the fitted price--points relationship.

```{r}
#| label: bacon-check2
#| include: false
stopifnot(sum(wine_bacon$bacon_x_outlier) == 0)
```

The largest BACON distances are shown below. These correspond to wines with prices that are most unusual after accounting for their points score.

```{r}
#| label: shiraz-bacon-largest
#| code-fold: false
wine_bacon |>
  filter(bacon_outlier) |>
  arrange(desc(bacon_distance)) |>
  select(country, winery, year, points, price, bacon_distance)
```

They are all over-priced wine (under this model), with the most anomalous being a 2009 Shiraz from the Henschke winery in the Eden Valley region of South Australia.

To see the effect of the anomalies on the fitted regression relationship, we refit the least-squares model using only the final BACON basic subset. This gives the regression line shown in @fig-shiraz-bacon that is obtained after omitting the nominated regression outliers. The two lines are almost identical because the nominated outliers have relatively low leverage.

```{r}
#| label: shiraz-bacon-fit
#| dependson: shiraz-bacon
#| echo: true
#| code-fold: false
fit_bacon_wine <- lm(log(price) ~ points, data = wine_bacon, subset = !bacon_outlier)
```

```{r}
#| label: fig-shiraz-bacon
#| fig-cap: "Regression BACON applied to Shiraz wine reviews. Points in red are nominated regression outliers. The dashed line is the ordinary least-squares fit using all observations, while the solid line is the least-squares fit after omitting BACON-nominated outliers."
#| warning: false
#| dependson: shiraz-bacon-fit
wine_grid <- tibble(points = range(wine_bacon$points)) |>
  mutate(
    least_squares = exp(predict(fit_wine, newdata = pick(everything()))),
    bacon = exp(predict(fit_bacon_wine, newdata = pick(everything())))
  )
wine_bacon |>
  ggplot(aes(y = price, x = points)) +
  geom_jitter(aes(colour = bacon_outlier), height = 0, width = 0.1, alpha = 0.6) +
  geom_line(
    data = wine_grid,
    aes(x = points, y = least_squares, linetype = "Least squares"),
    inherit.aes = FALSE
  ) +
  geom_line(
    data = wine_grid,
    aes(x = points, y = bacon, linetype = "BACON subset"),
    inherit.aes = FALSE
  ) +
  scale_y_log10() +
  scale_colour_manual(values = c("FALSE" = "#999999", "TRUE" = discrete_colours[1])) +
  scale_linetype_manual(
    values = c("Least squares" = "dashed", "BACON subset" = "solid")
  ) +
  labs(x = "Points", y = "Price", linetype = "Fit") +
  guides(colour = "none")
```

Finally, @fig-shiraz-bacon-distance shows the absolute scaled prediction errors used by the regression BACON algorithm. The dashed horizontal line is the final $t$ cutoff. Observations above this line are those excluded from the final basic subset.

```{r}
#| label: fig-shiraz-bacon-distance
#| fig-cap: "Absolute scaled prediction errors from regression BACON for Shiraz wine reviews. The dashed line shows the final cutoff used to nominate regression outliers."
#| dependson: shiraz-bacon-fit
regression_p <- length(coef(fit_bacon_wine))
regression_r <- sum(bacon_wine$subset)
bacon_wine_limit <- qt(
  bacon_alpha / (2 * (regression_r + 1)),
  df = regression_r - regression_p,
  lower.tail = FALSE
)
wine_bacon |>
  ggplot(aes(x = points, y = bacon_distance, colour = bacon_outlier)) +
  geom_jitter(height = 0, width = 0.1, alpha = 0.6) +
  geom_hline(yintercept = bacon_wine_limit, linetype = "dashed") +
  scale_colour_manual(values = c("FALSE" = "#999999", "TRUE" = discrete_colours[1])) +
  labs(x = "Points", y = "Absolute scaled prediction error") +
  guides(colour = "none")
```

The comparison with the ordinary residual plot highlights the purpose of the regression BACON algorithm. Least-squares residuals are measured relative to a line fitted to all observations, so the line may itself be affected by the anomalies. BACON instead repeatedly refits the model to a growing subset of observations that are internally consistent, and then nominates observations that remain inconsistent with that subset.

## Isolation forest

Isolation tree: Randomly partition data until every point is isolated or some depth is reached.

Combine many isolation trees to form an isolation forest.

Anomaly score inversely proportional to average depth of isolation trees.

isotree package

@Liu2008-dp
