# High-dimensional methods {#sec-highdim}

```{r}
#| include: false
#| cache: false
source("before-each-chapter.R")
```

## Projections



## Principal component analysis

## Multi-dimensional scaling

## Stahel-Donoho outlyingness

Stahel-Donoho outlyingness is a robust projection-based measure of how far each
observation is from the main body of the data [@Stahel81;@Donoho81].

For any unit
vector $\bm{a}$, project the data to one dimension via $\bm{a}^\intercal\bm{y}_i$, then
standardise the distance from a robust centre using a robust scale estimate. The
outlyingness of $\bm{y}_i$ is the largest standardised distance over all directions,
so an observation is flagged when it looks extreme in at least one projection.

This approach is useful in multivariate settings because it is affine equivariant and
less sensitive to masking than classical Mahalanobis-distance methods. In practice,
the optimisation over all directions is approximated using many candidate
projections, and skewness-adjusted variants are often preferred [@Brys2005-pc]. In
R, `robustbase::adjOutlyingness()` provides an efficient implementation for computing
these scores.

## Stray and HDoutliers algorithms

The HDoutliers algorithm of @Wilkinson2017-rb was designed to address several problems in anomaly detection:

1. Data containing a mixture of categorical and continuous variables.
2. Data with high dimensionality (a large number of variables).
3. Data with a large number of observations.
4. Data where anomalies occur in small clusters, rather than as isolated points.

```{r}
#| label: fig-stray
of_scores <- oldfaithful |>
  mutate(stray = stray_scores(duration))
of_scores |> arrange(desc(stray))
of_scores |>
  ggplot(aes(x = duration, y = stray)) +
  geom_point()
```

@Talagala2019-zv, @Wilkinson2017-rb

## PCout

@Filzmoser2008-qa, @Filzmoser2013-tc

* mvoutlier::pcout

## FastPCS

@Vakili2014-yc

* FastPCS::FastPCS

## DOBIN

@Kandanaarachchi2020-lo

## Comparisons using the O3 plot
