# High-dimensional methods {#sec-highdim}

```{r}
#| label: setup
#| include: false
#| cache: false
source("before-each-chapter.R")
# Set up for using rgl
options(rgl.useNULL = TRUE)
rgl::setupKnitr()
```

Suppose we have observations $\bm{y}_1, \ldots, \bm{y}_n$, where each observation is of length $m$. We can think of each observation as a point in $m$-dimensional space. Our data can be represented as an $n \times m$ matrix $\bm{Y}$, where each row is an observation and each column is a variable. In this chapter, we will focus on the case where $m$ is large, which we will refer to as "high-dimensional" data.

The curse of dimensionality (@sec-curse) makes it difficult to detect anomalies once the number of variables grows too large. In practice, even a moderate number of variables (e.g., $m > 5$) can cause problems for some methods. The high-dimensional methods described in this chapter are designed to address these challenges, by reducing the dimensionality of the data while preserving the relevant structure for anomaly detection.

## Projections

The simplest way to reduce the dimensionality of the data is to use **linear projections**: linear combinations of the original variables that map each $m$-dimensional observation to a lower-dimensional space. Different projection directions preserve different aspects of the joint structure, so the choice of direction matters. Understanding the geometry of projections sets the foundation for the methods described in the rest of this chapter.

To keep the geometry simple, we will scale all variables to have mean zero and unit variance, although to preserve interpretability, no rotation will be applied. The ideas are illustrated using air quality and meteorological variables measured hourly at the Aotizhongxin monitoring station (at the Olympic Sports Centre) in Beijing, over May 2014.

```{r}
#| label: aotizhongxin
#| code-fold: false
az_aq <- air_quality |>
  filter(station == "Aotizhongxin", year == 2014, month == 5) |>
  select(pm2_5:rainfall, wind_speed) |>
  mvscale(center = mean, scale = sd, cov = NULL) |>
  tidyr::drop_na()
az_aq
```

### Projecting from 2 to 1 dimension

The simplest case maps two variables to a single number. A projection onto a vector $\bm{a} = (a_1, a_2)^\intercal$ gives a score
$$
w_i = \bm{a}^\intercal \bm{y}_i = a_1 y_{i,1} + a_2 y_{i,2}
$$
for each observation $\bm{y}_i$, where $\bm{a}$ is a unit vector (i.e., $\|\bm{a}\| = \sqrt{a_1^2 + a_2^2} = 1$). The projection is a linear transformation that maps the 2D data cloud onto the line spanned by $\bm{a}$, with $w_i$ giving the position of the projected point along that line.

When we write this in matrix form, the projection of the $n \times 2$ data matrix $\bm{Y}$ onto the vector $\bm{a}$ gives an $n$-vector of scores $\bm{w} = \bm{Y}\bm{a}$.

@fig-proj-geometry shows an example using $\bm{a} = (0.8, 0.6)^\intercal$ applied to the standardised carbon monoxide (`co`) and nitrogen dioxide (`no2`) measurements at Aotizhongxin. The blue line is the projection direction. The five orange points are randomly selected observations, showing how they project onto the line. The resulting scores are the blue dots on the line.

```{r}
#| label: fig-proj-geometry
#| code-fold: true
#| fig-cap: "Projection of standardised carbon monoxide and nitrogen dioxide from Aotizhongxin onto one dimension. The blue line shows the projection direction $\\bm{a}$. Five randomly selected observations (in orange) are shown projecting onto the line, with the resulting scores shown as blue dots."
#| dependson: aotizhongxin
set.seed(4)
# The projection direction
a <- c(0.8, 0.6)
# Select five random observations to illustrate the projection
show_pts <- az_aq |>
  slice_sample(n = 5) |>
  mutate(
    w = a[1] * co + a[2] * no2,
    foot_co = w * a[1],
    foot_no2 = w * a[2]
  )
# Construct the plot
ggplot(az_aq, aes(x = co, y = no2)) +
  geom_point(alpha = 0.2) +
  geom_abline(slope = a[2] / a[1], intercept = 0, color = "#0072B2") +
  geom_segment(
    data = show_pts,
    aes(x = co, y = no2, xend = foot_co, yend = foot_no2),
    color = "#D55E00",
  ) +
  geom_point(
    data = show_pts,
    aes(x = foot_co, y = foot_no2),
    color = "#0072B2",
    size = 3
  ) +
  geom_point(data = show_pts, color = "#D55E00", size = 2) +
  coord_equal() +
  labs(x = "Standardised carbon monoxide", y = "Standardised nitrogen dioxide")
```

Projections can be thought of like shadows. Imagine a light source shining on the data cloud from a direction perpendicular to $\bm{a}$; the projection scores are the positions of the shadows cast by the observations onto the line spanned by $\bm{a}$.

The projection direction $\bm{a}$ can be any unit vector, so there are infinitely many ways to project the data. Each choice of $\bm{a}$ captures a different perspective on the joint variation in the data. Some directions may align with the main axis of variation (like the one shown in @fig-proj-geometry), giving a wide spread of scores; others may be nearly orthogonal to that axis, giving a narrow spread; still others may be perpendicular to the axes. For example $\bm{a} = (1, 0)^\intercal$ projects onto the carbon monoxide axis, while $\bm{a} = (0, 1)^\intercal$ projects onto the nitrogen dioxide axis.

The variance of the scores is a measure of how much of the joint variation is captured by that direction. The variance of the projection scores is
$$
\text{Var}(w) = \bm{a}^\intercal \bm{\Sigma} \bm{a},
$$
where $\bm{\Sigma}$ is the $2\times 2$ covariance matrix of the data $\bm{Y}$. The variance is maximised when $\bm{a}$ is the leading eigenvector of $\bm{\Sigma}$, and the maximum variance is the leading eigenvalue. This is the basis of principal component analysis, which we will discuss in the next section.

The "residual" from the projection is the difference between the original observation and its projection onto the line. So for an observation $\bm{y}_i$, the residual is $\bm{r}_i = \bm{y}_i - \bm{a} w_i$. The magnitude of the residual is a measure of how much of the variation in that observation is not captured by the projection. (The length of each orange line in @fig-proj-geometry is the magnitude of the residual for the corresponding observation.) If the residual is large, it means that there are aspects of that observation that are not well represented by the projection direction $\bm{a}$. This is important for anomaly detection, because an observation that appears normal in one projection may have a large residual, indicating that it is an anomaly in another projection.

### Projecting from 3 to 2 dimensions

Let's add a third variable, `pm2_5`, measuring the particulate matter with diameter less than 2.5 micrometers. @fig-aq3d shows the standardised `co`, `no2` and `pm2_5` measurements at Aotizhongxin in a 3D scatterplot. What you see on the screen is a 2D projection of the 3D data. Think of a light source directly behind the screen, casting shadows of the data points onto the screen. The projection direction is perpendicular to the screen, so you are seeing the data from that perspective. As you rotate the plot, you are changing the projection direction, which gives different perspectives on the joint structure.

::: {#fig-aq3d}

::: {.figure-content}

```{r}
#| label: aq3d-plot
#| webgl: true
#| code-fold: true
#| cache: false
#| dependson: aotizhongxin
rgl::plot3d(
  az_aq[, c("co", "no2", "pm2_5")],
  size = 4,
  alpha = 0.3
)
```

:::

Interactive 3D scatterplot of the carbon monoxide, nitrogen dioxide and particulate matter (diameter $< 2.5 \mu m$) at Aotizhongxin.

:::

A pairwise scatterplot matrix, such as that shown in @fig-proj-pairs3, shows three different 2D projections of the 3D data. Each panel is the projection you would see if you were facing the above cube perpendicular to the face not shown in the plot.


```{r}
#| label: fig-proj-pairs3
#| code-fold: false
#| fig-cap: "Pairwise scatterplots and correlations for standardised carbon monoxide, nitrogen dioxide and particulate matter at Aotizhongxin. All three pairs are strongly correlated, so the joint variation is nearly confined to a two-dimensional subspace."
#| dependson: aotizhongxin
#| message: false
#| warning: false
az_aq |>
  select(co, no2, pm2_5) |>
  GGally::ggpairs()
```

A projection from $m = 3$ variables onto a 2-dimensional subspace is defined by a $3 \times 2$ matrix $\bm{A}$ where $\bm{A}^\intercal\bm{A} = \bm{I}_2$. In other words, the columns of $\bm{A}$ are "orthonormal": each column is a unit vector, and all columns are mutually orthogonal (i.e., $\bm{a}_i^\intercal\bm{a}_j = 0$ for any two distinct columns $i$ and $j$). The columns of $\bm{A}$ define the projection directions. The projection of $\bm{y}_i$ is
$$
\bm{w}_i = \bm{A}^\intercal \bm{y}_i,
$$
and its residual is $\bm{r}_i = \bm{y}_i - \bm{A}\bm{w}_i$.

### General projections

In the general case, a projection from $m$ dimensions to $k$ dimensions uses an $m \times k$ matrix $\bm{A}$ with orthonormal columns, $\bm{A}^\intercal\bm{A} = \bm{I}_k$. Writing the $n \times m$ data matrix as $\bm{Y}$, the projected scores form the $n \times k$ matrix
$$
\bm{W} = \bm{Y}\bm{A}.
$$
Their residuals are given by
$$
\bm{R} = \bm{Y} - \bm{W}\bm{A}^\intercal = \bm{Y}(\bm{I} - \bm{A}\bm{A}^\intercal).
$$

The Aotizhongxin station data `az_aq` contains `r english::words(NCOL(az_aq))` variables. We can project this `r NCOL(az_aq)`-dimensional space onto a 2-dimensional subspace defined by a $`r NCOL(az_aq)` \times 2$ matrix $\bm{A}$ with orthonormal columns. The resulting 2D projection will capture some aspects of the joint structure, but not all. For example, suppose we choose the following projection matrix $\bm{A}$:

```{r}
#| label: 11dto2d
#| include: false
#A <- prcomp(az_aq)$rotation[, 1:2] %*%
#  -matrix(c(1, 0, 0, 1), nrow = 2, ncol = 2) |>
#  round(2)
A <- dobin::dobin(az_aq)$rotation[, 1:2] |> round(2)
colnames(A) <- c("a1", "a2")
rownames(A) <- colnames(az_aq)
```

```{r}
#| label: 11dto2dA
#| code-fold: false
#| dependson: 11dto2d
A
```

Then the resulting 2D projection is shown in @fig-proj-11dto2d.

```{r}
#| label: fig-proj-11dto2d
#| code-fold: true
#| fig-cap: "Projection of the 11 air quality and meteorological variables at Aotizhongxin onto a 2-dimensional subspace."
#| dependson: 11dto2d
W <- as.matrix(az_aq) %*% A
as_tibble(W) |>
  ggplot(aes(x = a1, y = a2)) +
  geom_point(alpha = 0.2) +
  labs(x = "w1", y = "w2")
```

This is a 2D shadow of the original 11D data cloud. Each point in the plot corresponds to an 11D observation, and the axes correspond to the two projection directions defined by the columns of $\bm{A}$.

A few observations in this projection appear to be anomalies, but we may also have missed some anomalies that are only visible in other projections. This illustrates the challenge of anomaly detection in high dimensions: we may not know which projection will reveal the anomalies, and there are infinitely many projections to choose from.

For an excellent introduction to interactively exploring all possible projections, see @mulgar.

## Principal component analysis {#sec-pca}

By far the most common projection method is **principal component analysis (PCA)**, which finds the projection directions that capture the most variance in the data. The first principal component is the direction of maximum variance, the second is the direction of maximum variance orthogonal to the first, and so on. Then the lower dimensional projection will contain as much of the variation of the original data as possible, given the number of dimensions in the projection.

As in the previous sections, let $\bm{Y}$ denote the $n\times m$ data matrix, and let $\bm{\Sigma}$ be its $m\times m$ covariance matrix. The projection scores are given by $\bm{W} = \bm{Y}\bm{A}$, where $\bm{A}$ is an $m \times k$ matrix with orthonormal columns.

For principal component projection, each column of $\bm{A}$ is found sequentially by maximising the variance of the scores it produces, subject to the orthonormality constraints. Specifically, the first direction $\bm{a}_1$ maximises $\text{Var}(\bm{a}_1^\intercal \bm{y}) = \bm{a}_1^\intercal \bm{\Sigma} \bm{a}_1$ subject to $\|\bm{a}_1\| = 1$. The second direction $\bm{a}_2$ maximises $\text{Var}(\bm{a}_2^\intercal \bm{y}) = \bm{a}_2^\intercal \bm{\Sigma} \bm{a}_2$ subject to $\|\bm{a}_2\| = 1$ and $\bm{a}_2^\intercal \bm{a}_1 = 0$, and so on.

This solution is equivalent to finding the eigenvectors of $\bm{\Sigma}$, and the resulting projection scores are called the principal component scores. The first $k$ eigenvectors of $\bm{\Sigma}$ form the columns of $\bm{A}$, and the corresponding eigenvalues give the variance captured by each principal component.

It is possible to calculate up to $k = \min(n, m)$ principal components, but in practice we often only use the first few that capture most of the variance. The proportion of variance in the original data that has been captured by the first $k$ principal components is given by the sum of the first $k$ eigenvalues divided by the sum of all eigenvalues.

For the 11 air quality and meteorological variables from Aotizhongxin, the 11 principal components can be computed using `prcomp()` as follows.

```{r}
#| label: pca
#| code-fold: false
#| dependson: aotizhongxin
pca_res <- prcomp(az_aq, center = FALSE, scale = FALSE)
pca_res$rotation
```

The total variance explained can also be computed from the `prcomp()` output.

```{r}
#| label: pca-var
#| code-fold: false
#| dependson: pca
100 * pca_res$sdev^2 / sum(pca_res$sdev^2)
```

```{r}
#| label: pca-cumvar
#| echo: false
#| dependson: pca
var_explained <- 100 * pca_res$sdev^2 / sum(pca_res$sdev^2)
cum_var_explained <- round(cumsum(var_explained))
var_explained <- round(var_explained)
```

So the first three principal components capture about `r var_explained[1]` + `r var_explained[2]` + `r var_explained[3]` = `r cum_var_explained[3]`% of the variance in the data.

Using only the first two principal components gives the scores shown in @fig-pca-2d.

::: {#fig-pca-2d}

```{r}
#| label: pca-2d-plot
#| code-fold: true
#| dependson: pca
pca_res$x[, 1:2] |>
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.2)
```

Two-dimensional principal components projections of the 11D air quality and meteorological data from Aotizhongxin. The first two principal components capture about `r cum_var_explained[2]`% of the variance in the data.

:::

Here the anomalies are much less obvious than in @fig-proj-11dto2d. This is relatively common with PCA, because any anomalies can have a large influence on the variance structure of the data, and therefore on the principal components. If the anomalies dominate the variance, they can cause the principal components to point in a direction that does not capture the main body of the data, making it difficult to detect them in the projected space. This is an example of "masking", where the anomalies are masked due to their influence on the projection directions.

### Types of anomalies

@Hubert2005-qa introduced a useful framework for classifying observations in the context of PCA, based on two distances.

The **score distance** measures how far an observation is from the centre *within* the PCA subspace. For an observation $\bm{y}_i$ with scores $\bm{w}_i = \bm{A}^\intercal \bm{y}_i$, it is
$$
\text{SD}_i = \sqrt{\sum_{j=1}^{k} \frac{w_{ij}^2}{\lambda_j}},
$$
where $\lambda_j$ is the variance of the $j$th principal component. This is the Mahalanobis distance of the projected observation from the centre of the PCA subspace. Observations with the largest score distances will be visible as potential anomalies in the PCA projection, because they lie far from the centre of the PCA subspace. They are called **leverage points** because they have a large influence on the estimation of the principal component directions.

The **orthogonal distance** measures how far an observation lies *from* the PCA subspace; i.e., the magnitude of its residual:
$$
\text{OD}_i = \|\bm{r}_i\| = \|\bm{y}_i - \bm{A}\bm{w}_i\|.
$$
This provides a measure of how much information has not been captured by the $k$ components used in the projection. Observations with the largest orthogonal distances will not be visible as anomalies in the PCA projection, because they lie close to the centre of the PCA subspace, but they will be visible in the residuals. They are sometimes called **orthogonal outliers** because they lie far from the PCA subspace in a direction orthogonal to it.

Using these two distances provides a rough classification of observations based on their score and orthogonal distances.

::: {.stub-column}
|                     | Small $\text{SD}_i$ | Large $\text{SD}_i$ |
|---------------------|---------------------|---------------------|
| Small $\text{OD}_i$ | Regular observation | Good leverage point |
| Large $\text{OD}_i$ | Orthogonal outlier  | Bad leverage point  |
:::

* **Regular observations** have small score and small orthogonal distances, so they lie close to the centre of the PCA subspace, and are well represented by the projection.
* **Good leverage points** have large score distances, but small orthogonal distances, so they lie far from the centre of the PCA subspace, but are still well represented by the projection. They inflate eigenvalues but do not distort the principal component directions. In fact, they can help improve the estimation of the principal component directions, because they provide more information about the structure of the data.
* **Bad leverage points** have large score and large orthogonal distances, so they lie far from both the centre and the PCA subspace. They can pull the principal component directions towards themselves — the mechanism behind masking.
* **Orthogonal outliers** have small score distances, but large orthogonal distances, so their projections lie close to the centre of the PCA subspace, but are not well represented by the projection. They will not be visible as anomalies in PCA projection, but will be visible in the residuals.

This suggests we need to consider anomalies in both the projected space and the residual space. Alternatively, we need to find projections that reveal the anomalies in the projected space, and do not contain anomalies in the residual space.



If the data were from a multivariate Normal distribution, then the score distances would follow a chi-squared distribution with $k$ degrees of freedom, and the orthogonal distances would follow a scaled chi-squared distribution with $m-k$ degrees of freedom. While we almost never believe that real data come from a multivariate Normal distribution, these results provide a useful reference point for understanding the geometry of the score and orthogonal distances, and for setting cutoffs to flag anomalies. We set cutoffs $c_\text{SD}$ and $c_\text{OD}$ such that observations with $\text{SD}_i > c_\text{SD}$ or $\text{OD}_i > c_\text{OD}$ are flagged as anomalies.

Plotting $\text{OD}_i$ against $\text{SD}_i$ gives an **outlier map** that partitions observations into four types. @fig-pca-types illustrates the geometry for the standardised PM2.5, NO2 and CO measurements from Aotizhongxin fitted by two principal components, and @fig-outlier-map shows the resulting outlier map. These three combustion-related pollutants normally covary closely, so regular observations nearly all lie in a two-dimensional subspace; orthogonal outliers are days when one pollutant deviates from the pattern set by the other two.

- **Regular observations** ($\text{SD}_i \le c_\text{SD}$, $\text{OD}_i \le c_\text{OD}$): lie close to the centre and within the PCA subspace.
- **Good leverage points** ($\text{SD}_i > c_\text{SD}$, $\text{OD}_i \le c_\text{OD}$): lie within the subspace but far from the centre. They inflate eigenvalues but do not distort the principal component directions.
- **Orthogonal outliers** ($\text{SD}_i \le c_\text{SD}$, $\text{OD}_i > c_\text{OD}$): appear unremarkable in the projected scores but lie far from the PCA subspace, visible only in the residuals.
- **Bad leverage points** ($\text{SD}_i > c_\text{SD}$, $\text{OD}_i > c_\text{OD}$): lie far from both the centre and the subspace. They can pull the principal component directions towards themselves — the mechanism behind masking.

```{r}
#| label: pca-3d
#| include: false
#| warning: false
#| dependson: aotizhongxin
az_3d <- az_aq |> select(co, no2, pm2_5)
pca_3d <- prcomp(az_3d, center=FALSE, scale = FALSE)
A <- pca_3d$rotation[, 1:2]
scores <- pca_3d$x[, 1:2]
sd <- sqrt(scores[,1]^2/pca_3d$sdev[1]^2 + scores[,2]^2/pca_3d$sdev[2]^2)
od <- sqrt(rowSums((az_3d - scores %*% t(A))^2))
cutoff_sd <- sqrt(qchisq(0.975, df = 2))
cutoff_od <- sqrt(qchisq(0.975, df = ncol(az_3d) - 2) * pca_3d$sdev[3]^2)
types_3d <- tibble(sd = sd, od = od) |>
  mutate(
    type = case_when(
      sd <= cutoff_sd & od <= cutoff_od ~ "Regular",
      sd > cutoff_sd & od <= cutoff_od ~ "Good leverage point",
      sd <= cutoff_sd & od > cutoff_od ~ "Orthogonal outlier",
      TRUE ~ "Bad leverage point"
    ),
    type = factor(
      type,
      c("Regular", "Good leverage point", "Orthogonal outlier", "Bad leverage point")
    )
  )
```

::: {#fig-pca-types}

::: {.figure-content}

```{r}
#| label: pca-types-plot
#| webgl: true
#| echo: false
#| cache: false
#| dependson: pca-3d
type_colors <- c(
  "Regular" = "#000000",
  "Good leverage point" = "#0072B2",
  "Orthogonal outlier" = "#E69F00",
  "Bad leverage point" = "#D55E00"
)
rgl::plot3d(
  az_3d,
  col = type_colors[as.character(types_3d$type)],
  alpha = ifelse(types_3d$type == "Regular", 0.25, 0.8),
  size = 5,
  xlab = "CO", ylab = "NO2", zlab = "PM2.5"
)
s1 <- seq(
  min(scores[, 1]) * 1.4, max(scores[, 1]) * 1.4,
  length.out = 20
)
s2 <- seq(
  min(scores[, 2]) * 1.4, max(scores[, 2]) * 1.4,
  length.out = 20
)
g <- expand.grid(s1 = s1, s2 = s2)
plane_coords <- cbind(g$s1, g$s2) %*% t(A)
rgl::surface3d(
  matrix(plane_coords[, 1], 20, 20),
  matrix(plane_coords[, 2], 20, 20),
  matrix(plane_coords[, 3], 20, 20),
  alpha = 0.25, color = "grey80", back = "culled"
)
proj_pts <- scores %*% t(A)
show_idx <- which(types_3d$type != "Regular")
az_mat <- as.matrix(az_3d)
n_show <- length(show_idx)
seg_mat <- matrix(NA, nrow = 2 * n_show, ncol = 3)
seg_mat[seq(1, 2 * n_show, by = 2), ] <- az_mat[show_idx, ]
seg_mat[seq(2, 2 * n_show, by = 2), ] <- proj_pts[show_idx, ]
rgl::segments3d(
  x = seg_mat[, 1], y = seg_mat[, 2], z = seg_mat[, 3],
  col = "grey50", lwd = 1
)
```

:::

Interactive 3D scatter of standardised carbon monoxide, nitrogen dioxide and PM2.5 from Aotizhongxin (May 2014), coloured by outlier type: regular (black), good leverage point (blue), orthogonal outlier (orange), bad leverage point (red). The grey surface is the two-component PCA subspace; segments connect each non-regular observation to its projection on that surface.

:::

```{r}
#| label: fig-outlier-map
#| code-fold: true
#| dependson: pca-3d
#| fig-cap: "Outlier map for the standardised carbon monoxide, nitrogen dioxide and PM2.5 data from Aotizhongxin [@Hubert2005-qa]. Dashed lines show the cutoffs $c_{\\text{SD}}$ and $c_{\\text{OD}}$."
ggplot(types_3d, aes(x = sd, y = od, color = type, shape = type)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_vline(xintercept = cutoff_sd, linetype = "dashed", color = "grey50") +
  geom_hline(yintercept = cutoff_od, linetype = "dashed", color = "grey50") +
  scale_color_manual(
    values = c(
      "Regular" = "black",
      "Good leverage point" = "#0072B2",
      "Orthogonal outlier" = "#E69F00",
      "Bad leverage point" = "#D55E00"
    )
  ) +
  scale_shape_manual(
    values = c(
      "Regular" = 16, "Good leverage point" = 17,
      "Orthogonal outlier" = 15, "Bad leverage point" = 18
    )
  ) +
  labs(x = "Score distance", y = "Orthogonal distance", color = NULL, shape = NULL) +
  guides(color = guide_legend(nrow = 2), shape = guide_legend(nrow = 2)) +
  theme(legend.position = "bottom")
```

### Robust principal components

There are several approaches to "robustifying" PCA, such as computing the eigenvectors of a robust covariance matrix, or using a different objective function that is less sensitive to outliers.

Suppose we used the robust covariance matrix computed by `robustbase::covMCD()` instead of the sample covariance matrix.

```{r}
#| label: robust-pca
#| warning: false
#| dependson: aotizhongxin
Sigma <- as.matrix(az_aq) |> robustbase::covMcd(alpha = 0.9)
robust_pca <- eigen(Sigma$cov)
(as.matrix(az_aq) %*% robust_pca$vectors[, 1:2]) |>
  as_tibble() |>
  ggplot(aes(x = V1, y = V2)) +
  geom_point(alpha = 0.2) +
  labs(x = "Robust PC1", y = "Robust PC2")
```

This has made almost no difference to the resulting projection, so we will need another approach.

The ROBPCA algorithm of @Hubert2005-qa is a more sophisticated method that combines projection pursuit with robust covariance estimation. It starts by finding an initial projection that captures the main structure of the data, then iteratively refines the projection directions to maximise a robust measure of variance. This method is designed to be more resistant to masking and swamping than classical PCA.

```{r}
#| label: robpca
#| warning: false
#| dependson: aotizhongxin
robust_pca <- rospca::robpca(az_aq, kmax=2, skew = TRUE)
as.matrix(az_aq) %*% robust_pca$loadings |>
  as_tibble() |>
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.2) +
  labs(x = "Robust PC1", y = "Robust PC2")
```


## Multi-dimensional scaling

## Stahel-Donoho outlyingness

Stahel-Donoho outlyingness is a robust projection-based measure of how far each
observation is from the main body of the data [@Stahel81;@Donoho81].

For any unit
vector $\bm{a}$, project the data to one dimension via $\bm{a}^\intercal\bm{y}_i$, then
standardise the distance from a robust centre using a robust scale estimate. The
outlyingness of $\bm{y}_i$ is the largest standardised distance over all directions,
so an observation is flagged when it looks extreme in at least one projection.

This approach is useful in multivariate settings because it is affine equivariant and
less sensitive to masking than classical Mahalanobis-distance methods. In practice,
the optimisation over all directions is approximated using many candidate
projections, and skewness-adjusted variants are often preferred [@Brys2005-pc]. In
R, `robustbase::adjOutlyingness()` provides an efficient implementation for computing
these scores.

## Stray and HDoutliers algorithms

The HDoutliers algorithm of @Wilkinson2017-rb was designed to address several problems in anomaly detection:

1. Data containing a mixture of categorical and continuous variables.
2. Data with high dimensionality (a large number of variables).
3. Data with a large number of observations.
4. Data where anomalies occur in small clusters, rather than as isolated points.

```{r}
#| label: fig-stray
#| fig-cap: "Stray outlier scores for the Old Faithful eruption durations. Higher scores indicate greater outlyingness."
of_scores <- oldfaithful |>
  mutate(stray = stray_scores(duration))
of_scores |> arrange(desc(stray))
of_scores |>
  ggplot(aes(x = duration, y = stray)) +
  geom_point()
```

@Talagala2019-zv, @Wilkinson2017-rb

## PCout

@Filzmoser2008-qa, @Filzmoser2013-tc

* mvoutlier::pcout

## FastPCS

@Vakili2014-yc

* FastPCS::FastPCS

## DOBIN

@Kandanaarachchi2020-lo

```{r}
#| label: fig-dobin
#| echo: false
#| dependson: aotizhongxin
dobin_scores <- dobin::dobin(az_aq)
```


## Comparisons using the O3 plot
