# High-dimensional methods {#sec-highdim}

```{r}
#| label: setup
#| include: false
#| cache: false
source("before-each-chapter.R")
# Set up for using rgl
options(rgl.useNULL = TRUE)
rgl::setupKnitr()
```

Suppose we have observations $\bm{y}_1, \ldots, \bm{y}_n$, where each observation is of length $m$. We can think of each observation as a point in $m$-dimensional space. Our data can be represented as an $n \times m$ matrix $\bm{Y}$, where each row is an observation and each column is a variable. In this chapter, we will focus on the case where $m$ is large, which we will refer to as "high-dimensional" data.

The curse of dimensionality (@sec-curse) makes it difficult to detect anomalies once the number of variables grows too large. In practice, even a moderate number of variables (e.g., $m > 5$) can cause problems for some methods. The high-dimensional methods described in this chapter are designed to address these challenges, by reducing the dimensionality of the data while preserving the relevant structure for anomaly detection.

## Projections

The simplest way to reduce the dimensionality of the data is to use **linear projections**: linear combinations of the original variables that map each $m$-dimensional observation to a lower-dimensional space. Different projection directions preserve different aspects of the joint structure, so the choice of direction matters. Understanding the geometry of projections sets the foundation for the methods described in the rest of this chapter.

To keep the geometry simple, we will scale all variables to have mean zero and unit variance, although to preserve interpretability, no rotation will be applied. The ideas are illustrated using air quality and meteorological variables measured hourly at the Aotizhongxin monitoring station (at the Olympic Sports Centre) in Beijing, over May 2014.

```{r}
#| label: aotizhongxin
#| code-fold: false
az_aq <- air_quality |>
  filter(station == "Aotizhongxin", year == 2014, month == 5) |>
  tidyr::drop_na()
Y <- az_aq |>
  select(pm2_5:rainfall, wind_speed) |>
  mvscale(center = mean, scale = sd, cov = NULL)
Y
```

### Projecting from 2 to 1 dimension

The simplest case maps two variables to a single number. A projection onto a vector $\bm{a} = (a_1, a_2)^\intercal$ gives a score
$$
w_i = \bm{a}^\intercal \bm{y}_i = a_1 y_{i,1} + a_2 y_{i,2}
$$
for each observation $\bm{y}_i$, where $\bm{a}$ is a unit vector (i.e., $\|\bm{a}\| = \sqrt{a_1^2 + a_2^2} = 1$). The projection is a linear transformation that maps the 2D data cloud onto the line defined by $\bm{a}$, with $w_i$ giving the position of the projected point along that line.

When we write this in matrix form, the projection of the $n \times 2$ data matrix $\bm{Y}$ onto the vector $\bm{a}$ gives an $n$-vector of scores $\bm{w} = \bm{Y}\bm{a}$.

@fig-proj-geometry shows an example using $\bm{a} = (0.8, 0.6)^\intercal$ applied to the standardised carbon monoxide (`co`) and nitrogen dioxide (`no2`) measurements at Aotizhongxin. The blue line is the projection direction. The five orange points are randomly selected observations, showing how they project onto the line. The resulting scores are the blue dots on the line.

```{r}
#| label: fig-proj-geometry
#| code-fold: true
#| fig-cap: "Projection of standardised carbon monoxide and nitrogen dioxide from Aotizhongxin onto one dimension. The blue line shows the projection direction $\\bm{a}$. Five randomly selected observations (in orange) are shown projecting onto the line, with the resulting scores shown as blue dots."
#| dependson: aotizhongxin
set.seed(4)
# The projection direction
a <- c(0.8, 0.6)
# Select five random observations to illustrate the projection
show_pts <- Y |>
  slice_sample(n = 5) |>
  mutate(
    w = a[1] * co + a[2] * no2,
    foot_co = w * a[1],
    foot_no2 = w * a[2]
  )
# Construct the plot
ggplot(Y, aes(x = co, y = no2)) +
  geom_point(alpha = 0.25) +
  geom_abline(slope = a[2] / a[1], intercept = 0, colour = "#0072B2") +
  geom_segment(
    data = show_pts,
    aes(x = co, y = no2, xend = foot_co, yend = foot_no2),
    colour = "#D55E00",
  ) +
  geom_point(
    data = show_pts,
    aes(x = foot_co, y = foot_no2),
    colour = "#0072B2",
    size = 3
  ) +
  geom_point(data = show_pts, colour = "#D55E00", size = 2) +
  coord_equal() +
  labs(x = "Standardised carbon monoxide", y = "Standardised nitrogen dioxide")
```

Projections can be thought of like shadows. Imagine a light source shining on the data cloud from a direction perpendicular to $\bm{a}$; the projection scores are the positions of the shadows cast by the observations onto the line defined by $\bm{a}$.

The projection direction $\bm{a}$ can be any unit vector, so there are infinitely many ways to project the data. Each choice of $\bm{a}$ captures a different perspective on the joint variation in the data. Some directions may align with the main axis of variation (like the one shown in @fig-proj-geometry), giving a wide spread of scores; others may be nearly orthogonal to that axis, giving a narrow spread; still others may be perpendicular to the axes. For example, $\bm{a} = (1, 0)^\intercal$ projects onto the carbon monoxide axis, while $\bm{a} = (0, 1)^\intercal$ projects onto the nitrogen dioxide axis.

The variance of the scores is a measure of how much of the joint variation is captured by that direction, and is given by
$$
\text{Var}(w) = \bm{a}^\intercal \bm{\Sigma} \bm{a},
$$
where $\bm{\Sigma}$ is the $2\times 2$ covariance matrix of the data $\bm{Y}$. The variance is maximised when $\bm{a}$ is the leading eigenvector of $\bm{\Sigma}$; then the leading eigenvalue is the maximum variance. This is the basis of principal component analysis, which we will discuss in @sec-pca.

The "residual" from the projection is the difference between the original observation and its projection onto the line. So for an observation $\bm{y}_i$, the residual is $\bm{r}_i = \bm{y}_i - \bm{a} w_i$. The magnitude of the residual is a measure of how much of the variation in that observation is not captured by the projection, and is known as the "orthogonal distance". (The length of each orange line in @fig-proj-geometry is the orthogonal distance for the corresponding observation.) If the residual is large, it means that there are aspects of that observation that are not well represented by the projection direction $\bm{a}$. This is important for anomaly detection, because an observation that appears normal in one projection may have a large residual, indicating that it is an anomaly in another projection.

### Projecting from 3 to 2 dimensions

Let's add a third variable, `pm2_5`, measuring the particulate matter with diameter less than 2.5 micrometers. @fig-aq3d shows the standardised `co`, `no2` and `pm2_5` measurements at Aotizhongxin in a 3D scatterplot. What you see on the screen is a 2D projection of the 3D data. Think of a light source directly behind the screen, casting shadows of the data points onto the screen. The projection direction is perpendicular to the screen, so you are seeing the data from that perspective. As you rotate the plot, you are changing the projection direction, which gives different perspectives on the joint structure.

::: {#fig-aq3d}

::: {.figure-content}

```{r}
#| label: aq3d-plot
#| webgl: true
#| code-fold: true
#| cache: false
#| dependson: aotizhongxin
rgl::plot3d(
  Y[, c("co", "pm2_5", "no2")],
  size = 1,
  alpha = 0.25,
  type = "s"
)
```

:::

Interactive 3D scatterplot of the carbon monoxide, nitrogen dioxide and particulate matter (diameter $< 2.5 \mu m$) at Aotizhongxin.

:::

A pairwise scatterplot matrix, such as that shown in @fig-proj-pairs3, shows three different 2D projections of the 3D data. Each panel is the projection you would see if you were facing the above cube perpendicular to the face not shown in the plot.


```{r}
#| label: fig-proj-pairs3
#| code-fold: false
#| fig-cap: "Pairwise scatterplots and correlations for standardised carbon monoxide, nitrogen dioxide and particulate matter at Aotizhongxin. All three pairs are strongly correlated, so the joint variation is nearly confined to a two-dimensional subspace."
#| dependson: aotizhongxin
#| message: false
#| warning: false
Y |>
  select(co, no2, pm2_5) |>
  GGally::ggpairs()
```

### General projections

A projection from $m$ variables onto a $k$-dimensional subspace is defined by a $m \times k$ matrix $\bm{A}$ where $\bm{A}^\intercal\bm{A} = \bm{I}_k$. In other words, the columns of $\bm{A}$ are "orthonormal": each column is a unit vector, and all columns are mutually orthogonal (i.e., $\bm{a}_i^\intercal\bm{a}_j = 0$ for any two distinct columns $i$ and $j$). The columns of $\bm{A}$ define the projection directions. The projection of $\bm{y}_i$ is
$$
\bm{w}_i = \bm{A}^\intercal \bm{y}_i,
$$
and its residual is $\bm{r}_i = \bm{y}_i - \bm{A}\bm{w}_i$.

In matrix form, the projected scores form the rows of the $n\times k$ matrix
$$
\bm{W} = \bm{Y}\bm{A},
$$
and their residuals are given by
$$
\bm{R} = \bm{Y} - \bm{W}\bm{A}^\intercal = \bm{Y}(\bm{I} - \bm{A}\bm{A}^\intercal).
$$
The covariance matrix of the scores is given by
$$
  \text{Cov}(\bm{W}) = \bm{A}^\intercal \bm{\Sigma} \bm{A}.
$$

The Aotizhongxin station data contains `r english::words(NCOL(Y))` variables. We can project this `r NCOL(Y)`-dimensional space onto a 2-dimensional subspace defined by a $`r NCOL(Y)` \times 2$ matrix $\bm{A}$ with orthonormal columns. The resulting 2D projection will capture some aspects of the joint structure, but not all. For example, suppose we choose the following projection matrix $\bm{A}$, using a method that we will discuss later in @sec-dobin.

```{r}
#| label: 11dto2d
#| code-fold: false
#| dependson: aotizhongxin
A <- dobin::dobin(Y)$rotation[, 1:2]
colnames(A) <- c("a1", "a2")
rownames(A) <- colnames(Y)
A
```

Then the resulting 2D projection is shown in @fig-proj-11dto2d.

```{r}
#| label: fig-proj-11dto2d
#| code-fold: true
#| fig-cap: "Projection of the 11 air quality and meteorological variables at Aotizhongxin onto a 2-dimensional subspace."
#| dependson: 11dto2d
W <- as.matrix(Y) %*% A
az_aq_proj <- az_aq |>
  bind_cols(W)
az_aq_proj |>
  ggplot(aes(x = a1, y = a2)) +
  geom_point(alpha = 0.2, colour = "#0072B2")
```

This is a 2D shadow of the original 11D data cloud. Each point in the plot corresponds to an 11D observation, and the axes correspond to the two projection directions defined by the columns of $\bm{A}$.

A few observations in this projection appear to be anomalies, but we may also have missed some anomalies that are only visible in other projections. This illustrates the challenge of anomaly detection in high dimensions: we may not know which projection will reveal the anomalies, and there are infinitely many projections to choose from.

For an excellent introduction to interactively exploring all possible projections, see @mulgar.

## Types of projection anomalies

There are two types of projection-based anomalies:

1. **Anomalies in the projection space**. For example, they may be far from the other observations in the projection, or they may be in a region of the projection space that has low density. We can identify them by applying anomaly-detection methods to the projected scores.
2. **Anomalies in the residual space**. For example, observations with large residuals, or observations that have otherwise unusual residual magnitudes. We can detect them by applying anomaly-detection methods to the residual magnitudes. It is not usually possibly to identify anomalies in the residual space by looking at the residuals themselves, because they are still in a high-dimensional space.

### Anomalies in the projection space

@fig-proj-11dto2d shows some potential anomalies in the projection space. The general distance-based and density-based anomaly detection methods discussed in @sec-distances and @sec-density-methods can be applied to the projected scores to flag these anomalies.

For example, using KDE surprisals (@sec-kdesurprisals), we can flag the observations with the lowest density in the projection as potential anomalies. Here the threshold probability of 0.02 has been chosen after some experimentation.

```{r}
#| label: proj-kde
#| code-fold: false
#| dependson: fig-proj-11dto2d
az_aq_proj <- az_aq_proj |>
  mutate(
    prob = surprisals_prob(pick(a1, a2), loo = TRUE, approximation = "gpd"),
    surprisal_anomaly = prob < 0.02
  )
az_aq_proj |>
  filter(surprisal_anomaly)
```

```{r}
#| label: fig-proj-kde
#| code-fold: true
#| dependson: proj-kde
#| fig-cap: "Projection of the 11 air quality and meteorological variables at Aotizhongxin onto a 2-dimensional subspace, with points coloured by whether their KDE surprisal probability is less than 0.02."
az_aq_proj |>
  ggplot(aes(x = a1, y = a2, colour = surprisal_anomaly)) +
  geom_point(alpha = 0.2 + 0.8 * az_aq_proj$surprisal_anomaly) +
  scale_colour_manual(values = c("FALSE" = "#0072B2", "TRUE" = "black"))
```

This shows that there was an unusual period between 8am and 12noon on 22 May 2024. These correspond to the cluster of five observations near the bottom of @fig-proj-kde. Further inspection shows that this was a period of high particulate matter (`pm2_5` and `pm10`), `co` and `dew_point`, and moderately high `so2`. The point on the far right of @fig-proj-kde corresponds to the last line of the table above. The observations at that time showed high `co` and `dew_point` and very high `rainfall`.

Similar results are obtained with other anomaly detection methods. For example, in @fig-proj-isoforest we use an Isolation Forest to find the top 10 most anomalous observations.

```{r}
#| label: fig-proj-isoforest
#| code-fold: true
#| dependson: fig-proj-11dto2d
#| fig-cap: "Projection of the 11 air quality and meteorological variables at Aotizhongxin onto a 2-dimensional subspace, with points coloured by whether their Isolation Forest score is in the top 10 most anomalous observations."
iso_model <- isotree::isolation.forest(W)
az_aq_proj <- az_aq_proj |>
  mutate(
    iso_score = predict(iso_model, W),
    top_ten = iso_score > sort(iso_score, decreasing = TRUE)[10]
  )
az_aq_proj |>
  ggplot(aes(x = a1, y = a2, colour = top_ten)) +
  geom_point(alpha = 0.2 + 0.8 * az_aq_proj$top_ten) +
  scale_colour_manual(values = c("FALSE" = "#0072B2", "TRUE" = "black"))
```

### Anomalies in the residual space

To find anomalies in the residual space, we first calculate the **orthogonal distances**, measuring how far an observation lies *from* the projection subspace; i.e., the magnitude of its residual:
$$
\text{OD}_i = \|\bm{r}_i\| = \|\bm{y}_i - \bm{A}\bm{w}_i\|.
$$
This provides a measure of how much information has *not* been captured by the $k$ components used in the projection. Observations with the largest orthogonal distances may not be visible as anomalies in the projection space, but they will be visible in the residuals.

```{r}
#| label: fig-proj-residuals1
#| dependson: fig-proj-11dto2d
#| code-fold: false
#| fig-asp: 0.2
#| fig-cap: HDR boxplot of the orthogonal distances of the observations from the projection subspace. Anomalies are shown in black.
R <- Y - as.matrix(W) %*% t(A)
az_aq_proj <- az_aq_proj |>
  mutate(od = sqrt(rowSums(R^2)))
az_aq_proj |> arrange(desc(od))
az_aq_proj |> gg_hdrboxplot(od)
```

We can highlight these residual anomalies in the projection space, as shown in @fig-proj-residuals2. The points with the largest orthogonal distances are highlighted in black.

```{r}
#| label: fig-proj-residuals2
#| dependson: fig-proj-residuals1
#| fig-cap: Projection of the 11 air quality and meteorological variables at Aotizhongxin onto a 2-dimensional subspace, with points coloured by whether their orthogonal distance from the projection subspace is in the top 1% of all orthogonal distances."
threshold <- quantile(az_aq_proj$od, 0.99, type = 8)
az_aq_proj |>
  ggplot(aes(x = a1, y = a2, colour = od >= threshold)) +
  geom_point(alpha = 0.2 + 0.8 * (az_aq_proj$od >= threshold)) +
  scale_colour_manual(values = c("FALSE" = "#0072B2", "TRUE" = "black")) +
  guides(colour = "none")
```

The observation with the most extreme orthogonal distance, is also the most extreme in the projection space, so it is visible as an anomaly in both spaces. However, the other observations with large orthogonal distances are not visible as anomalies in the projection space, illustrating the importance of considering both spaces when looking for anomalies.

### Outlier map

To explore outlying anomalies, we can look at the relationship between two different distances: the **score distance**, which measures how far an observation lies from the centre *within* the projection space, and the **orthogonal distance** (defined above), which measures how far it lies *from* the projection space.

The **score distance** is the Mahalanobis distance of the projected observation from the centre of the projection subspace (@sec-distances). For an observation $\bm{y}_i$ with scores $\bm{w}_i = \bm{A}^\intercal \bm{y}_i$, it is
$$
\text{SD}_i = \sqrt{\bm{w}_i^\intercal (\bm{A}^\intercal \bm{\Sigma} \bm{A})^{-1} \bm{w}_i}.
$$
Observations with the largest score distances will be visible as potential anomalies, because they lie far from the centre of the projection subspace.

```{r}
#| label: score-distances
#| code-fold: false
#| dependson: fig-proj-11dto2d
az_aq_proj <- az_aq_proj |>
  mutate(sd = sqrt(diag(W %*% solve(t(A) %*% cov(Y) %*% A) %*% t(W))))
```

We can plot the orthogonal distance against the score distance to explore the relationship between these two types of anomalies. This is known as an "outlier map" [@Hubert2005-qa].

```{r}
#| label: fig-odsdplot
#| code-fold: true
#| dependson: score-distances
#| fig-cap: "Orthogonal distance plotted against score distance for the 11 air quality and meteorological variables at Aotizhongxin projected onto a 2-dimensional subspace."
az_aq_proj |>
  ggplot(aes(x = sd, y = od)) +
  geom_point(alpha = 0.4) +
  labs(x = "Score distance", y = "Orthogonal distance")
```

Here there are two observations with large orthogonal distances, but moderate score distances, and one observation with both large orthogonal and score distances.

## Stahel-Donoho outlyingness {#sec-stahel-donoho}

Stahel-Donoho outlyingness is a robust projection-based measure of how far each
observation is from the main body of the data [@Stahel81;@Donoho81] in at least one projection.

For any unit vector $\bm{a}$, project the data to one dimension via $\bm{a}^\intercal\bm{y}_i$, then scale the result using a robust z-score (@sec-scaling), adjusted for skewness [@Brys2005-pc]. The outlyingness of $\bm{y}_i$ is the largest standardised absolute z-score over all possible directions $\bm{a}$, so an observation is flagged when it looks extreme in at least one projection.

In practice, the optimisation over all directions is approximated using many candidate projections. In R, `robustbase::adjOutlyingness()` provides an efficient implementation for computing these scores.

We can apply this to the same standardised air quality and meteorological variables from Aotizhongxin used in @fig-proj-11dto2d. Because the outlyingness is maximised over all directions, it does not require us to choose a projection in advance: an observation receives a high score if it is extreme in *any* direction, whether that direction lies within the 2-dimensional projection or orthogonal to it.

```{r}
#| label: stahel-donoho
#| code-fold: false
#| dependson: fig-proj-11dto2d
set.seed(1)
ao <- robustbase::adjOutlyingness(Y)
az_aq_proj <- az_aq_proj |>
  mutate(top_1pc = ao$adjout > quantile(ao$adjout, prob = 0.99, type = 8))
```

@fig-stahel-donoho shows the 2-dimensional projection from @fig-proj-11dto2d, with the top 1% most outlying observations highlighted in black.

```{r}
#| label: fig-stahel-donoho
#| code-fold: true
#| dependson: stahel-donoho
#| fig-cap: "Projection of the 11 air quality and meteorological variables at Aotizhongxin onto a 2-dimensional subspace, with the observations flagged by their Stahel–Donoho outlyingness highlighted in black."
az_aq_proj |>
  ggplot(aes(x = a1, y = a2, colour = top_1pc)) +
  geom_point(alpha = 0.2 + 0.8 * az_aq_proj$top_1pc) +
  scale_colour_manual(values = c("FALSE" = "#0072B2", "TRUE" = "black")) +
  guides(colour = "none")
```

Clearly the 2D projection does not capture all the anomalies, but it does capture one or two of them. Also, the 1D projections used in the Stahel-Donoho method misses anomalies that can only be seen in multiple dimensions, such as the group of five observations near the bottom of @fig-stahel-donoho.

## Principal component analysis {#sec-pca}

So far, we haven't said where the projection matrix $\bm{A}$ comes from. By far the most common projection method is **principal component analysis (PCA)**, which finds the projection directions that capture the most variance in the data. The first principal component is the direction of maximum variance, the second is the direction of maximum variance orthogonal to the first, and so on. Then the lower dimensional projection will contain as much of the variation of the original data as possible, given the number of dimensions in the projection.

For principal component projection, each column of $\bm{A}$ is found sequentially by maximising the variance of the scores it produces, subject to the orthonormality constraints. If $\bm{\Sigma}$ is the $m\times m$ covariance matrix of the data $\bm{Y}$, then the first direction $\bm{a}_1$ maximises $\text{Var}(\bm{a}_1^\intercal \bm{y}) = \bm{a}_1^\intercal \bm{\Sigma} \bm{a}_1$ subject to $\|\bm{a}_1\| = 1$. The second direction $\bm{a}_2$ maximises $\text{Var}(\bm{a}_2^\intercal \bm{y}) = \bm{a}_2^\intercal \bm{\Sigma} \bm{a}_2$ subject to $\|\bm{a}_2\| = 1$ and $\bm{a}_2^\intercal \bm{a}_1 = 0$, and so on.

This solution is equivalent to finding the eigenvectors of $\bm{\Sigma}$, and the resulting projection scores are called the principal component scores. The first $k$ eigenvectors of $\bm{\Sigma}$ form the columns of $\bm{A}$.

The covariance matrix of the scores is given by
$$
  \text{Cov}(\bm{W}) = \bm{A}^\intercal \bm{\Sigma} \bm{A} = \text{diag}(\lambda_1, \ldots, \lambda_k),
$$
where $\lambda_1 \ge \lambda_2 \ge \cdots \ge \lambda_k$ are the first $k$ eigenvalues of $\bm{\Sigma}$. The covariance matrix of the scores is diagonal, which means that the principal component scores are uncorrelated. The variance of the $j$th principal component score is $\lambda_j$, so the eigenvalues give the variance captured by each principal component. The proportion of variance in the original data that has been captured by the first $k$ principal components is given by the sum of the first $k$ eigenvalues divided by the sum of all eigenvalues.

It is possible to calculate up to $k = \min(n, m)$ principal components, but in practice we often only use the first few that capture most of the variance.

For the 11 air quality and meteorological variables from Aotizhongxin, the 11 principal components can be computed using `prcomp()`, or using `rrcov::PcaClassic()`. We will use the latter, because it will make it easier to compare with other methods.

```{r}
#| label: pca
#| code-fold: false
#| dependson: aotizhongxin
#| message: false
library(rrcov)
pca_res <- PcaClassic(Y)
pca_res$loadings
```

The variance explained by the first $k$ components can also be computed from the output.

```{r}
#| label: pca-var
#| code-fold: false
#| dependson: pca
100 * cumsum(pca_res$eigenvalues) / sum(pca_res$eigenvalues)
```

```{r}
#| label: pca-cumvar
#| echo: false
#| dependson: pca
cum_var_explained <- round(100 * cumsum(pca_res$eigenvalues) / sum(pca_res$eigenvalues))
```

Using only the first two principal components gives the scores shown in @fig-pca-2d.

::: {#fig-pca-2d}

```{r}
#| label: pca-2d-plot
#| code-fold: true
#| dependson: pca
pca_res$scores[, 1:2] |>
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.2)
```

Two-dimensional principal components projections of the 11D air quality and meteorological data from Aotizhongxin. The first two principal components capture about `r cum_var_explained[2]`% of the variance in the data.

:::

Here the anomalies are much less obvious than in @fig-proj-11dto2d. This can happen for two reasons:

1. If the anomalies are not aligned with the main axes of variation, they may be hidden in the lower variance components that we have not included in the 2D projection, although they may be visible in the residuals.
2. The anomalies can cause the principal components to point in a direction that does not capture the main body of the data, making it difficult to detect them in the projection space. This is an example of "masking", where the anomalies are hidden due to their influence on the projection directions.

### Diagnostic plot

To distinguish these two situations, @Hubert2005-qa introduced a useful framework for classifying observations in the context of PCA, based on the score distance and orthogonal distance.

::: {.stub-column}
|                     | Small $\text{SD}_i$ | Large $\text{SD}_i$ |
|---------------------|---------------------|---------------------|
| Small $\text{OD}_i$ | Regular observation | Good leverage point |
| Large $\text{OD}_i$ | Orthogonal outlier  | Bad leverage point  |
:::

* **Regular observations** have small score and small orthogonal distances, so they lie close to the centre of the PCA subspace, and are well represented by the projection.
* **Good leverage points** have large score distances, but small orthogonal distances, so they lie far from the centre of the PCA subspace, but are still well represented by the projection. They inflate eigenvalues but do not distort the principal component directions. In fact, they can help improve the estimation of the principal component directions, because they provide more information about the structure of the data.
* **Bad leverage points** have large score and large orthogonal distances, so they lie far from both the centre and the PCA subspace. They can pull the principal component directions towards themselves — the mechanism behind masking.
* **Orthogonal outliers** have small score distances, but large orthogonal distances, so their projections lie close to the centre of the PCA subspace, but are not well represented by the projection. They will not be visible as anomalies in PCA projection, but they will be visible in the residuals.

If the data were from a multivariate Normal distribution, then the score distances would approximately follow a chi-squared distribution with $k$ degrees of freedom. While we almost never believe that real data come from a multivariate Normal distribution, this provides a useful reference point for understanding the geometry of the score distances, and for setting a cutoff to classify points. We flag an observation as a high leverage point if $\text{SD}_i$ is larger than the 97.5% quantile of the chi-squared distribution with $k$ degrees of freedom. So under a Normal model, we would expect about 2.5% of the observations to have large score distances and be flagged as leverage points.

The distribution of orthogonal distances is more complicated, but a common approach is to flag an observation as having a large orthogonal distance if $\text{OD}_i$ is greater than the 97.5% quantile of an appropriate Normal distribution. So again, about 2.5% of the observations would be expected to have large orthogonal distances under a Normal model.

When combined with the outlier map shown above, these cutoffs partition the observations into the four types described in the table. @fig-pca-outliermap illustrates the idea for the Aotizhongxin data using the two principal components shown in @fig-pca-2d.


```{r}
#| label: fig-pca-outliermap
#| code-fold: true
#| dependson: pca
#| fig-cap: "Outlier map for the standardised carbon monoxide, nitrogen dioxide and PM2.5 data from Aotizhongxin [@Hubert2005-qa]. Dashed lines show the cutoffs $c_{\\text{SD}}$ and $c_{\\text{OD}}$."
outlier_map <- function(Y, k, pca_fn = rrcov::PcaClassic, guide = FALSE, ...) {
  pca_res <- pca_fn(Y, k = k, ...)
  df <- as_tibble(pca_res$scores) |>
    mutate(
      sd = pca_res$sd,
      od = pca_res$od,
      type = case_when(
        sd <= pca_res$cutoff.sd & od <= pca_res$cutoff.od ~ "Regular",
        sd > pca_res$cutoff.sd & od <= pca_res$cutoff.od ~ "Good leverage point",
        sd <= pca_res$cutoff.sd & od > pca_res$cutoff.od ~ "Orthogonal outlier",
        sd > pca_res$cutoff.sd & od > pca_res$cutoff.od ~ "Bad leverage point"
      ),
      type = factor(
        type,
        levels = c(
          "Orthogonal outlier",
          "Regular",
          "Bad leverage point",
          "Good leverage point"
        )
      )
    )
  p <- df |>
    ggplot(aes(x = sd, y = od, colour = type)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_vline(xintercept = pca_res$cutoff.sd, linetype = "dashed", colour = "grey50") +
    geom_hline(yintercept = pca_res$cutoff.od, linetype = "dashed", colour = "grey50") +
    labs(
      x = "Score distance",
      y = "Orthogonal distance",
      title = paste0("PCA with ", k, " component", ifelse(k == 1, "", "s"))
    ) +
    scale_colour_manual(
      values = c(
        "Regular" = "#0072B2",
        "Good leverage point" = "#009E73",
        "Orthogonal outlier" = "#D55E00",
        "Bad leverage point" = "black"
      )
    )
  if (guide) {
    p <- p + guides(colour = guide_legend(nrow = 2)) + theme(legend.position = "bottom")
  } else {
    p <- p + guides(colour = "none")
  }
  return(p)
}
outlier_map(Y, 2, guide = TRUE)
```

Here we can see that there are no points that are particularly bad leverage points (the two black points are very close to the cutoff). So the most anomalous observations are orthogonal outliers in the residual space, but these are not having a strong effect on the first two principal component directions.

We can repeat this analysis using different values of $k$, the number of principal components used in the projection. As $k$ increases, more of the variance is captured by the projection, which can cause the orthogonal distances to decrease, and the score distances to increase, because the projection is capturing more of the structure of the data. Any outlying anomalies that are not well represented by the first few principal components will have large orthogonal distances, but as we include more components, they may become better represented in the projection. This can lead to some anomalies being classified as good leverage points in higher dimensions, even if they were orthogonal outliers in lower dimensions.


```{r}
#| label: fig-pca-outliermaps
#| code-fold: true
#| dependson: fig-pca-outliermap
#| fig-cap: "Outlier maps for the Aotizhongxin data, using different numbers of principal components in the projection."
patchwork::wrap_plots(
  outlier_map(Y, 1),
  outlier_map(Y, 2),
  outlier_map(Y, 3),
  outlier_map(Y, 4),
  ncol = 2,
  byrow = TRUE
)
```

This can be seen in the second-largest score distance in the $k=4$ plot. The corresponding observation was an orthogonal outlier in the other plots, but is now a good leverage point, because it is now well represented in the projection.

Notice that we now also have some bad leverage values in $k=3$ and $k=4$. So the third and subsequent principal components are being pulled towards some anomalies, making them harder to identify in the lower dimensional space. This could be masking some of the anomalies.

### Robust principal components

There are several approaches to "robustifying" PCA, such as computing the eigenvectors of a robust covariance matrix, or using a different objective function that is less sensitive to outliers. One approach is the ROBPCA algorithm of @Hubert2005-qa, which combines projection pursuit with robust covariance estimation. Before describing the algorithm, let's see its effect on the outlier maps.

```{r}
#| label: fig-pca-outliermap2
#| code-fold: true
#| dependson: fig-pca-outliermap
#| fig-cap: "Outlier maps for the Aotizhongxin data, using different numbers of principal components in the projection."
#| message: false
patchwork::wrap_plots(
  outlier_map(Y, 1, pca_fn = rrcov::PcaHubert, alpha = 0.9, skew = TRUE, mcd = FALSE),
  outlier_map(Y, 2, pca_fn = rrcov::PcaHubert, alpha = 0.9, skew = TRUE, mcd = FALSE),
  outlier_map(Y, 3, pca_fn = rrcov::PcaHubert, alpha = 0.9, skew = TRUE, mcd = FALSE),
  outlier_map(Y, 4, pca_fn = rrcov::PcaHubert, alpha = 0.9, skew = TRUE, mcd = FALSE),
  ncol = 2,
  byrow = TRUE
)
```

Now there are far fewer bad leverage points, and the map changes much less as we increase $k$. This suggests that the ROBPCA algorithm has found projection directions that are less influenced by the anomalies, so they are easier to identify in the projection space.

The ROBPCA algorithm proceeds in three stages, combining projection pursuit ideas (used to obtain a robust starting point) with robust covariance estimation (used to refine the result).

**Stage 1: Reduce the working dimension.** When $m$ is large, the $n$ observations span a subspace of dimension at most $n-1$. A singular value decomposition of the mean-centred data matrix is used to express the data in this subspace without any loss of information. This is purely an affine transformation — no components are discarded yet — and it ensures that the subsequent steps operate in a space of dimension no larger than $\min(n-1, m)$.

**Stage 2: Find a robust subspace.** For each observation, the Stahel–Donoho outlyingness (@sec-stahel-donoho) is computed by projecting onto many candidate directions and recording the largest robustly standardised distance from the centre. The $h$ least-outlying observations are retained, where $h = \lceil \alpha n \rceil$ for some $\alpha$ between 0.5 and 1 (the default is $\alpha = 0.75$); a larger $\alpha$ gives more efficient estimates, while a smaller $\alpha$ gives greater robustness. The eigenvalues of the covariance matrix of these $h$ points are used to choose the number of components $k$ (for example, using a scree plot or a proportion-of-variance criterion), and the data are projected onto the first $k$ eigenvectors.

**Stage 3: Robustly estimate the centre and scatter.** Within this $k$-dimensional subspace, the minimum covariance determinant (MCD) estimator is applied to obtain a robust centre $\hat{\bm{\mu}}$ and scatter matrix $\bm{S}$. The estimate is then reweighted to improve efficiency, downweighting observations whose robust distance is large. The eigenvectors of $\bm{S}$ give the final robust principal component directions, which are transformed back to the original $m$-dimensional space. The score and orthogonal distances are computed exactly as before, but using the robust centre and the robust loadings, so that the bad leverage points no longer distort the projection.

Because the projection directions are estimated from the bulk of the data rather than from all observations, anomalies cannot pull the components towards themselves. This is why the outlier maps in @fig-pca-outliermap2 show so few bad leverage points and change little as $k$ increases.


## Multidimensional scaling


## Stray and HDoutliers algorithms

The HDoutliers algorithm of @Wilkinson2017-rb was designed to address several problems in anomaly detection:

1. Data containing a mixture of categorical and continuous variables.
2. Data with high dimensionality (a large number of variables).
3. Data with a large number of observations.
4. Data where anomalies occur in small clusters, rather than as isolated points.

```{r}
#| label: fig-stray
#| fig-cap: "Stray outlier scores for the Old Faithful eruption durations. Higher scores indicate greater outlyingness."
of_scores <- oldfaithful |>
  mutate(stray = stray_scores(duration))
of_scores |> arrange(desc(stray))
of_scores |>
  ggplot(aes(x = duration, y = stray)) +
  geom_point()
```

@Talagala2019-zv, @Wilkinson2017-rb

## PCout

@Filzmoser2008-qa, @Filzmoser2013-tc

* mvoutlier::pcout

## FastPCS

@Vakili2014-yc

* FastPCS::FastPCS

## DOBIN {#sec-dobin}

@Kandanaarachchi2020-lo

```{r}
#| label: fig-dobin
#| echo: false
#| dependson: aotizhongxin
dobin_scores <- dobin::dobin(Y)
```


## Comparisons using the O3 plot
