# High-dimensional methods {#sec-highdim}

```{r}
#| label: setup
#| include: false
#| cache: false
source("before-each-chapter.R")
# Set up for using rgl
options(rgl.useNULL = TRUE)
rgl::setupKnitr()
```

The curse of dimensionality (@sec-curse) makes it difficult to detect anomalies once the number of variables grows too large. In practice, even a moderate number of variables (e.g., $m > 5$) can cause problems for some methods. The high-dimensional methods described in this chapter are designed to address these challenges, by reducing the dimensionality of the data while preserving the relevant structure for anomaly detection.

## Projections

The simplest way to reduce the dimensionality of the data is to use **linear projections**: linear combinations of the original variables that map each $p$-dimensional observation to a lower-dimensional space. Different projection directions preserve different aspects of the joint structure, so the choice of direction matters. Understanding the geometry of projections sets the foundation for the methods described in the rest of this chapter.

To keep the geometry simple, we will scale all variables to have mean zero and unit variance, although to preserve interpretability, no rotation will be applied. The ideas are illustrated using meteorological variables measured hourly at the Aotizhongxin monitoring station (at the Olympic Sports Centre) in Beijing, over May 2014.

```{r}
#| label: aotizhongxin
#| code-fold: false
az_aq <- air_quality |>
  filter(station == "Aotizhongxin", year == 2014, month == 5) |>
  select(temperature, pressure, dew_point, wind_speed, rainfall) |>
  mvscale(center = mean, scale = sd, cov = NULL)
```

### Projecting from 2 to 1 dimension

The simplest case maps two variables to a single number. A projection onto a vector $\bm{a} = (a_1, a_2)^\intercal$ gives a score
$$
w_i = \bm{a}^\intercal \bm{y}_i = a_1 y_{i,1} + a_2 y_{i,2}
$$
for each observation, where $\bm{a}$ is a unit vector (i.e., $\|\bm{a}\| = \sqrt{a_1^2 + a_2^2} = 1$). The projection is a linear transformation that maps the 2D data cloud onto the line spanned by $\bm{a}$, with $w_i$ giving the position of the projected point along that line.

@fig-proj-geometry shows an example using $\bm{a} = (-0.8, 0.6)^\intercal$ applied to the standardised temperature and pressure measurements at Aotizhongxin. The blue line is the projection direction. The five orange points are randomly selected observations, showing how they project onto the line. The resulting scores are the blue dots on the line.

```{r}
#| label: fig-proj-geometry
#| code-fold: true
#| fig-cap: "Projection of standardised temperature and pressure from Aotizhongxin onto one dimension. The blue line shows the projection direction. Five randomly selected observations (in orange) are shown projecting onto the line, with the resulting scores shown as blue dots."
#| dependson: aotizhongxin
set.seed(2)
# The projection direction
a <- c(-0.8, 0.6)
# Select five random observations to illustrate the projection
show_pts <- az_aq |>
  slice_sample(n = 5) |>
  mutate(
    w = a[1] * temperature + a[2] * pressure,
    foot_temp = w * a[1],
    foot_press = w * a[2]
  )
# Construct the plot
ggplot(az_aq, aes(x = temperature, y = pressure)) +
  geom_point(alpha = 0.2) +
  geom_abline(slope = a[2] / a[1], intercept = 0, color = "#0072B2") +
  geom_segment(
    data = show_pts,
    aes(x = temperature, y = pressure, xend = foot_temp, yend = foot_press),
    color = "#D55E00",
  ) +
  geom_point(
    data = show_pts,
    aes(x = foot_temp, y = foot_press),
    color = "#0072B2",
    size = 3
  ) +
  geom_point(data = show_pts, color = "#D55E00", size = 2) +
  coord_equal() +
  labs(x = "Standardised temperature", y = "Standardised pressure")
```

Projections can be thought of like shadows. Imagine a light source shining on the data cloud from a direction perpendicular to $\bm{a}$; the projection scores are the positions of the shadows cast by the observations onto the line spanned by $\bm{a}$.

The projection direction $\bm{a}$ can be any unit vector, so there are infinitely many ways to project the data. Each choice of $\bm{a}$ captures a different perspective on the joint variation in the data. Some directions may align with the main axis of variation (like the one shown in @fig-proj-geometry), giving a wide spread of scores; others may be nearly orthogonal to that axis, giving a narrow spread; still others may be perpendicular to the axes. For example $\bm{a} = (1, 0)^\intercal$ projects onto the temperature axis, while $\bm{a} = (0, 1)^\intercal$ projects onto the pressure axis.

The variance of the scores is a measure of how much of the joint variation is captured by that direction. The variance of the projection scores is
$$
\text{Var}(w) = \bm{a}^\intercal \bm{\Sigma} \bm{a},
$$
where $\bm{\Sigma}$ is the covariance matrix of the data $\bm{y}$. The variance is maximised when $\bm{a}$ is the leading eigenvector of $\bm{\Sigma}$, and the maximum variance is the leading eigenvalue. This is the basis of principal component analysis, which we will discuss in the next section.

### Projecting from 3 to 2 dimensions

@fig-aq3d shows the standardised temperature, pressure and dew point measurements at Aotizhongxin in a 3D scatterplot. What you see on the screen is a 2D projection of the 3D data. Think of a light source directly behind the screen, casting shadows of the data points onto the screen. The projection direction is perpendicular to the screen, so you are seeing the data from that perspective. As you rotate the plot, you are changing the projection direction, which gives different perspectives on the joint structure. 

::: {#fig-aq3d}

::: {.figure-content}

```{r}
#| webgl: true
#| code-fold: true
#| cache: false
rgl::plot3d(
  az_aq[, c("temperature", "pressure", "dew_point")],
  size = 4, alpha = 0.3
)
```

:::

Interactive 3D scatterplot of the temperature, pressure and dew point at Aotizhongxin.

:::

A pairwise scatterplot matrix, such as that shown in @fig-proj-pairs3, shows three different 2D projections of the 3D data. Each panel is a projection onto the plane spanned by the two variables shown on the axes. 


```{r}
#| label: fig-proj-pairs3
#| code-fold: false
#| fig-cap: "Pairwise scatterplots and correlations for standardised temperature, pressure and dew point at Aotizhongxin. All three pairs are strongly correlated, so the joint variation is nearly confined to a two-dimensional subspace."
#| dependson: aotizhongxin
#| message: false
#| warning: false
az_aq |>
  select(temperature, pressure, dew_point) |>
  GGally::ggpairs()
```

A projection from $p = 3$ variables onto a 2-dimensional subspace is defined by a $3 \times 2$ matrix $\bm{A}$ where $\bm{A}^\intercal\bm{A} = \bm{I}_2$. In other words, the columns of $\bm{A}$ are "orthonormal": each column is a unit vector, and all columns are mutually orthogonal (i.e., $\bm{a}_i^\intercal\bm{a}_j = 0$ for any two distinct columns $i$ and $j$). The columns of $\bm{A}$ define the projection directions.

The 2-dimensional projection of $\bm{y}_i$ is
$$
\bm{w}_i = \bm{A}^\intercal \bm{y}_i 
$$

### General projections

In the general case, a projection from $p$ dimensions to $k$ dimensions uses a $p \times k$ matrix $\bm{A}$ with orthonormal columns, $\bm{A}^\intercal\bm{A} = \bm{I}_k$. Writing the $n \times p$ data matrix as $\bm{Y}$, the projected scores form the $n \times k$ matrix
$$
\bm{W} = \bm{Y}\bm{A}.
$$

The Aotizhongxin station records five meteorological variables: temperature, pressure, dew point, wind speed, and rainfall. @fig-proj-all shows the pairwise relationships among all five.

```{r}
#| label: fig-proj-all
#| code-fold: false
#| dependson: aotizhongxin
#| fig-cap: "Pairwise scatterplots and correlations for five standardised meteorological variables at Aotizhongxin. Wind speed and rainfall are largely uncorrelated with the temperature-pressure-dew point group, so the five variables span a more complex joint distribution than the three-variable case."
az_aq |>
  select(temperature, pressure, dew_point, wind_speed, rainfall) |>
  GGally::ggpairs()
```

```{r}
#| label: 5dto2d
#| include: false
A <- prcomp(az_aq, center = FALSE, scale = FALSE, rank = 2)$rotation |>
  round(2)
A <- dobin::dobin(az_aq)$rotation[,1:2] |>
  round(2)
colnames(A) <- c("a1", "a2")
```

We can project these five variables onto a 2-dimensional subspace defined by a $5 \times 2$ matrix $\bm{A}$ with orthonormal columns. The resulting 2D projection will capture some aspects of the joint structure, but not all. For example, suppose we choose the following projection matrix $\bm{A}$:

```{r}
#| label: 5dto2dA
#| code-fold: false
A
```

Then the resulting 2D projection is shown in @fig-proj-5dto2d. 

```{r}
#| label: fig-proj-5dto2d
#| code-fold: true
#| fig-cap: "Projection of the five standardised meteorological variables at Aotizhongxin onto a 2-dimensional subspace."
as.matrix(az_aq) %*% A |>
  as_tibble() |>
  ggplot(aes(x = a1, y = a2)) +
  geom_point(alpha = 0.2) +
  labs(x = "w1", y = "w2")
```

A few observations in this projection appear to be anomalies, but we may also have missed some anomalies that are only visible in other projections. This illustrates the challenge of anomaly detection in high dimensions: we may not know which projection will reveal the anomalies, and there are infinitely many projections to choose from.

For an excellent introduction to interactively exploring all possible projections, see @mulgar.

## Principal component analysis {#sec-pca}

By far the most common projection method is **principal component analysis (PCA)**, which finds the projection directions that capture the most variance in the data. The first principal component is the direction of maximum variance, the second is the direction of maximum variance orthogonal to the first, and so on. 



## Multi-dimensional scaling

## Stahel-Donoho outlyingness

Stahel-Donoho outlyingness is a robust projection-based measure of how far each
observation is from the main body of the data [@Stahel81;@Donoho81].

For any unit
vector $\bm{a}$, project the data to one dimension via $\bm{a}^\intercal\bm{y}_i$, then
standardise the distance from a robust centre using a robust scale estimate. The
outlyingness of $\bm{y}_i$ is the largest standardised distance over all directions,
so an observation is flagged when it looks extreme in at least one projection.

This approach is useful in multivariate settings because it is affine equivariant and
less sensitive to masking than classical Mahalanobis-distance methods. In practice,
the optimisation over all directions is approximated using many candidate
projections, and skewness-adjusted variants are often preferred [@Brys2005-pc]. In
R, `robustbase::adjOutlyingness()` provides an efficient implementation for computing
these scores.

## Stray and HDoutliers algorithms

The HDoutliers algorithm of @Wilkinson2017-rb was designed to address several problems in anomaly detection:

1. Data containing a mixture of categorical and continuous variables.
2. Data with high dimensionality (a large number of variables).
3. Data with a large number of observations.
4. Data where anomalies occur in small clusters, rather than as isolated points.

```{r}
#| label: fig-stray
of_scores <- oldfaithful |>
  mutate(stray = stray_scores(duration))
of_scores |> arrange(desc(stray))
of_scores |>
  ggplot(aes(x = duration, y = stray)) +
  geom_point()
```

@Talagala2019-zv, @Wilkinson2017-rb

## PCout

@Filzmoser2008-qa, @Filzmoser2013-tc

* mvoutlier::pcout

## FastPCS

@Vakili2014-yc

* FastPCS::FastPCS

## DOBIN

@Kandanaarachchi2020-lo

## Comparisons using the O3 plot
