# Functional data {#sec-functional}

```{r}
#| label: setup12
#| include: false
#| cache: false

source("before-each-chapter.R")
```

Functional data occur when the natural observational representation is a function.
The most commonly occurring examples are where observations are taken of functions over one or two dimensions.
For example, mortality rates are naturally represented as 1-dimensional functions of age, and surface temperatures are naturally represented as functions over the 2-dimensional plane of latitude and longitude.

@fig-fr-mortality shows French male mortality rates between 1816 and 1999.
Each line denotes the mortality rates as a function of age for one year, with the colours in rainbow order corresponding to the years of observation.
Overall, we see a large decrease in mortality rates during early childhood years, then an increase during teenage years.
After about age 30, the rates increase almost linearly on a log scale.
Comparing the curves over time, we see that the rates have steadily fallen for all ages, with nearly a 100-fold reduction in mortality rates at around age 10.
The noticeable gap between the curves occurred after World War II due to the effective use of antibiotics, and widespread vaccinations for polio, diphtheria, tetanus, whooping cough, measles, mumps and rubella.

```{r}
#| label: frmale
#| code-fold: false

fr_male <- fr_mortality |> filter(Sex == "Male") |> select(-Sex)
```

```{r}
#| label: fig-fr-mortality
#| fig-cap: French male mortality rates between 1816 and 1999.

fr_male |>
  ggplot() +
  aes(x = Age, y = log10(Mortality), colour = Year, group = Year) +
  geom_line() +
  scale_colour_gradientn(colours = rainbow(100)[1:80]) +
  labs(y = "Log (base 10) mortality rate") +
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ 10^(.),
      breaks = 10^(-(0:4)),
      labels = format(10^(-(0:4)), scientific = FALSE),
      name = "Mortality rate"
    )
  )
```

As we are interested in anomalies, we will mostly focus on the unusual curves.
There are about 10--15 years when the mortality rate for 20--40 year olds is much higher than in other years.
These are the war years.
Any anomaly detection method for functional data should identify these years as unusual.

We will denote the functional data as $y_i(x)$.
Here, $i$ is a discrete index over all observations, and $x$ is the continuous variable over which each function is defined.
For the mortality data, $y_{i}(x)$ represent the mortality rate for age $x$ in year $i$.

It is possible to have multiple indexes (e.g., mortality rates for males and females, or for different countries, or both), and the function may be defined over more than one variable (e.g., mortality rates over age and BMI).
It is also possible to have multivariate observations (e.g., mortality and fertility rates).

## Interpolation and smoothing {#sec-interpolation}

The distinguishing feature of functional data is that we think of the data as smooth and continuous functions.
In practice, our observations are usually at discrete values, even though the underlying functions are continuous.
In @fig-fr-mortality, for example, we observe the mortality rate for discrete ages 0, 1, 2, \dots, while we think of the rate as defined for any age $x \in [0,\infty)$.

It is common to form continuous functions using some kind of interpolation or smoothing of the raw data.
Smoothing is used if the underlying curve has been observed with measurement error.

The data shown in @fig-fr-mortality are observed with error, as the "wiggliness" of the curves shows.
The mortality rate (or central death rate) at age $x$ and year $t$ is defined as the number of people who died at age $x$ in year $t$ divided by the number of people who were aged $x$ in the middle of year $t$.
We can smooth the data to obtain the mortality rates at noninteger values of $x$.

In the following code, we use a smoothing method proposed by @HU07, that takes account of the characteristics of mortality rates.

```{r}
#| label: fr_smooth
#| message: false
#| code-fold: false

library(vital)
fr_smooth <- fr_male |>
  as_vital(index = Year, key = Age, .age = "Age") |>
  smooth_mortality(Mortality, k = 21)
```

@fig-fr-smooth shows the resulting smooth curve for one year along with the original data, while @fig-fr-smooth2 shows the smooth curves for all years.

```{r}
#| label: fig-fr-smooth
#| dependson: fr_smooth
#| fig-cap: Smoothed French male mortality rates for 1999.

fr_male |>
  filter(Year == 1999) |>
  ggplot(aes(x = Age, y = Mortality)) +
  geom_point() +
  geom_line(
    data = fr_smooth |> filter(Year == 1999),
    aes(y = .smooth)
  ) +
  scale_y_log10(
    breaks = 10^(-(0:4)),
    labels = format(10^(-(0:4)), scientific = FALSE)
  ) +
  labs(y = "Mortality rate")
```

```{r}
#| label: fig-fr-smooth2
#| dependson: fr_smooth
#| fig-cap: Smoothed French male mortality rates for all years.

fr_smooth |>
  ggplot() +
  aes(x = Age, y = log10(.smooth), colour = Year, group = Year) +
  geom_line() +
  scale_colour_gradientn(colours = rainbow(100)[1:80]) +
  labs(y = "Log mortality rate") +
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ 10^(.),
      breaks = 10^(-(0:4)),
      labels = format(10^(-(0:4)), scientific = FALSE),
      name = "Mortality rate"
    )
  )
```

Working with the smoothed curves rather than the raw mortality rates has several advantages for anomaly detection.
Smoothing removes measurement noise that otherwise be mistaken for genuine functional variation, so isolated noisy observations are not flagged as anomalies, and real anomalies are not masked by noise.
It also gives us a curve that can be evaluated, and differentiated, at any age, which is needed for the basis function and depth-based methods used later in this chapter.
Finally, smoothing borrows strength across neighbouring ages, using the fact that mortality varies smoothly with age to produce more stable and reliable estimates than the raw data alone.

## Dimension reduction using basis functions {#sec-basis-functions}

We can approximate functional data using a linear combination of basis functions:
$$
  \hat{y}_i(x) = b_0(x) + \sum_{j = 1}^J \beta_{ij} b_j(x),
$$
where $b_0(x)$ is a constant function, $\{b_1(x),\dots,b_J(x)\}$ are a set of basis functions, and $\beta_{ij}$ are the coefficients.
The basis functions span the space of all functions.
For example, the Fourier basis functions are a set of sine and cosine functions that can be used to represent any periodic function.
The B-spline basis functions are a set of piecewise polynomial functions that can be used to represent any smooth function.
The choice of basis functions depends on the characteristics of the data.

The number of basis functions $J$ is usually much smaller than the number of observations, with larger $J$ corresponding to a more flexible approximation.

Given a set of basis functions, we can estimate the coefficients $\beta_{ij}$ by minimising the sum of squared errors between the observed data $y_i(x)$ and the approximated data $\hat{y}_i(x)$.

Alternatively, we can use functional principal component analysis to find the empirical basis functions that explain the most variance in the data.
Then $b_0(x)$ is set to be the mean function,
$$
  b_0(x) = \frac{1}{n} \sum_{i = 1}^n y_i(x),
$$
and the first principal component, $b_1(x)$, is the function that minimises
$$
  \big[y_i(x) - b_0(x) - \beta_{i1}b_1(x)\big]^2
$$
subject to the constraint that $\int b_1^2(x)dx = 1$.
The second principal component, $b_2(x)$, is the function that minimises
$$
  \big[y_i(x) - b_0(x) - \beta_{i1}b_1(x) - \beta_{i2}b_2(x)\big]^2
$$
subject to the constraints that $\int b_2^2(x)dx = 1$ and $\int b_1(x)b_2(x)dx = 0$.
And so on.

If the functions are discretised on a fine grid over $x$, we can use regular principal component decomposition to get approximately the same results as functional principal components.

In the following example, we find the first few principal components (empirical basis functions) using the French male mortality data.

```{r}
#| echo: true
#| code-fold: false
#| message: false
#| label: pca-bases

# Wide version of smoothed log Mortality with ages on columns
frmort_wide <- fr_smooth |>
  as_tibble() |>
  mutate(logmx = log(.smooth)) |>
  select(Year, Age, logmx) |>
  tidyr::pivot_wider(
    names_from = Age,
    values_from = logmx,
    names_prefix = "Age"
  )

# Compute first three principal components
pca <- frmort_wide |>
  select(-Year) |>
  prcomp(center = TRUE, scale = FALSE, rank = 3)
summary(pca)
```

The first three principal components explain `r round(100 * sum(pca$sdev[1:3]^2) / sum(pca$sdev^2), 1)`% of the variation in the data.
Instead of working in the 86 dimensional space containing all ages, we can work in the 3-dimensional space of the first three principal component scores, which is a much simpler problem.

```{r}
#| echo: true
#| code-fold: false
#| message: false
#| label: fig-pca-bases
#| dependson: pca-bases
#| fig-cap: First three principal components (empirical basis functions) for the French male mortality data.

pca_bases <- broom::tidy(pca, "loadings") |>
  mutate(
    Age = readr::parse_number(column),
    PC = as.factor(PC)
  )
pca_bases |>
  ggplot() +
  aes(x = Age, y = value, group = PC, colour = PC) +
  geom_line()
```
@fig-pca-bases shows the first three empirical basis functions.

1. The first basis function is negative at every age, and largest in magnitude in early childhood, so a high score on this component corresponds to lower mortality at all ages, but especially for children.
2. The second basis function contrasts men of fighting age (roughly ages 18--40) against children, so it picks out years where mortality among young men was unusually high (or low) relative to mortality in childhood.
3. The third basis function contrasts infant mortality and older mortality (above age 40), with mortality of pre-teens and early 20s.

```{r}
#| echo: true
#| code-fold: false
#| message: false
#| label: fig-pca-scores
#| dependson: pca-bases
#| fig-cap: First two principal component scores for the French male mortality data.

# Compute principal component scores
pca_scores <- broom::augment(pca, frmort_wide[, "Year"]) |>
  select(-.rownames) |>
  rename_with(~ gsub("\\.fitted", "", .x))
# Time series plot of scores
pca_scores |>
  tidyr::pivot_longer(
    -Year,
    names_to = "PC",
    values_to = "value"
  ) |>
  ggplot(aes(x = Year, y = value)) +
  geom_line(aes(colour = PC))
```

@fig-pca-scores shows how the two component scores evolve over time.
Other than during the two world wars, the first component score increases steadily and fairly smoothly across the study period, reflecting the long-run, near-continuous decline in mortality at all ages.
The second component score sits close to zero for most years but spikes sharply during 1914--1918 and 1940--1945, capturing the additional mortality among young French men during the two world wars.
The third component is more subtle, and reflects slow changes in mortality patterns over time, rather than the sharp spikes seen in the second component.

```{r}
#| echo: true
#| code-fold: false
#| message: false
#| label: fig-pca-scores2
#| fig-cap: Scatterplot of first two principal component scores for the French male mortality data.

pca_scores |>
  select(-Year) |>
  GGally::ggpairs()
```

@fig-pca-scores2 plots the pairwise scatterplots of the first three component scores against each other.
Most years fall along a narrow band in each plot, but there are some clear anomalies, especially in the second component.

This low-dimensional view foreshadows the anomaly detection approach of the next section, where we treat points that lie far from the bulk of the data in this principal component space as candidate anomalies.

## Anomalies in the principal component space {#sec-functional-pca}

Once the functional data have been reduced to a handful of principal component scores, each curve (here, each year) is just a point in a low-dimensional numerical space, and we can apply the same methods introduced previously in this book.
For example, we can compute a surprisal probability for each year from the first two component scores, using a GPD approximation for the tail (@sec-approximate-surprisal-prob), and flag any year with a probability below 0.05 as anomalous.
We restrict attention to the first two components here so the results can be shown more easily in a scatterplot; we return to all three components below using the O3 plot.

```{r}
#| echo: true
#| code-fold: false
#| label: fig-pca-outliers
#| dependson: fig-pca-scores
#| fig-cap: French principal component scores, with anomalous years (surprisal probability below 0.05) shown in red.

# Find outliers in the PCs
outliers <- pca_scores |>
  mutate(prob = surprisals_prob(pick(PC1, PC2), approximation = "gpd")) |>
  filter(prob < 0.05)

pca_scores |>
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point() +
  geom_point(data = outliers, colour = "red") +
  ggrepel::geom_label_repel(data = outliers, aes(label = Year))
```

```{r}
#| label: format-years
#| echo: false

# Format a vector of years as individual years and ranges (3+ consecutive years)
format_years <- function(years) {
  years <- sort(unique(years))
  grp <- cumsum(c(1L, diff(years) != 1L))
  parts <- unlist(lapply(
    split(years, grp),
    \(g) {
      if (length(g) >= 3) paste0(g[1], "--", g[length(g)]) else as.character(g)
    }
  ))
  if (length(parts) > 1) {
    paste(
      paste(parts[-length(parts)], collapse = ", "),
      "and",
      parts[length(parts)]
    )
  } else {
    parts
  }
}
```

@fig-pca-outliers identifies `r english::words(nrow(outliers))` anomalous years: `r format_years(outliers$Year)`.
These are unsurprising given @fig-fr-mortality, and most coincide with years already flagged in @sec-weird-times using the Hampel identifier applied directly to the raw age-specific series --- a quite different method, applied to the raw data rather than to PCA scores.
The following historical events are relevant:

- 1871: Repression of the 'Commune de Paris'
- 1914--1918: World War I
- 1919: Spanish flu pandemic
- 1940--1944: World War II

To take another perspective, we can apply the O3 plot (@sec-o3) to the first three principal component scores.

```{r}
#| label: pca_o3prep
#| dependson: fig-pca-outliers
#| code-fold: false

o3 <- pca_scores |>
  select(-Year) |>
  OutliersO3::O3prep(
    method = c("PCS", "adjOut", "MCD", "BAC", "DDC"),
    tolPCS = 0.05,
    toladj = 0.15,
    tolMCD = 0.001,
    tolBAC = 0.001,
    tolDDC = 0.025
  )
o3_plots <- OutliersO3::O3plotM(
  o3,
  coltxtsize = 10,
  caseNames = paste0("Y", pca_scores$Year)
)
```

```{r}
#| label: fig-o3-pca
#| fig-width: 14
#| fig-asp: 0.2
#| dependson: pca_o3prep
#| code-fold: false

o3_plots$gO3
```

The years flagged at least 7 times across all five methods and variable combinations are given below.

```{r}
#| label: o3-outliers
#| dependson: pca_o3prep

red_case <- o3_plots$outsTable |>
  distinct(Case, Combination, Method) |>
  count(Case, Combination, name = "n_agree") |>
  filter(n_agree == 5L)
out_table <- o3_plots$outsTable |>
  count(Case, name = "n_flagged") |>
  filter(n_flagged > 6) |>
  mutate(red = Case %in% red_case$Case)
pca_scores |>
  slice(out_table$Case) |>
  mutate(n_flagged = out_table$n_flagged, red = out_table$red) |>
  arrange(Year)
```

```{r}
#| label: o3-summary
#| echo: false
#| dependson: o3-outliers

o3_years <- sort(pca_scores$Year[out_table$Case])
o3_red_years <- sort(pca_scores$Year[out_table$Case[out_table$red]])
n_pc_o3 <- ncol(pca_scores) - 1
n_pc_surprisal <- pca_scores |> select(-Year, -PC3) |> ncol()
n_methods <- length(unique(o3_plots$outsTable$Method))
same_years <- setequal(outliers$Year, o3_years)
```

Reassuringly, the O3 plot, using all `r english::words(n_pc_o3)` principal components and `r english::words(n_methods)` quite different outlier-detection methods, flags `r english::words(length(o3_years))` years (`r format_years(o3_years)`) --- `r if (same_years) "exactly the same years" else "largely the same years"` identified by the two-component surprisal approach above.
Of these, `r english::words(length(o3_red_years))` --- `r format_years(o3_red_years)` --- are flagged by every method and variable combination (shown in red in @fig-o3-pca), while the rest are flagged consistently but not unanimously.
This agreement across methods, and with the time series analysis of @sec-weird-times, gives us some confidence that these years are genuine anomalies rather than artefacts of any one method.

## Functional bagplot {#sec-functional-bagplot}

A boxplot summarises a univariate sample using a central box and whiskers, with points beyond the whiskers flagged as outliers.
Three different ways of extending this idea to functional data are in common use here.
The functional bagplot and the functional HDR boxplot, covered in this section and the next, were proposed by @HS08: reduce each curve to its first two functional principal component scores (@sec-basis-functions), draw the ordinary bivariate bagplot or HDR boxplot of those scores (@sec-bagplots, @sec-hdr-boxplots), and then map the boundary of the resulting central region back through the principal component basis to obtain an envelope of curves.
A third approach, the functional depth boxplot (@sec-functional-depth), instead works directly with the modified band depth of the curves, without going through a low-dimensional projection first.

For the two PCA-based versions, we need a way to turn a point $(s_1, s_2)$ in principal component score space back into a curve.
Using the notation of @sec-basis-functions, this is just
$$
  \hat y(x) = b_0(x) + s_1 b_1(x) + s_2 b_2(x),
$$
the mean function plus a weighted combination of the first two empirical basis functions.

Before mapping anything back to curve space, it helps to see the bagplot of the scores themselves.
@fig-pca-bagplot shows the ordinary bivariate bagplot of the first two principal component scores.
Each point is a year; the dark region is the bag, the light region is the loop, and any points beyond the loop are flagged as bivariate outliers.

```{r}
#| label: fig-pca-bagplot
#| dependson: pca-reconstruct
#| code-fold: false
#| fig-cap: Bagplot of the first two principal component scores of the French male mortality curves.

pca_scores |> gg_bagplot(PC1, PC2)
```

Every point in @fig-pca-bagplot corresponds to a whole mortality curve, via the basis reconstruction above, so the bag and loop shown here trace out regions of "typical" curves, while the flagged points correspond to years whose curves depart from that typical shape.
Mapping the bag and loop boundaries back through the basis functions, rather than just the flagged points, is what turns this ordinary bagplot into a functional bagplot.

The following code identifies the curves contained in the bag and the loop, and those beyond that loop corresponding to the anomalies in @fig-pca-bagplot.

```{r}
#| label: pca-reconstruct
#| dependson: fig-pca-scores
#| code-fold: true

reconstruct_curve <- function(s1, s2, pca) {
  p <- length(pca$center)
  matrix(pca$center, nrow = length(s1), ncol = p, byrow = TRUE) +
    outer(s1, pca$rotation[, 1]) +
    outer(s2, pca$rotation[, 2])
}
envelope <- function(hull, age, pca) {
  curves <- reconstruct_curve(hull[, 1], hull[, 2], pca)
  tibble(Age = age, ymin = apply(curves, 2, min), ymax = apply(curves, 2, max))
}
classify_bag_region <- function(scores, bag) {
  match_rows <- function(target) {
    if (NROW(target) == 0) {
      return(integer(0))
    }
    apply(as.matrix(target), 1, function(r) {
      which(scores$PC1 == r[1] & scores$PC2 == r[2])
    }) |>
      unlist()
  }
  region <- rep("bag", nrow(scores))
  region[match_rows(bag$pxy.outer)] <- "loop"
  region[match_rows(bag$pxy.outlier)] <- "outlier"
  region
}
```

```{r}
#| label: bagplot-scores
#| dependson: pca-reconstruct
#| code-fold: false

bag <- aplpack::compute.bagplot(as.matrix(pca_scores[, c("PC1", "PC2")]))
pca_scores <- pca_scores |>
  mutate(bag_region = classify_bag_region(pca_scores, bag))
bagplot_outliers <- pca_scores |> filter(bag_region == "outlier")
```

Mapping the bag and loop boundary vertices back to curves with `reconstruct_curve()`, and taking the pointwise range of each set of boundary curves, gives an envelope for the "typical" region of the functional bagplot at every age.

```{r}
#| label: fig-fbagplot
#| dependson: bagplot-scores
#| fig-cap: Functional bagplot of French male mortality.
#|   The dark band shows the envelope of the bag (50% region), the light band shows the envelope of the loop; the raw curves for years flagged as bagplot outliers are overlaid.

age <- as.numeric(gsub("Age", "", rownames(pca$rotation)))
bag_env <- envelope(bag$hull.bag, age, pca)
loop_env <- envelope(bag$hull.loop, age, pca)

outlier_curves <- fr_male |>
  filter(Year %in% bagplot_outliers$Year) |>
  mutate(logmx = log(Mortality))

ggplot() +
  geom_ribbon(
    data = loop_env,
    aes(x = Age, ymin = ymin, ymax = ymax),
    fill = "grey80"
  ) +
  geom_ribbon(
    data = bag_env,
    aes(x = Age, ymin = ymin, ymax = ymax),
    fill = "grey50"
  ) +
  geom_line(
    data = outlier_curves,
    aes(x = Age, y = logmx, group = Year, colour = factor(Year))
  ) +
  labs(y = "Log mortality rate", colour = "Outlier year")
```

@fig-fbagplot flags `r english::words(nrow(bagplot_outliers))` years as functional bagplot outliers: `r format_years(bagplot_outliers$Year)`.
These are almost exactly the war years already identified in @sec-functional-pca and @sec-weird-times, and the outlying curves are clearly visible bulging above the typical envelope for young men --- exactly where the second principal component loads most heavily (@fig-pca-bases).

## Functional HDR boxplot {#sec-functional-hdr-boxplot}

The functional HDR boxplot follows the same recipe, but uses the highest density regions of a bivariate kernel density estimate of the scores (@sec-mhdr) in place of the bagplot's bag and loop.
We use `hdrcde::hdr.2d()`, which returns the density estimated at each data point (`$fxy`) together with the density threshold corresponding to each requested coverage probability (`$falpha`).
Note that `hdrcde` labels these thresholds by the tail probability $1 - p$ rather than the coverage $p$, so the threshold for the 99% HDR is named `"1%"`.

```{r}
#| label: hdr-scores
#| dependson: pca-reconstruct
#| code-fold: false
#| message: false

h <- hdrcde::hdr.2d(
  pca_scores$PC1,
  pca_scores$PC2,
  prob = c(50, 99),
  kde.package = "ks"
)
pca_scores <- pca_scores |>
  mutate(
    hdr_region = case_when(
      h$fxy >= h$falpha["50%"] ~ "50%",
      h$fxy >= h$falpha["1%"] ~ "99%",
      TRUE ~ "outlier"
    )
  )
hdr_outliers <- pca_scores |> filter(hdr_region == "outlier")
```

As with the bagplot, we extract the boundary of each HDR as a contour of the density grid, and map it back to a curve envelope.
The 99% contour is sometimes broken into several disconnected pieces by the kernel density estimate; we keep only the largest piece, as the others are small artefacts of the grid used to estimate the density.

```{r}
#| label: fig-fhdrboxplot
#| dependson: hdr-scores
#| fig-cap: Functional HDR boxplot of French male mortality.
#|   The dark band shows the envelope of the 50% HDR, the light band the 99% HDR; the raw curves for years flagged as outliers (falling outside the 99% HDR) are overlaid.

largest_piece <- function(contour_lines) {
  contour_lines[[which.max(sapply(contour_lines, \(piece) length(piece$x)))]]
}
contour50 <- largest_piece(
  grDevices::contourLines(h$den$x, h$den$y, h$den$z, levels = h$falpha["50%"])
)
contour99 <- largest_piece(
  grDevices::contourLines(h$den$x, h$den$y, h$den$z, levels = h$falpha["1%"])
)
hdr50_env <- envelope(cbind(contour50$x, contour50$y), age, pca)
hdr99_env <- envelope(cbind(contour99$x, contour99$y), age, pca)

outlier_curves <- fr_male |>
  filter(Year %in% hdr_outliers$Year) |>
  mutate(logmx = log(Mortality))

ggplot() +
  geom_ribbon(
    data = hdr99_env,
    aes(x = Age, ymin = ymin, ymax = ymax),
    fill = "grey80"
  ) +
  geom_ribbon(
    data = hdr50_env,
    aes(x = Age, ymin = ymin, ymax = ymax),
    fill = "grey50"
  ) +
  geom_line(
    data = outlier_curves,
    aes(x = Age, y = logmx, group = Year, colour = factor(Year))
  ) +
  labs(y = "Log mortality rate", colour = "Outlier year")
```

@fig-fhdrboxplot is more conservative than the functional bagplot, flagging only `r english::words(nrow(hdr_outliers))` years, `r format_years(hdr_outliers$Year)`, as lying outside the 99% HDR of the score distribution.
Both are among the years already flagged by every method used so far in this chapter.

## Functional depth boxplot {#sec-functional-depth}

There are several other approaches to identifying functional outliers, summarised in @hubert2015multivariate, almost all based on functional depth.

The idea of depth generalises the univariate idea of a median and quantiles to more complex objects.
We have already seen this with multivariate numeric data (@sec-depth); now we will extend the idea to functional data.
Given a sample of curves $y_1(x),\dots,y_n(x)$, a depth measure assigns each curve $y_i$ a value between 0 and 1 describing how central it is within the sample: curves near the middle of the distribution receive high depth, while unusual curves receive low depth.
Curves can then be ranked from the deepest (most typical) to the shallowest (most unusual), giving an ordering analogous to ranking univariate observations from the median outward.

One of the simplest depths is the band depth: it counts how often a curve $y_i$ lies completely within the "band" traced out by every possible group of $J$ other curves in the sample, so that curves often enveloped by others are deep, while curves that stick out --- even briefly --- have low depth.
Looking at @fig-fr-mortality, we would expect that some of the most recent years, which have low mortality at almost every age, would be shallow; the war years would also be shallow, since they have high mortality at ages 20--40 that is not seen in other years.

Because the exact band depth is expensive to compute for anything beyond a handful of curves, in practice it is usually replaced by the more tractable modified band depth (MBD), which instead measures the proportion of the domain $x$ over which $y_i(x)$ lies within each band, averaged over all bands [@LopezPintado2009].
MBD, and the other depth-based methods used in the rest of this chapter, are available in the `roahd` package, which we use from here on.
MBD is also the depth measure underlying the classic functional boxplot introduced below.

```{r}
#| label: fdata-build
#| dependson: fr_smooth
#| code-fold: false
#| message: false

library(roahd)
fr_fd <- fData(sort(unique(fr_smooth$Age)), as.matrix(frmort_wide[, -1]))

depth_tbl <- tibble(
  Year = frmort_wide$Year,
  MBD = MBD(fr_fd)
)
depth_tbl |> arrange(MBD)
```

```{r}
#| label: shallowyears
#| echo: false

shallow_years <- depth_tbl |>
  arrange(MBD) |>
  slice_head(n = 10) |>
  pull(Year)
shallow_1800s <- shallow_years[shallow_years < 1900] |> sort()
stopifnot(identical(shallow_1800s, c(1854L, 1870L, 1871L)))
```

Ranking curves by MBD alone identifies the years whose entire trajectory lies furthest from the bulk of the sample: the shallowest curves are `r format_years(shallow_years)`, whose mortality is lower at almost every age than in any other year.
Those from the 1800s correspond to the 1854 cholera outbreak, the 1870 Franco-Prussian war and the 1871 Repression of the 'Commune de Paris'.

This illustrates an important feature of depth-based ranking: because French mortality has fallen more or less steadily across the study period, the most recent curves sit at one edge of the whole 184-year sample and are therefore the least "enveloped" by the rest, even though nothing untoward happened in those years.
Depth ranking on its own cannot distinguish a genuinely anomalous curve from one that simply lies at the extreme end of a long-running, gradual trend.

The classic functional boxplot of @Sun2011 works directly with the smoothed curves via `roahd::fbplot()`, using the modified band depth introduced above rather than a 2-dimensional score projection.
The central region is formed from the 50% deepest curves, and the fence is this region inflated by a factor (1.5, by default) of its own range at each age; any curve that leaves the fence at any age is flagged as an outlier.

```{r}
#| label: fig-fbplot-sungenton
#| dependson: fdata-build
#| fig-cap: Sun–Genton functional boxplot of the smoothed French male mortality curves.
#|   The dark band shows the central 50% region, the light band the fence; the raw curves for the flagged outlier years are overlaid.

fb <- fbplot(fr_fd, display = FALSE)
fb_outlier_years <- frmort_wide$Year[fb$ID_outliers]

central <- fr_fd$values[
  order(-fb$Depth)[seq_len(round(nrow(fr_fd$values) / 2))],
  ,
  drop = FALSE
]
central_range <- apply(central, 2, range)
fence_width <- fb$Fvalue * (central_range[2, ] - central_range[1, ])
fb_env <- tibble(
  Age = sort(unique(fr_smooth$Age)),
  cmin = central_range[1, ],
  cmax = central_range[2, ],
  fmin = central_range[1, ] - fence_width,
  fmax = central_range[2, ] + fence_width
)

outlier_curves <- fr_smooth |>
  as_tibble() |>
  filter(Year %in% fb_outlier_years) |>
  mutate(loglog = log(.smooth))

ggplot() +
  geom_ribbon(
    data = fb_env,
    aes(x = Age, ymin = fmin, ymax = fmax),
    fill = "grey80"
  ) +
  geom_ribbon(
    data = fb_env,
    aes(x = Age, ymin = cmin, ymax = cmax),
    fill = "grey50"
  ) +
  geom_line(
    data = outlier_curves,
    aes(x = Age, y = loglog, group = Year, colour = factor(Year))
  ) +
  labs(y = "Log mortality rate", colour = "Outlier year")
```

@fig-fbplot-sungenton tells a different story from the previous two figures: it flags `r english::words(length(fb_outlier_years))` years, `r format_years(fb_outlier_years)`, none of which are war years.
These are simply the most recent years in the sample, whose mortality is lower at almost every age than any earlier year --- the same phenomenon already noted for the raw MBD ranking above.
This is a useful reminder that a functional boxplot, like MBD ranking, measures how atypical a whole curve is relative to the rest of the sample, not whether anything unusual happened in a particular year; it will always flag the edges of a strong, gradual trend as "outliers", even though the war years remain the more interesting anomalies from a historical point of view.

The functional bagplot and HDR boxplot do not share this issue, because reducing each curve to two principal component scores loses most of the information about the gradual, whole-of-curve trend captured by MBD, while retaining the specific age-pattern (young men versus children) that made the war years unusual.
Together, the methods introduced in this chapter --- surprisal probabilities and the O3 plot in principal component space, and the three functional boxplots --- agree closely on the years that matter historically, while disagreeing, instructively, on how to treat the long-run decline in mortality itself.
