# Functional data {#sec-functional}

```{r}
#| label: setup12
#| include: false
#| cache: false

source("before-each-chapter.R")
```

Functional data occur when the natural observational representation is a function.
The most commonly occurring examples are where observations are taken of functions over one or two dimensions.
For example, mortality rates are naturally represented as 1-dimensional functions of age, and surface temperatures are naturally represented as functions over the 2-dimensional plane of latitude and longitude.

@fig-fr-mortality shows French male mortality rates between 1816 and 1999.
Each line denotes the mortality rates as a function of age for one year, with the colours in rainbow order corresponding to the years of observation.
Overall, we see a large decrease in mortality rates during early childhood years, then an increase during teenage years.
After about age 30, the rates increase almost linearly on a log scale.
Comparing the curves over time, we see that the rates have steadily fallen for all ages, with nearly a 100-fold reduction in mortality rates at around age 10.
The noticeable gap between the curves occurred after World War II due to the effective use of antibiotics, and widespread vaccinations for polio, diphtheria, tetanus, whooping cough, measles, mumps and rubella.

```{r}
#| label: fig-fr-mortality
#| fig-cap: French male mortality rates between 1816 and 1999.

fr_mortality |>
  filter(Sex == "Male") |>
  ggplot() +
  aes(x = Age, y = log10(Mortality), colour = Year, group = Year) +
  geom_line() +
  scale_colour_gradientn(colours = rainbow(100)[1:80]) +
  labs(y = "Log (base 10) mortality rate") +
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ 10^(.),
      breaks = 10^(-(0:4)),
      labels = format(10^(-(0:4)), scientific = FALSE),
      name = "Mortality rate"
    )
  )
```

As we are interested in anomalies, we will mostly focus on the unusual curves.
There are about 10--15 years when the mortality rate for 20--40 year olds is much higher than in other years.
These are the war years.
Any anomaly detection method for functional data should identify these years as unusual.

We will denote the functional data as $y_i(x)$.
Here, $i$ is a discrete index over all observations, and $x$ is the continuous variable over which each function is defined.
For the mortality data, $y_{i}(x)$ represent the mortality rate for age $x$ in year $i$.

It is possible to have multiple indexes (e.g., mortality rates for males and females, or for different countries, or both), and the function may be defined over more than one variable (e.g., mortality rates over age and BMI).
It is also possible to have multivariate observations (e.g., mortality and fertility rates).

## Interpolation and smoothing {#sec-interpolation}

The distinguishing feature of functional data is that we think of the data as smooth and continuous functions.
In practice, our observations are usually at discrete values, even though the underlying functions are continuous.
In @fig-fr-mortality, for example, we observe the mortality rate for discrete ages 0, 1, 2, \dots, while we think of the rate as defined for any age $x \in [0,\infty)$.

It is common to form continuous functions using some kind of interpolation or smoothing of the raw data.
Smoothing is used if the underlying curve has been observed with measurement error.

The data shown in @fig-fr-mortality are observed with error, as the "wiggliness" of the curves shows.
The mortality rate (or central death rate) at age $x$ and year $t$ is defined as the number of people who died at age $x$ in year $t$ divided by the number of people who were aged $x$ in the middle of year $t$.
We can smooth the data to obtain the mortality rates at noninteger values of $x$.

In the following code, we use a smoothing method proposed by @HU07, that takes account of the characteristics of mortality rates.

```{r}
#| label: fr_smooth
#| message: false
#| code-fold: false

library(vital)
fr_smooth <- fr_mortality |>
  as_vital(index = Year, key = c(Age, Sex), .age = "Age", .sex = "Sex") |>
  smooth_mortality(Mortality, k = 21)
```

@fig-fr-smooth shows the resulting smooth curve for one year along with the original data, while @fig-fr-smooth2 shows the smooth curves for all years.

```{r}
#| label: fig-fr-smooth
#| dependson: fr_smooth
#| fig-cap: Smoothed French male mortality rates for 1999.

fr_mortality |>
  filter(Year == 1999, Sex == "Male") |>
  ggplot(aes(x = Age, y = Mortality)) +
  geom_point() +
  geom_line(
    data = fr_smooth |> filter(Year == 1999, Sex == "Male"),
    aes(y = .smooth)
  ) +
  scale_y_log10(
    breaks = 10^(-(0:4)),
    labels = format(10^(-(0:4)), scientific = FALSE)
  ) +
  labs(y = "Mortality rate")
```

```{r}
#| label: fig-fr-smooth2
#| dependson: fr_smooth
#| fig-cap: Smoothed French male mortality rates for all years.

fr_smooth |>
  filter(Sex == "Male") |>
  ggplot() +
  aes(x = Age, y = log10(.smooth), colour = Year, group = Year) +
  geom_line() +
  scale_colour_gradientn(colours = rainbow(100)[1:80]) +
  labs(y = "Log mortality rate") +
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ 10^(.),
      breaks = 10^(-(0:4)),
      labels = format(10^(-(0:4)), scientific = FALSE),
      name = "Mortality rate"
    )
  )
```

Working with the smoothed curves rather than the raw mortality rates has several advantages for anomaly detection.
Smoothing removes measurement noise that otherwise be mistaken for genuine functional variation, so isolated noisy observations are not flagged as anomalies, and real anomalies are not masked by noise.
It also gives us a curve that can be evaluated, and differentiated, at any age, which is needed for the basis function and depth-based methods used later in this chapter.
Finally, smoothing borrows strength across neighbouring ages, using the fact that mortality varies smoothly with age to produce more stable and reliable estimates than the raw data alone.

## Dimension reduction using basis functions {#sec-basis-functions}

We can approximate functional data using a linear combination of basis functions:
$$
  \hat{y}_i(x) = b_0(x) + \sum_{j = 1}^J \beta_{ij} b_j(x),
$$
where $b_0(x)$ is a constant function, $\{b_1(x),\dots,b_J(x)\}$ are a set of basis functions, and $\beta_{ij}$ are the coefficients.
The basis functions span the space of all functions.
For example, the Fourier basis functions are a set of sine and cosine functions that can be used to represent any periodic function.
The B-spline basis functions are a set of piecewise polynomial functions that can be used to represent any smooth function.
The choice of basis functions depends on the characteristics of the data.

The number of basis functions $J$ is usually much smaller than the number of observations, with larger $J$ corresponding to a more flexible approximation.

Given a set of basis functions, we can estimate the coefficients $\beta_{ij}$ by minimising the sum of squared errors between the observed data $y_i(x)$ and the approximated data $\hat{y}_i(x)$.

Alternatively, we can use functional principal component analysis to find the empirical basis functions that explain the most variance in the data.
Then $b_0(x)$ is set to be the mean function,
$$
  b_0(x) = \frac{1}{n} \sum_{i = 1}^n y_i(x),
$$
and the first principal component, $b_1(x)$, is the function that minimises
$$
  \big[y_i(x) - b_0(x) - \beta_{i1}b_1(x)\big]^2
$$
subject to the constraint that $\int b_1^2(x)dx = 1$.
The second principal component, $b_2(x)$, is the function that minimises
$$
  \big[y_i(x) - b_0(x) - \beta_{i1}b_1(x) - \beta_{i2}b_2(x)\big]^2
$$
subject to the constraints that $\int b_2^2(x)dx = 1$ and $\int b_1(x)b_2(x)dx = 0$.
And so on.

If the functions are discretised on a fine grid over $x$, we can use regular principal component decomposition to get approximately the same results as functional principal components.

In the following example, we find the first few principal components (empirical basis functions) using the French male mortality data.

```{r}
#| echo: true
#| code-fold: false
#| message: false
#| label: pca-bases

# Wide version of log Mortality with ages on columns
frmort_wide <- fr_mortality |>
  filter(Sex == "Male") |>
  mutate(logmx = log(Mortality)) |>
  select(-Mortality, -Sex) |>
  tidyr::pivot_wider(
    names_from = Age,
    values_from = logmx,
    names_prefix = "Age"
  )

# Compute first three principal components
pca <- frmort_wide |>
  select(-Year) |>
  prcomp(center = TRUE, scale = FALSE, rank = 3)
summary(pca)
```

The first three principal components explain `r round(100 * sum(pca$sdev[1:3]^2) / sum(pca$sdev^2), 1)`% of the variation in the data.
Instead of working in the 86 dimensional space containing all ages, we can work in the 3-dimensional space of the first three principal component scores, which is a much simpler problem.

```{r}
#| echo: true
#| code-fold: false
#| message: false
#| label: fig-pca-bases
#| dependson: pca-bases
#| fig-cap: First three principal components (empirical basis functions) for the French male mortality data.

pca_bases <- broom::tidy(pca, "loadings") |>
  mutate(
    Age = readr::parse_number(column),
    PC = as.factor(PC)
  )
pca_bases |>
  ggplot() +
  aes(x = Age, y = value, group = PC, colour = PC) +
  geom_line()
```
@fig-pca-bases shows the first three empirical basis functions.

1. The first basis function is negative at every age, and largest in magnitude in early childhood, so a high score on this component corresponds to lower mortality at all ages, but especially for children.
2. The second basis function contrasts men of fighting age (roughly ages 18--40) against children, so it picks out years where mortality among young men was unusually high (or low) relative to mortality in childhood.
3. The third basis function contrasts infant mortality and older mortality (above age 40), with mortality of pre-teens and early 20s.

```{r}
#| echo: true
#| code-fold: false
#| message: false
#| label: fig-pca-scores
#| dependson: pca-bases
#| fig-cap: First two principal component scores for the French male mortality data.

# Compute principal component scores
pca_scores <- broom::augment(pca, frmort_wide[, "Year"]) |>
  select(-.rownames) |>
  rename_with(~ gsub("\\.fitted", "", .x))
# Time series plot of scores
pca_scores |>
  tidyr::pivot_longer(
    -Year,
    names_to = "PC",
    values_to = "value"
  ) |>
  ggplot(aes(x = Year, y = value)) +
  geom_line(aes(colour = PC))
```

@fig-pca-scores shows how the two component scores evolve over time.
Other than during the two world wars, the first component score increases steadily and fairly smoothly across the study period, reflecting the long-run, near-continuous decline in mortality at all ages.
The second component score sits close to zero for most years but spikes sharply during 1914--1918 and 1940--1945, capturing the additional mortality among young French men during the two world wars.
The third component is more subtle, and reflects slow changes in mortality patterns over time, rather than the sharp spikes seen in the second component.

```{r}
#| echo: true
#| code-fold: false
#| message: false
#| label: fig-pca-scores2
#| fig-cap: Scatterplot of first two principal component scores for the French male mortality data.

pca_scores |>
  select(-Year) |>
  GGally::ggpairs()
```

@fig-pca-scores2 plots the pairwise scatterplots of the first three component scores against each other.
Most years fall along a narrow band in each plot, but there are some clear anomalies, especially in the second component.

This low-dimensional view foreshadows the anomaly detection approach of the next section, where we treat points that lie far from the bulk of the data in this principal component space as candidate anomalies.

## Anomalies in the principal component space {#sec-functional-pca}

Once the functional data have been reduced to a handful of principal component scores, each curve (here, each year) is just a point in a low-dimensional numerical space, and we can apply the same methods introduced previously in this book.
For example, we can compute a surprisal probability for each year from the first two component scores, using a GPD approximation for the tail (@sec-approximate-surprisal-prob), and flag any year with a probability below 0.05 as anomalous.
We restrict attention to the first two components here so the results can be shown more easily in a scatterplot; we return to all three components below using the O3 plot.

```{r}
#| echo: true
#| code-fold: false
#| label: fig-pca-outliers
#| dependson: fig-pca-scores
#| fig-cap: French principal component scores, with anomalous years (surprisal probability below 0.05) shown in red.

# Find outliers in the PCs
outliers <- pca_scores |>
  mutate(prob = surprisals_prob(pick(PC1, PC2), approximation = "gpd")) |>
  filter(prob < 0.05)

pca_scores |>
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point() +
  geom_point(data = outliers, colour = "red") +
  ggrepel::geom_label_repel(data = outliers, aes(label = Year))
```

```{r}
#| label: format-years
#| echo: false

# Format a vector of years as individual years and ranges (3+ consecutive years)
format_years <- function(years) {
  years <- sort(unique(years))
  grp <- cumsum(c(1L, diff(years) != 1L))
  parts <- unlist(lapply(
    split(years, grp),
    \(g) {
      if (length(g) >= 3) paste0(g[1], "--", g[length(g)]) else as.character(g)
    }
  ))
  if (length(parts) > 1) {
    paste(
      paste(parts[-length(parts)], collapse = ", "),
      "and",
      parts[length(parts)]
    )
  } else {
    parts
  }
}
```

@fig-pca-outliers identifies `r english::words(nrow(outliers))` anomalous years: `r format_years(outliers$Year)`.
These are unsurprising given @fig-fr-mortality, and most coincide with years already flagged in @sec-weird-times using the Hampel identifier applied directly to the raw age-specific series --- a quite different method, applied to the raw data rather than to PCA scores.
The following historical events are relevant:

- 1871: Repression of the 'Commune de Paris'
- 1914--1918: World War I
- 1919: Spanish flu pandemic
- 1940--1944: World War II

To take another perspective, we can apply the O3 plot (@sec-o3) to the first three principal component scores.

```{r}
#| label: pca_o3prep
#| dependson: fig-pca-outliers
#| code-fold: false

o3 <- pca_scores |>
  select(-Year) |>
  OutliersO3::O3prep(
    method = c("PCS", "adjOut", "MCD", "BAC", "DDC"),
    tolPCS = 0.05,
    toladj = 0.15,
    tolMCD = 0.001,
    tolBAC = 0.001,
    tolDDC = 0.025
  )
o3_plots <- OutliersO3::O3plotM(
  o3,
  coltxtsize = 10,
  caseNames = paste0("Y", pca_scores$Year)
)
```

```{r}
#| label: fig-o3-pca
#| fig-width: 14
#| fig-asp: 0.2
#| dependson: pca_o3prep
#| code-fold: false

o3_plots$gO3
```

The years flagged at least 7 times across all five methods and variable combinations are given below.

```{r}
#| label: o3-outliers
#| dependson: pca_o3prep

red_case <- o3_plots$outsTable |>
  distinct(Case, Combination, Method) |>
  count(Case, Combination, name = "n_agree") |>
  filter(n_agree == 5L)
out_table <- o3_plots$outsTable |>
  count(Case, name = "n_flagged") |>
  filter(n_flagged > 6) |>
  mutate(red = Case %in% red_case$Case)
pca_scores |>
  slice(out_table$Case) |>
  mutate(n_flagged = out_table$n_flagged, red = out_table$red) |>
  arrange(Year)
```

```{r}
#| label: o3-summary
#| echo: false
#| dependson: o3-outliers

o3_years <- sort(pca_scores$Year[out_table$Case])
o3_red_years <- sort(pca_scores$Year[out_table$Case[out_table$red]])
n_pc_o3 <- ncol(pca_scores) - 1
n_pc_surprisal <- pca_scores |> select(-Year, -PC3) |> ncol()
n_methods <- length(unique(o3_plots$outsTable$Method))
same_years <- setequal(outliers$Year, o3_years)
```

Reassuringly, the O3 plot, using all `r english::words(n_pc_o3)` principal components and `r english::words(n_methods)` quite different outlier-detection methods, flags `r english::words(length(o3_years))` years (`r format_years(o3_years)`) --- `r if (same_years) "exactly the same years" else "largely the same years"` identified by the two-component surprisal approach above.
Of these, `r english::words(length(o3_red_years))` --- `r format_years(o3_red_years)` --- are flagged by every method and variable combination (shown in red in @fig-o3-pca), while the rest are flagged consistently but not unanimously.
This agreement across methods, and with the time series analysis of @sec-weird-times, gives us some confidence that these years are genuine anomalies rather than artefacts of any one method.

## Functional depth measures {#sec-functional-depth}

There are several other approaches to identifying functional outliers, summarised in @hubert2015multivariate, almost all based on functional depth.
