# Functional data {#sec-functional}

```{r}
#| label: setup12
#| include: false
#| cache: false

source("before-each-chapter.R")
```

Functional data occur when the natural observational representation is a function.
The most commonly occurring examples are where observations are curves (functions over one dimension) or surfaces (functions over two dimensions).
For example, mortality rates are naturally represented as 1-dimensional functions of age, and surface temperatures are naturally represented as functions over the 2-dimensional plane of latitude and longitude.

@fig-fr-mortality shows French male mortality rates between 1816 and 1999.
In @sec-weird-times, we looked at these as time series, treating each age as a separate series.
Here, we will treat the same data as curves, with each year represented as a function of mortality rates over age.
Each line in @fig-fr-mortality denotes the mortality rates for one year as a function of age, with the colours in rainbow order corresponding to the years of observation.
Overall, we see a large decrease in mortality rates during early childhood years, then an increase during teenage years.
After about age 30, the rates increase almost linearly on a log scale.
Comparing the curves over time, we see that the rates have steadily fallen for all ages, with more than a 40-fold reduction in mortality rates at around age 10.
The noticeable gap between the curves occurred after World War II due to the effective use of antibiotics, and widespread vaccinations for polio, diphtheria, tetanus, whooping cough, measles, mumps and rubella.

```{r}
#| label: fig-fr-mortality
#| code-fold: false
#| fig-cap: French male mortality rates between 1816 and 1999.

fr_male <- fr_mortality |> filter(Sex == "Male") |> select(-Sex)
fr_male |>
  ggplot() +
  aes(x = Age, y = Mortality, colour = Year, group = Year) +
  geom_line() +
  scale_colour_gradientn(colours = rainbow(100)[1:80]) +
  scale_y_log10(
    breaks = 10^(-(0:4)),
    labels = format(10^(-(0:4)), scientific = FALSE)
  ) +
  labs(y = "Mortality rate")
```

As we are interested in anomalies, we will mostly focus on the unusual curves.
There are about 10--15 years when the mortality rate for 20--40 year olds is much higher than in other years.
These are the war years.
Any anomaly detection method for functional data should identify these years as unusual.

We will denote the functional data as $y_i(x)$.
Here, $i$ is a discrete index over all observations, and $x$ is the continuous variable over which each function is defined.
For the mortality data, $y_{i}(x)$ represent the mortality rate for age $x$ in year $i$.

It is possible to have multiple indexes (e.g., mortality rates for males and females, or for different countries, or both), and the function may be defined over more than one variable (e.g., mortality rates over age and BMI).
It is also possible to have multivariate observations (e.g., mortality and fertility rates).

## Interpolation and smoothing {#sec-interpolation}

The distinguishing feature of functional data is that we think of the data as smooth and continuous functions.
In practice, our observations are usually at discrete values, even though the underlying functions are continuous.
In @fig-fr-mortality, for example, we observe the mortality rate for discrete ages 0, 1, 2, \dots, while we think of the rate as defined for any age $x \in [0,\infty)$.

It is common to form continuous functions using some kind of interpolation or smoothing of the raw data.
Smoothing is used if the underlying curve has been observed with measurement error.

The data shown in @fig-fr-mortality are observed with error, as the "wiggliness" of the curves shows.
The mortality rate (or central death rate) at age $x$ and year $i$ is defined as the number of people who died at age $x$ in year $i$ divided by the number of people who were aged $x$ in the middle of year $i$.
We can smooth the data to obtain the mortality rates at noninteger values of $x$.

In the following code, we use a smoothing method proposed by @HU07, which takes account of the characteristics of mortality rates.

```{r}
#| label: fr_smooth
#| dependson: fig-fr-mortality
#| message: false
#| code-fold: false

library(vital)
fr_smooth <- fr_male |>
  as_vital(index = Year, key = Age, .age = "Age") |>
  smooth_mortality(Mortality, k = 21) |>
  as_tibble()
```

@fig-fr-smooth shows the resulting smooth curve for one year along with the original data, while @fig-fr-smooth2 shows the smooth curves for all years.

```{r}
#| label: fig-fr-smooth
#| dependson: [fig-fr-mortality, fr_smooth]
#| fig-cap: Smoothed French male mortality rates for 1999.

fr_male |>
  filter(Year == 1999) |>
  ggplot(aes(x = Age, y = Mortality)) +
  geom_point() +
  geom_line(
    data = fr_smooth |> filter(Year == 1999),
    aes(y = .smooth)
  ) +
  scale_y_log10(
    breaks = 10^(-(0:4)),
    labels = format(10^(-(0:4)), scientific = FALSE)
  ) +
  labs(y = "Mortality rate")
```

```{r}
#| label: fig-fr-smooth2
#| dependson: fr_smooth
#| fig-cap: Smoothed French male mortality rates for all years.

fr_smooth |>
  ggplot() +
  aes(x = Age, y = .smooth, colour = Year, group = Year) +
  geom_line() +
  scale_colour_gradientn(colours = rainbow(100)[1:80]) +
  scale_y_log10(
    breaks = 10^(-(0:4)),
    labels = format(10^(-(0:4)), scientific = FALSE)
  ) +
  labs(y = "Mortality rate")
```

Working with the smoothed curves rather than the raw mortality rates has several advantages for anomaly detection.
Smoothing removes measurement noise that would otherwise be mistaken for genuine functional variation, so isolated noisy observations are not flagged as anomalies, and real anomalies are not masked by noise.
It also gives more stable and reliable estimated rates than the raw data alone, by borrowing strength across neighbouring ages, using the fact that mortality varies smoothly with age.

## Dimension reduction using basis functions {#sec-basis-functions}

We can approximate functional data using a linear combination of basis functions:
$$
  \hat{y}_i(x) = b_0(x) + \sum_{j=1}^J \beta_{ij} b_j(x),
$$
where $b_0(x)$ is a fixed function common to every observation (analogous to an intercept), $\{b_1(x),\dots,b_J(x)\}$ are a set of basis functions, and $\beta_{ij}$ are the coefficients.
With enough basis functions, this representation can approximate a wide class of functions.
For example, the Fourier basis functions are a set of sine and cosine functions that can be used to represent any smooth periodic function.
The B-spline basis functions are a set of piecewise polynomial functions that can be used to represent any smooth function.
The choice of basis functions depends on the characteristics of the data.

The number of basis functions $J$ is usually much smaller than the number of observations, with larger $J$ corresponding to a more flexible approximation.

Given a set of basis functions, we can estimate the coefficients $\beta_{ij}$ by minimising the sum of squared errors between the observed data $y_i(x)$ and the approximated data $\hat{y}_i(x)$.

Alternatively, we can use functional principal component analysis to find the empirical basis functions that explain the most variance in the data.
Then $b_0(x)$ is set to be the mean function,
$$
  b_0(x) = \frac{1}{n} \sum_{i=1}^n y_i(x),
$$
and the first principal component, $b_1(x)$, is the function that minimises
$$
  \sum_{i=1}^n \int \big[y_i(x) - b_0(x) - \beta_{i1}b_1(x)\big]^2 dx
$$
subject to the constraint that $\int b_1^2(x)dx = 1$.
The second principal component, $b_2(x)$, is the function that minimises
$$
  \sum_{i=1}^n \int \big[y_i(x) - b_0(x) - \beta_{i1}b_1(x) - \beta_{i2}b_2(x)\big]^2 dx
$$
subject to the constraints that $\int b_2^2(x)dx = 1$ and $\int b_1(x)b_2(x)dx = 0$.
And so on.
The principal components $b_1(x), b_2(x), \dots$, are also called the "loadings", while the coefficients $\beta_{i,1},\beta_{i,2},\dots$ are called the "scores".

If the functions are discretised on a fine grid over $x$, we can use regular principal component decomposition to get approximately the same results as functional principal components.

In the following example, we find the first few principal components using the French male mortality data.
(The smoothed data are already on a fine grid of ages, so we don't need to discretise them further.)

```{r}
#| echo: true
#| code-fold: false
#| message: false
#| label: pca-bases
#| dependson: fr_smooth

# Wide version of smoothed log Mortality with ages on columns
frmort_wide <- fr_smooth |>
  mutate(logmx = log(.smooth)) |>
  select(Year, Age, logmx) |>
  tidyr::pivot_wider(
    names_from = Age,
    values_from = logmx,
    names_prefix = "Age"
  )
# Matrix containing only the log Mortality values, not the year
frmort_mat <- frmort_wide |>
  select(-Year) |>
  as.matrix()
# Compute first three principal components
pca <- frmort_mat |>
  prcomp(center = TRUE, scale = FALSE, rank = 3)
summary(pca)
```

The first three principal components explain `r round(100 * sum(pca$sdev[1:3]^2) / sum(pca$sdev^2), 1)`% of the variation in the data.
Instead of working in the `r NCOL(frmort_mat)` dimensional space containing all ages, we can work in the 3-dimensional space of the first three principal component scores, which is a much simpler problem, without losing much information.

```{r}
#| echo: true
#| code-fold: false
#| message: false
#| label: fig-pca-bases
#| dependson: pca-bases
#| fig-cap: First three principal components (empirical basis functions) for the French male mortality data.

# Extract PCA basis functions
pca_bases <- broom::tidy(pca, "loadings") |>
  mutate(
    Age = readr::parse_number(column),
    Basis = paste0("PC", as.factor(PC))
  )
pca_bases |>
  ggplot() +
  aes(x = Age, y = value, group = Basis, colour = Basis) +
  geom_line() +
  labs(y = latex2exp::TeX("$b_i(x)$"), x = "Age (x)")
```

@fig-pca-bases shows the first three empirical basis functions.

1. The first basis function is negative at every age, and largest in magnitude in early childhood, so a high score on this component corresponds to lower mortality at all ages, but especially for children.
2. The second basis function contrasts men of fighting age (roughly ages 18--40) against children, so it picks out years where mortality among young men was unusually high (or low) relative to mortality in childhood.
3. The third basis function contrasts infant mortality and older mortality (above age 40), with mortality of pre-teens and early 20s.

```{r}
#| echo: true
#| code-fold: false
#| message: false
#| label: fig-pca-scores
#| dependson: pca-bases
#| fig-cap: First three principal component scores for the French male mortality data.

# Extract PCA scores
pca_scores <- broom::augment(pca, frmort_wide[, "Year"]) |>
  select(-.rownames) |>
  rename_with(~ gsub("\\.fitted", "", .x))
# Time series plot of scores
pca_scores |>
  tidyr::pivot_longer(
    -Year,
    names_to = "PC",
    values_to = "value"
  ) |>
  ggplot(aes(x = Year, y = value)) +
  geom_line(aes(colour = PC)) +
  labs(y = latex2exp::TeX("$\\beta_{i,t}$"), x = "Year (t)")
```

@fig-pca-scores shows how the two component scores evolve over time.
Other than during the two world wars, the first component score increases steadily and fairly smoothly across the study period, reflecting the long-run, near-continuous decline in mortality at all ages.
The second component score sits close to zero for most years but spikes sharply during 1914--1918 and 1940--1945, capturing the additional mortality among young French men during the two world wars.
The third component is more subtle, and reflects slow changes in mortality patterns over time, rather than the sharp spikes seen in the second component.

```{r}
#| echo: true
#| code-fold: false
#| message: false
#| label: fig-pca-scores2
#| dependson: fig-pca-scores
#| fig-cap: Scatterplot of first three principal component scores for the French male mortality data.

pca_scores |>
  select(-Year) |>
  GGally::ggpairs()
```

@fig-pca-scores2 plots the pairwise scatterplots of the first three component scores against each other.
Most years fall along a narrow band in each plot, but there are some clear anomalies, especially in the second component.

This low-dimensional view foreshadows the anomaly detection approach of the next section, where we treat points that lie far from the bulk of the data in this principal component space as candidate anomalies.

## Anomalies in the principal component space {#sec-functional-pca}

Once the functional data have been reduced to a handful of principal component scores, each curve (here, each year) is just a point in a low-dimensional numerical space, and we can apply the same methods introduced previously in this book.
For example, we can compute a surprisal probability for each year from the first two component scores, using a GPD approximation for the tail (@sec-approximate-surprisal-prob), and flag any year with a probability below 0.05 as anomalous.
We restrict attention to the first two components here so the results can be shown more easily in a scatterplot; we return to all three components below using the O3 plot.

```{r}
#| echo: true
#| code-fold: false
#| label: fig-pca-outliers
#| dependson: fig-pca-scores
#| fig-cap: French principal component scores, with anomalous years (surprisal probability below 0.05) shown in red.

# Find outliers in the PCs
outliers <- pca_scores |>
  mutate(prob = surprisals_prob(pick(PC1, PC2), approximation = "gpd")) |>
  filter(prob < 0.05)

pca_scores |>
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point() +
  geom_point(data = outliers, colour = "red") +
  ggrepel::geom_label_repel(data = outliers, aes(label = Year))
```

```{r}
#| label: format-years
#| echo: false

# Format a vector of years as individual years and ranges (3+ consecutive years)
format_years <- function(years) {
  years <- sort(unique(years))
  grp <- cumsum(c(1L, diff(years) != 1L))
  parts <- unlist(lapply(
    split(years, grp),
    \(g) {
      if (length(g) >= 3) paste0(g[1], "--", g[length(g)]) else as.character(g)
    }
  ))
  if (length(parts) > 1) {
    paste(
      paste(parts[-length(parts)], collapse = ", "),
      "and",
      parts[length(parts)]
    )
  } else {
    parts
  }
}
```

@fig-pca-outliers identifies `r english::words(nrow(outliers))` anomalous years: `r format_years(outliers$Year)`.
These are unsurprising given @fig-fr-mortality, and most coincide with years already flagged in @sec-weird-times using the Hampel identifier applied directly to the raw age-specific series --- a quite different method, applied to the raw data rather than to PCA scores.
The following historical events are relevant:

- 1871: Repression of the 'Commune de Paris'
- 1914--1918: World War I
- 1919: Spanish flu pandemic
- 1940--1944: World War II

To take another perspective, we can apply the O3 plot (@sec-o3) to the first three principal component scores.

```{r}
#| label: pca_o3prep
#| dependson: fig-pca-outliers
#| code-fold: false

o3 <- pca_scores |>
  select(-Year) |>
  OutliersO3::O3prep(
    method = c("PCS", "adjOut", "MCD", "BAC", "DDC"),
    tolPCS = 0.05,
    toladj = 0.15,
    tolMCD = 0.001,
    tolBAC = 0.001,
    tolDDC = 0.025
  )
o3_plots <- OutliersO3::O3plotM(
  o3,
  coltxtsize = 10,
  caseNames = paste0("Y", pca_scores$Year)
)
```

```{r}
#| label: fig-o3-pca
#| fig-width: 14
#| fig-asp: 0.2
#| dependson: pca_o3prep
#| code-fold: false

o3_plots$gO3
```

The years flagged at least 7 times across all `r english::words(length(unique(o3_plots$outsTable$Method)))` methods and variable combinations are given below.

```{r}
#| label: o3-outliers
#| dependson: pca_o3prep

red_case <- o3_plots$outsTable |>
  distinct(Case, Combination, Method) |>
  count(Case, Combination, name = "n_agree") |>
  filter(n_agree == 5L)
out_table <- o3_plots$outsTable |>
  count(Case, name = "n_flagged") |>
  filter(n_flagged > 6) |>
  mutate(red = Case %in% red_case$Case)
pca_scores |>
  slice(out_table$Case) |>
  mutate(n_flagged = out_table$n_flagged, red = out_table$red) |>
  arrange(Year)
```

```{r}
#| label: o3-summary
#| echo: false
#| dependson: o3-outliers

o3_years <- sort(pca_scores$Year[out_table$Case])
o3_red_years <- sort(pca_scores$Year[out_table$Case[out_table$red]])
n_pc_o3 <- ncol(pca_scores) - 1
n_pc_surprisal <- pca_scores |> select(-Year, -PC3) |> ncol()
n_methods <- length(unique(o3_plots$outsTable$Method))
same_years <- setequal(outliers$Year, o3_years)
```

Reassuringly, the O3 plot, using all `r english::words(n_pc_o3)` principal components and `r english::words(n_methods)` quite different outlier-detection methods, flags `r english::words(length(o3_years))` years (`r format_years(o3_years)`) --- `r if (same_years) "exactly the same years" else "largely the same years"` identified by the two-component surprisal approach above.
Of these, `r english::words(length(o3_red_years))` --- `r format_years(o3_red_years)` --- are flagged by every method and variable combination (shown in red in @fig-o3-pca), while the rest are flagged consistently but not unanimously.
This agreement across methods, along with the time series analysis of @sec-weird-times, gives us some confidence that these years are genuine anomalies rather than artefacts of any one method.

## Functional bagplot {#sec-functional-bagplot}

A boxplot summarises a univariate sample using a central box and whiskers, with points beyond the whiskers flagged as outliers.
Several different ways of extending this idea to functional data are described over the sections that follow.
The functional bagplot and the functional HDR boxplot, covered in this section and the next, were proposed by @HS08 [@rainbow].
They follow a similar approach: reduce each curve to its first two functional principal component scores (@sec-basis-functions), draw the bivariate bagplot or HDR boxplot of those scores, and then map each region back to a functional envelope formed from the curves whose scores fall inside it.
So the inner region of the functional bagplot is bounded by the curves whose scores lie in the bivariate bag, and the outer region is bounded by the curves whose scores lie in the bivariate loop.
The bagplot's central point (the depth median of the scores) need not coincide with any observed point, so the "median curve" is instead taken to be the observed curve whose scores are closest (in Euclidean distance) to that central point.

@fig-pca-bagplot shows the ordinary bivariate bagplot (@sec-bagplots) of the first two principal component scores.
Each point corresponds to one year's mortality curve; the dark region is the bag, the light region is the loop, and any points beyond the loop are flagged as bivariate outliers, coloured by year to match the corresponding curves in @fig-fbagplot below.

```{r}
#| label: fig-pca-bagplot
#| dependson: fig-pca-scores
#| code-fold: true
#| fig-cap: Bagplot of the first two principal component scores of the French male mortality curves.

bag <- aplpack::compute.bagplot(as.matrix(pca_scores[, c("PC1", "PC2")]))
bagplot_outliers <- as.data.frame(bag$pxy.outlier) |>
  left_join(pca_scores, by = c("PC1", "PC2"))

pca_scores |>
  gg_bagplot(PC1, PC2) +
  geom_point(data = bagplot_outliers, aes(colour = factor(Year)), size = 2) +
  labs(colour = "Outlier year")
```

The bag and loop shown here trace out regions of "typical" curves, while the flagged points correspond to years whose curves depart from that typical shape.
Taking the pointwise range of the curves whose scores fall inside the bag, and separately of the curves whose scores fall inside the loop, gives an envelope for the "typical" region of the functional bagplot at every age.

The following code identifies the curves contained in the bag and the loop, and those beyond that loop corresponding to the anomalies in @fig-pca-bagplot.

```{r}
#| label: fig-fbagplot
#| dependson: fig-pca-bagplot
#| code-fold: true
#| fig-cap: Functional bagplot of French male mortality.
#|   The dark band shows the envelope of the bag (50% region), the light band shows the envelope of the loop; the solid curve is the curve corresponding to the bagplot's central point; the raw curves for years flagged as bagplot outliers are overlaid.

# Pointwise envelope of the actual (smoothed) curves for a set of years
curve_envelope <- function(years) {
  fr_smooth |>
    filter(Year %in% years) |>
    summarise(ymin = min(.smooth), ymax = max(.smooth), .by = Age)
}
bag_years <- as.data.frame(bag$pxy.bag) |>
  left_join(pca_scores, by = c("PC1", "PC2")) |>
  pull(Year)
loop_years <- as.data.frame(bag$pxy.outer) |>
  left_join(pca_scores, by = c("PC1", "PC2")) |>
  pull(Year)
bag_env <- curve_envelope(bag_years)
loop_env <- curve_envelope(c(bag_years, loop_years))
# Central curve: the observed curve closest to the bagplot's depth median
centre_point <- which.min(
  (pca_scores$PC1 - bag$center[1])^2 + (pca_scores$PC2 - bag$center[2])^2
)
centre_curve <- fr_smooth |> filter(Year == pca_scores$Year[centre_point])
outlier_curves <- fr_male |>
  filter(Year %in% bagplot_outliers$Year)
bag_colors <- c("#00659e", "#5397BD", "#A7CADD")
ggplot() +
  geom_ribbon(
    data = loop_env,
    aes(x = Age, ymin = ymin, ymax = ymax),
    fill = bag_colors[3]
  ) +
  geom_ribbon(
    data = bag_env,
    aes(x = Age, ymin = ymin, ymax = ymax),
    fill = bag_colors[2]
  ) +
  geom_line(
    data = centre_curve,
    aes(x = Age, y = .smooth),
    colour = bag_colors[1]
  ) +
  geom_line(
    data = outlier_curves,
    aes(x = Age, y = Mortality, group = Year, colour = factor(Year))
  ) +
  scale_y_log10(
    breaks = 10^(-(0:4)),
    labels = format(10^(-(0:4)), scientific = FALSE)
  ) +
  labs(y = "Mortality rate", colour = "Outlier year")
```

@fig-fbagplot flags `r english::words(nrow(bagplot_outliers))` years as functional bagplot outliers: `r format_years(bagplot_outliers$Year)`.
These are almost exactly the war years already identified in @sec-functional-pca and @sec-weird-times, and the outlying curves are clearly visible bulging above the typical envelope for young men.

Another implementation of the functional bagplot is provided by `rainbow::fboxplot()`.

## Functional HDR boxplot {#sec-functional-hdr-boxplot}

The functional HDR boxplot follows the same recipe, but uses the highest density regions of a bivariate kernel density estimate of the scores (@sec-mhdr) in place of the bagplot's bag and loop.

@fig-pca-hdrboxplot shows the ordinary bivariate HDR boxplot (@sec-hdr-boxplots) of the first two principal component scores.
Because of the strange shape that the principal component scores lie on, the resulting kernel density estimate is particularly sensitive to the bandwidth choice, and the resulting HDRs are broken into several disconnected pieces.
After some experimentation, we chose the "scv" (smoothed cross-validation) bandwidth method.
To roughly match the number of years flagged by the bagplot, we use a 93% coverage probability for the larger HDR.
Thus, the dark region is the 50% HDR (containing 50% of the years), the light region is the 93% HDR (containing 93% of the years), and the points outside the 93% HDR are flagged as outliers, coloured by year to match the corresponding curves in @fig-fhdrboxplot below.

```{r}
#| label: fig-pca-hdrboxplot
#| dependson: fig-pca-scores
#| fig-cap: HDR boxplot of the first two principal component scores of the French male mortality curves.

hdr_labels <- pca_scores[, c("PC1", "PC2")] |>
  dist_kde(method = "scv") |>
  hdr_regions(prob = c(0.5, 0.93))
pca_scores <- pca_scores |>
  mutate(
    hdr_region = case_when(
      !is.na(hdr_labels$hdr_50) ~ "50%",
      !is.na(hdr_labels$hdr_93) ~ "93%",
      TRUE ~ "outlier"
    ),
    hdr_cluster = hdr_labels$hdr_93
  )
hdr_outliers <- pca_scores |> filter(hdr_region == "outlier")
pca_scores |>
  gg_hdrboxplot(PC1, PC2, prob = c(0.5, 0.93), method = "scv") +
  geom_point(
    data = hdr_outliers,
    aes(x = PC1, y = PC2, colour = factor(Year)),
    size = 2
  ) +
  labs(colour = "Outlier year") +
  guides(colour = "legend")
```

Here, we have used `hdr_regions()` to classify each year's scores as either inside the 50% HDR, inside the 93% HDR but outside the 50% HDR, or outside the 93% HDR (and thus flagged as an outlier).

All but one of the `r english::words(nrow(hdr_outliers))` years flagged as outliers have previously been identified as anomalous above (in @fig-pca-outliers, @fig-o3-pca or @fig-pca-bagplot); the exception, 1855, was not picked out by any of the earlier methods, but corresponds to a year during the Crimean War [@french_mortality].

```{r}
#| label: hdr-consensus-check
#| dependson: [fig-pca-hdrboxplot, fig-pca-bagplot, o3-summary]
#| include: false

stopifnot(
  "the 'previously identified' claim above no longer matches hdr_outliers -- update the text" = setequal(
    setdiff(hdr_outliers$Year, union(o3_years, bagplot_outliers$Year)),
    1855
  )
)
```

As with the bagplot, the envelope of each functional HDR is formed directly from the curves whose scores fall inside it, using the `hdr_region` classification already computed above.
Because the bivariate scores split into two separated clusters (@fig-pca-hdrboxplot), so does each HDR: `hdr_regions()` labels which disconnected piece of the HDR each year's scores fall into, via `hdr_cluster`, and we build a separate envelope for each piece, so the split remains visible wherever the two pieces' envelopes do not overlap.

```{r}
#| label: fig-fhdrboxplot
#| dependson: [fig-fbagplot, fig-pca-hdrboxplot, o3-summary]
#| fig-cap: Functional HDR boxplot of French male mortality.
#|   The dark bands show the envelope of the 50% HDR, the light bands the 93% HDR; the raw curves for years flagged as outliers (falling outside the 93% HDR) are overlaid.

hdr_scores <- pca_scores |> filter(hdr_region != "outlier")
cluster_envelope <- function(years) {
  fr_smooth |>
    filter(Year %in% years) |>
    left_join(hdr_scores[, c("Year", "hdr_cluster")], by = "Year") |>
    summarise(
      ymin = min(.smooth),
      ymax = max(.smooth),
      .by = c(Age, hdr_cluster)
    )
}
hdr50_env <- cluster_envelope(hdr_scores$Year[hdr_scores$hdr_region == "50%"])
hdr93_env <- cluster_envelope(hdr_scores$Year)
outlier_curves <- fr_male |>
  filter(Year %in% hdr_outliers$Year)
ggplot() +
  geom_ribbon(
    data = hdr93_env,
    aes(x = Age, ymin = ymin, ymax = ymax, group = hdr_cluster),
    fill = bag_colors[3]
  ) +
  geom_ribbon(
    data = hdr50_env,
    aes(x = Age, ymin = ymin, ymax = ymax, group = hdr_cluster),
    fill = bag_colors[2]
  ) +
  geom_line(
    data = outlier_curves,
    aes(x = Age, y = Mortality, group = Year, colour = factor(Year))
  ) +
  scale_y_log10(
    breaks = 10^(-(0:4)),
    labels = format(10^(-(0:4)), scientific = FALSE)
  ) +
  labs(y = "Mortality rate", colour = "Outlier year")
```

The bulk of the curves fall into two distinct clusters, corresponding to before and after the end of World War II.

Another implementation of the functional HDR boxplot is provided by `rainbow::fboxplot()`.

## Functional depth {#sec-functional-depth}

Rather than reducing each curve to a handful of principal component scores, the remaining approaches in this chapter work directly with the curves themselves, via functional depth or a related measure.
See @hubert2015multivariate for a more detailed review of these ideas.

The idea of depth generalises the univariate idea of a median and quantiles to more complex objects.
We have already seen this with multivariate numeric data (@sec-depth); now we will extend the idea to functional data.
Given a sample of curves $y_1(x),\dots,y_n(x)$, a depth measure assigns each curve $y_i$ a value between 0 and 1 describing how central it is within the sample: curves near the middle of the distribution receive high depth, while unusual curves receive low depth.
Curves can then be ranked from the deepest (most typical) to the shallowest (most unusual), giving an ordering analogous to ranking univariate observations from the median outward.

One of the simplest depths is the band depth: it counts how often a curve $y_i$ lies completely within the "band" traced out by every possible group of $k$ other curves in the sample, so that curves often enveloped by others are deep, while curves that stick out --- even briefly --- have low depth.
Looking at @fig-fr-mortality, we would expect that some of the most recent years, which have low mortality at almost every age, would be shallow; the war years would also be shallow, since they have high mortality at ages 20--40 that is not seen in other years.

Because the exact band depth is expensive to compute for anything beyond a handful of curves, in practice it is usually replaced by the more tractable modified band depth (MBD), which instead measures the proportion of the domain $x$ over which $y_i(x)$ lies within each band, averaged over all bands [@LopezPintado2009].

```{r}
#| label: mbd-build
#| dependson: [fr_smooth, pca-bases]
#| code-fold: false
#| message: false

mbd <- function(Y) {
  n <- NROW(Y)
  m <- NCOL(Y)
  r <- apply(Y, 2, rank)
  # (r-1)(n-r): pairs of curves whose band contains y at this point.
  # + (n-1): pairs where y is itself one of the two band-forming curves.
  (rowSums((r - 1) * (n - r)) / m + (n - 1)) / choose(n, 2)
}
depth_tbl <- tibble(Year = frmort_wide$Year, MBD = mbd(frmort_mat))
depth_tbl |> arrange(MBD)
```

```{r}
#| label: shallowyears
#| dependson: mbd-build
#| echo: false

shallow_years <- depth_tbl |> arrange(MBD) |> head(10) |> pull(Year)
shallow_1800s <- shallow_years[shallow_years < 1900] |> sort()
stopifnot(identical(shallow_1800s, c(1854L, 1870L, 1871L)))
```

Ranking curves by MBD alone identifies the years whose entire trajectory lies furthest from the bulk of the sample: the shallowest curves are `r format_years(shallow_years)`.
The `r english::words(length(shallow_years) - length(shallow_1800s))` most recent of these have mortality lower at almost every age than in any other year, while the `r english::words(length(shallow_1800s))` from the 1800s are shallow for the opposite reason --- their mortality is distinctly higher than in surrounding years at almost every age.
These earlier shallow years correspond to the 1854 cholera outbreak, the 1870 Franco-Prussian war and the 1871 Repression of the 'Commune de Paris'.

This illustrates an important feature of depth-based ranking: it can fail to distinguish between curves that are unusual due to location and those that are unusual due to shape.
Because French mortality has fallen more or less steadily across the study period, the most recent curves sit at one edge of the whole 184-year sample and are therefore the least "enveloped" by the rest, even though their shape is not unusual.
On the other hand, the 1854, 1870 and 1871 curves are unusual in shape, but they are not extreme in location relative to the rest of the 19th century curves.

Other types of functional depth are computed by `rainbow::fdepth()`, and the `fda.usc`, `fdaoutliers` and `mrfDepth` packages.

## Sun--Genton functional boxplot {#sec-sungenton-boxplot}

The Sun--Genton functional boxplot [@Sun2011] works directly with the (possibly smoothed) curves, using the modified band depth introduced in @sec-functional-depth.
The central region is formed from the 50% deepest curves (those with MBD at or above the median), and a curve is flagged as an outlier if, at any age, it leaves a fence formed by inflating this central region by a multiple of 1.5 of its own range.
@fig-fbplot-sungenton shows the resulting functional boxplot for the French male mortality curves.

```{r}
#| label: fig-fbplot-sungenton
#| dependson: [mbd-build, fig-fbagplot]
#| fig-cap: Sun--Genton functional boxplot of the smoothed French male mortality curves.
#|   The dark band shows the central 50% region, the light band the envelope of the curves not flagged as outliers; the solid curve is the deepest (most central) curve; the raw curves for the flagged outlier years are overlaid.

central_years <- depth_tbl$Year[depth_tbl$MBD >= median(depth_tbl$MBD)]
centre_year <- depth_tbl$Year[which.max(depth_tbl$MBD)]
# The fence is a linear inflation of the central region's width, so it must be
# computed on the log scale in this example
central_env <- fr_smooth |>
  filter(Year %in% central_years) |>
  summarise(cmin = min(log(.smooth)), cmax = max(log(.smooth)), .by = Age) |>
  mutate(
    flower = cmin - 1.5 * (cmax - cmin),
    fupper = cmax + 1.5 * (cmax - cmin)
  )
outlier_years <- fr_smooth |>
  mutate(logm = log(.smooth)) |>
  left_join(central_env, by = "Age") |>
  filter(logm < flower | logm > fupper) |>
  distinct(Year) |>
  pull(Year)
fr_smooth |>
  summarise(
    cmin = min(.smooth[Year %in% central_years]),
    cmax = max(.smooth[Year %in% central_years]),
    fmin = min(.smooth[!Year %in% outlier_years]),
    fmax = max(.smooth[!Year %in% outlier_years]),
    .by = Age
  ) |>
  ggplot() +
  geom_ribbon(
    aes(x = Age, ymin = fmin, ymax = fmax),
    fill = bag_colors[3]
  ) +
  geom_ribbon(
    aes(x = Age, ymin = cmin, ymax = cmax),
    fill = bag_colors[2]
  ) +
  geom_line(
    data = fr_smooth |> filter(Year == centre_year),
    aes(x = Age, y = .smooth),
    colour = bag_colors[1]
  ) +
  geom_line(
    data = fr_smooth |> filter(Year %in% outlier_years),
    aes(x = Age, y = .smooth, group = Year, colour = factor(Year))
  ) +
  scale_y_log10(
    breaks = 10^(-(0:4)),
    labels = format(10^(-(0:4)), scientific = FALSE)
  ) +
  labs(y = "Mortality rate", colour = "Outlier year")
```

@fig-fbplot-sungenton fails to identify the war years or pandemics as outliers, and inflates the size of the central region to include almost all of the curves.
In fact, it flags `r english::words(length(outlier_years))` years, `r format_years(outlier_years)`, none of which are war years.
These are simply the most recent years in the sample, whose mortality is lower at almost every age than any earlier year.

This may seem inconsistent with 1854, 1870 and 1871 being the shallowest 19th century curves, but a low MBD only means a curve is poorly enveloped by bands of other curves; it does not mean the curve's values are pointwise extreme, which is what the boxplot fence tests.
The mortality in those three years, though unusual in shape, never falls outside the range spanned by other 19th century curves, so they never cross the fence, unlike the six most recent years.

The functional bagplot and HDR boxplot do not share the problem of trend, because reducing each curve to two principal component scores loses most of the information about the gradual, whole-of-curve trend captured by MBD, while retaining the specific age-pattern (young men versus children) that made the war years unusual.

While the Sun--Genton functional boxplot did not do well in this example, it can still be a useful tool for identifying unusual curves in other contexts, especially where the first two principal component scores do not capture the most interesting variation in the data.

We coded the functional boxplot from scratch here, but versions are available in other packages including `fda::fbplot()`, `GET::fBoxplot()`, `fdaoutlier::functional_boxplot()` and `roahd::fbplot()`.

The next two sections look at ways of separating genuine shape anomalies from the kind of trend-driven magnitude effect that misled the depth boxplot here.

## Outliergram {#sec-outliergram}

The modified band depth used above conflates two different reasons a curve can be unusual: it might have a genuinely atypical *shape*, or it might sit at one extreme of the bulk of the data, and so be poorly enveloped by the rest of the sample.
The outliergram of @Arribas-Gil2014-ll separates these two effects by comparing MBD against a second measure, the modified epigraph index (MEI) of @LopezPintadoRomo2011.

Whereas MBD measures how often a curve is enveloped by bands formed from pairs of other curves, MEI measures a curve's rank position: it is close to 1 for a curve lying below almost every other curve across almost the whole domain, close to 0 for a curve lying above almost every other curve, and around 0.5 for a curve in the middle of the pack.
So MEI captures a curve's overall level relative to the rest of the sample, in much the same way that the first principal component score did in @sec-functional-pca, while MBD captures how well-enveloped it is.

For most collections of curves, curves with an extreme MEI (very close to 0 or 1) are, by virtue of sitting near the edge of the sample, poorly enveloped and so also have somewhat low MBD.
In fact, there is a well-defined upper bound on MBD as a function of MEI, tracing out a parabola.
A curve with an unusual shape stands out from this pattern: its MBD falls noticeably *below* the parabola for its MEI value, meaning it is even less enveloped than curves with a similar rank would typically be.
The outliergram plots MEI against MBD along with this parabola, and flags as shape outliers those curves whose vertical distance below the parabola exceeds a boxplot-style fence (the third quartile of these distances, plus 1.5 times their interquartile range, by default).

```{r}
#| label: fig-outliergram
#| dependson: mbd-build
#| code-fold: true
#| fig-cap: Outliergram of the smoothed French male mortality curves.
#|   The shaded region shows the theoretical MBD/MEI parabola down to the outlier fence; years beyond the fence are shown in red.

mei <- function(Y) {
  n <- NROW(Y)
  m <- NCOL(Y)
  r <- apply(Y, 2, rank, ties.method = "min")
  rowSums(n - r + 1) / (n * m)
}
mei_tbl <- depth_tbl |> mutate(MEI = mei(frmort_mat))

# Theoretical upper bound of MBD as a function of MEI (Arribas-Gil & Romo 2014)
n_years <- NROW(frmort_wide)
a0 <- -2 / (n_years * (n_years - 1))
a1 <- 2 * (n_years + 1) / (n_years - 1)
a2 <- n_years^2 * a0
mbd_bound <- function(mei) a0 + a1 * mei + a2 * mei^2
mei_tbl <- mei_tbl |>
  mutate(d = mbd_bound(MEI) - MBD)

# Shape outliers: distance below the parabola exceeds a boxplot-style fence
fence <- quantile(mei_tbl$d, 0.75) + 1.5 * IQR(mei_tbl$d)
og_outliers <- mei_tbl |> filter(d >= fence)
parabola <- tibble(MEI = seq(0, 1, length.out = 100)) |>
  mutate(MBD = mbd_bound(MEI), lower = MBD - fence)

mei_tbl |>
  ggplot(aes(x = MEI, y = MBD)) +
  geom_ribbon(
    data = parabola,
    aes(ymin = lower, ymax = MBD),
    fill = "#00659e",
    alpha = 0.2
  ) +
  geom_line(data = parabola, aes(y = MBD), colour = "#00659e") +
  geom_point() +
  geom_point(data = og_outliers, colour = "#D55E00") +
  ggrepel::geom_label_repel(
    data = og_outliers,
    aes(label = Year),
    colour = "#D55E00",
    nudge_y = -0.05,
    direction = "y"
  )
```

@fig-outliergram flags `r english::words(nrow(og_outliers))` years as shape outliers: `r format_years(sort(og_outliers$Year))`.
These were in the middle of World War II, and are the years where the age profile changes shape most, with mortality rising sharply for young men while barely moving for children or the elderly.
More shape anomalies would be identified if we reduced the fence multiplier from 1.5.
Unlike the depth boxplot fence of @fig-fbplot-sungenton, none of the low-mortality recent years are flagged, because their low MBD is entirely explained by their extreme MEI --- they sit at one end of the rank ordering, but they do not deviate from the shape that curves at that end of the ranking are expected to have.

## Magnitude-shape plot {#sec-ms-plot}

A second way to separate magnitude and shape effects is the magnitude-shape plot of @Dai2018.
For each curve, and at each point $x$ on the grid, they compute a *directional outlyingness*: a signed measure of how far, and in which direction, the curve departs from the centre of the distribution of values at that $x$.
Averaging this quantity over the domain gives a magnitude score (MO) for each curve --- large when a curve sits consistently on one side of the bulk of the data.
The variation of the directional outlyingness over the domain gives a separate shape score (VO) --- large when a curve's departure from the centre changes direction or size as $x$ varies, which happens when a curve crosses others or has an unusual local pattern, even if its overall MO is modest.
Plotting VO against MO for every curve gives the magnitude-shape plot; @Dai2018 then flag outliers using a robust distance in this two-dimensional (MO, VO) space, with a finite-sample correction due to @Hardin2005.

These calculations are implemented in the `fdaoutlier` package, which we use here.

```{r}
#| label: fig-ms-plot
#| dependson: mbd-build
#| code-fold: true
#| fig-cap: Magnitude-shape plot of the smoothed French male mortality curves.
#|   Years flagged as outliers are shown in red.

ms <- fdaoutlier::msplot(frmort_mat, seed = 1, plot = FALSE)
ms_tbl <- tibble(
  Year = frmort_wide$Year,
  MO = ms$mean_outlyingness,
  VO = ms$var_outlyingness
)
ms_outliers <- ms_tbl |> slice(ms$outliers)

ms_tbl |>
  ggplot(aes(x = MO, y = VO)) +
  geom_point() +
  geom_point(data = ms_outliers, colour = "red")
```

```{r}
#| label: ms-plot-groups
#| dependson: fig-ms-plot
#| echo: false

ms_recent <- ms_outliers |> filter(Year >= 1946) |> pull(Year) |> sort()
ms_early <- ms_outliers |> filter(Year < 1946) |> pull(Year) |> sort()
```

@fig-ms-plot flags `r english::words(nrow(ms_outliers))` years as outliers, far more than either the outliergram or the depth boxplot.
Almost all of these, `r format_years(ms_recent)`, are years from `r min(ms_recent)` onwards --- every year after World War II.
The remainder, `r format_years(ms_early)`, are war and conflict years already familiar from earlier sections.
So the magnitude-shape plot re-discovers the same problem as the Sun--Genton functional boxplot of @sec-sungenton-boxplot: because mortality has fallen steadily across the study period, every sufficiently recent curve has a large, consistently-signed directional outlyingness relative to the whole sample, driving up its MO score regardless of whether anything unusual happened in that year.
Two conceptually different methods --- one based on band depth, one on directional outlyingness --- fail here for the same underlying reason: magnitude-based measures cannot cope with trending data, regardless of implementation.

## Summary {#sec-functional-summary}

We have applied six different methods to the same French male mortality data in this chapter, and they do not all agree, because they are not all measuring the same thing.
Reducing each curve to its first two or three principal component scores, and then applying surprisal probabilities, the O3 plot, the bagplot, or the HDR boxplot, all give a similar answer: the war years and a handful of other historical crises are flagged as anomalies, because these methods retain the specific age-pattern of a curve while discarding most of its whole-of-sample trend.
Working directly with the curves via band depth or directional outlyingness gives a different answer, because these magnitude-based measures cannot distinguish a curve that is genuinely unusual in shape from one that sits at the end of a long, gradual trend; on this data, that confounding is severe enough that both the Sun--Genton depth boxplot and the magnitude-shape plot flag most of the twentieth century as anomalous.
The outliergram sits in between, using the same band depth but comparing it against a curve's rank (via MEI) rather than its raw value, which lets it isolate genuine shape anomalies and, on this data, recovers a conservative version of the same war years found by the PCA-based methods.

No single method here is more "correct" than the others.
Trend is common in real functional data, and whether it should be treated as the expected background against which anomalies are judged, as the PCA- and shape-based methods effectively do here, or as something that can itself produce anomalies, is a modelling choice, not a purely statistical one.
In practice, applying more than one method, as we have done here, and using the disagreements between them to understand *why* particular curves are flagged is more informative than relying on any single number to decide what counts as anomalous.
