---
code-fold: false
---

# Matrix algebra {#sec-appendix}

```{r}
#| include: false
#| cache: false
source("before-each-chapter.R")
```


Working with multivariate data inevitably involves writing some of the ideas using matrix expressions and matrix algebra. In this appendix, a few of the results that we need are collected for convenient reference. Readers who have not previously studied matrix algebra are recommended to do some self-study using a good introductory textbook such as @Gentle07 or @Searle06.

## Identity matrix {#sec-identity-matrix}

The identity matrix is like the number 1: multiplying a matrix by the identity matrix leaves the matrix unchanged. The identity matrix is a square matrix with ones on the diagonal and zeros elsewhere, denoted by $\bm{I}$. Sometimes we will use subscripts to denote the size of the identity matrix; e.g. $\bm{I}_n$ is an $n \times n$ identity matrix.

```{r}
I3 <- diag(3)
I3
```

## Matrices and directions {#sec-matrices-directions}

A matrix can be applied to a vector $\bm{v}$ by matrix multiplication, producing a new vector $\bm{A}\bm{v}$. Geometrically, this transforms $\bm{v}$, generally changing both its length and the direction it points. Different starting vectors are typically transformed in different ways: a matrix may stretch some directions, shrink or rotate others, or even collapse two distinct directions onto the same output.

```{r}
#| label: matrixA
A <- matrix(c(3, 1, 1, 2), 2, 2)
A
v <- c(1, 0)
A %*% v
```

Geometrically, $\bm{v}$ originally pointed along the $x$-axis with length 1; after $\bm{A}$ is applied, it points up and to the right and is about three times longer.

```{r}
#| echo: false
#| dependson: matrixA
#| warning: false
#| fig-asp: 0.4
#| fig-width: 7
#| out-width: "80%"
rbind(v, c(A %*% v)) |>
  as_tibble() |>
  ggplot() +
  geom_segment(
    aes(x = 0, y = 0, xend = V1, yend = V2),
    arrow = arrow(length = unit(0.25, "cm")),
    linewidth = 0.8,
    col = discrete_colours[1:2]
  ) +
  annotate(
    "text",
    x = c(1, 3.0),
    y = c(-0.1, 1.1),
    label = c("bolditalic(v)", "bolditalic(Av)"),
    parse = TRUE,
    col = discrete_colours[1:2],
    family = "serif"
  ) +
  labs(x = "x", y = "y") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.5), minor_breaks = NULL) +
  coord_equal()
```

## Rank {#sec-rank}

The **rank** of a matrix measures how many genuinely independent rows (or equivalently, columns) it contains. A **rank-deficient** matrix has redundancy — some rows are duplicates or weighted sums of others; a **full-rank** matrix has no redundancy.

More formally, the rank is the number of linearly independent rows, which is always equal to the number of linearly independent columns. A set of rows is **linearly independent** when none of them can be written as a linear combination (i.e., a weighted sum) of the others. For an $m \times n$ matrix, the rank is at most $\min(m, n)$, and a matrix has full rank when this maximum is attained.

The rank of a matrix can be computed in R using `qr(<matrix>)$rank`.

```{r}
qr(A)$rank
```

For a rank-deficient example, the matrix $\begin{pmatrix} 1 & 2 \\ 2 & 4 \end{pmatrix}$ has rank 1, since its second row is twice its first row.

```{r}
qr(matrix(c(1, 2, 2, 4), 2, 2))$rank
```

## Matrix inverse {#sec-matrix-inverse}

The inverse $\bm{A}^{-1}$ of a matrix $\bm{A}$ is the matrix analogue of $1/x$ for a number: multiplying by it undoes what $\bm{A}$ did. Just as $1/x$ does not exist when $x = 0$, a matrix inverse does not always exist. When it does, $\bm{A}$ is called **invertible** (or **non-singular**); otherwise it is called **singular**. Only square, full-rank matrices have inverses; a rank-deficient matrix has lost information that the inverse would need to recover.

Formally, for a square matrix $\bm{A}$, the inverse $\bm{A}^{-1}$ is defined by $\bm{A}^{-1}\bm{A} = \bm{A}\bm{A}^{-1} = \bm{I}$, where $\bm{I}$ is the identity matrix.

The inverse of a matrix can be computed using the `solve` function in R.

```{r}
Ainv <- solve(A)
Ainv
Ainv %*% A
```

## Determinants {#sec-determinants}

The **determinant** of a square matrix measures how much the matrix stretches or shrinks space when applied to vectors. A determinant of zero means the matrix has collapsed at least one direction down to nothing — which is why such a matrix cannot be inverted: the information lost in the flattening cannot be recovered. So a square matrix is singular (equivalently, rank-deficient) if and only if its determinant is zero.

The determinant of a square matrix $\bm{A}$ is denoted by $|\bm{A}|$ or $\text{det}(\bm{A})$. The determinant of a $1 \times 1$ matrix is just the value of the element. For a $2 \times 2$ matrix,
$$
|\bm{A}| = a_{11}a_{22} - a_{12}a_{21},
$$
where $a_{ij}$ is the element in row $i$ and column $j$. For larger matrices, the formula is usually computed recursively using the Laplace expansion: for any choice of row $i$,
$$
|\bm{A}| = \sum_{j=1}^n (-1)^{i+j} a_{ij} |\bm{A}_{ij}|,
$$
where $\bm{A}_{ij}$ is the matrix obtained by deleting the $i$th row and $j$th column of $\bm{A}$.

The determinant of a matrix can be computed using the `det` function in R.

```{r}
det(A)
```

## Eigenvalues and eigenvectors {#sec-eigen}

Some matrices have special directions along which they only stretch or shrink vectors (possibly reversing them), without rotating them to point a different way. Any nonzero vector along such a direction is called an **eigenvector**, and the corresponding stretching factor is the **eigenvalue**. Formally, a nonzero vector $\bm{v}$ is an eigenvector of $\bm{A}$, with eigenvalue $\lambda$, if $\bm{A}\bm{v} = \lambda \bm{v}$. A square matrix has full rank (see @sec-rank) if and only if all of its eigenvalues are nonzero.

The eigenvalues and eigenvectors of a matrix can be computed using the `eigen` function in R.

```{r}
eigen(A)
```

Eigenvectors are essential tools in dimension reduction (@sec-highdim).

## Positive definite matrices {#sec-positive-definite}

Just as some numerical values must be positive (for example, a distance or a variance), or at least non-negative, sometimes we need something similar for matrices.

A **positive-definite** matrix is a special type of symmetric matrix where all the eigenvalues are positive. A positive-definite matrix is invertible, and its inverse is also positive definite.

A **positive semi-definite** matrix is a symmetric matrix where all the eigenvalues are non-negative. A positive semi-definite matrix is invertible if and only if it is positive definite.

Covariance matrices are used throughout this book, and they must always be at least positive semi-definite.

## Cholesky decomposition {#sec-cholesky}

A Cholesky decomposition is like a square root for a matrix. It is a decomposition of a positive semi-definite matrix $\bm{A}$ into a product $\bm{A} = \bm{L}\bm{L}^\intercal$, where $\bm{L}$ is a lower triangular matrix with non-negative diagonal entries. If $\bm{A}$ is positive definite, then the diagonal elements of $\bm{L}$ are positive.

Alternatively, it can be expressed using an upper triangular matrix, $\bm{U} = \bm{L}^\intercal$, so that $\bm{A} = \bm{U}^\intercal \bm{U}$. The `chol` function in R computes the upper triangular version of the Cholesky decomposition (although only for positive-definite matrices).

```{r}
U <- chol(A)
U
t(U) %*% U
```

Cholesky decompositions are particularly useful in multivariate scaling (@sec-scaling).
