---
code-fold: false
---

# Matrix algebra {#sec-appendix}

```{r}
#| label: setupA
#| include: false
#| cache: false

source("before-each-chapter.R")
```

Working with multivariate data inevitably involves writing some of the ideas using matrix expressions and matrix algebra.
In this appendix, a few of the results that we need are collected for convenient reference.
Readers who have not previously studied matrix algebra are recommended to do some self-study using a good introductory textbook such as @johnston2021introduction.

## Vectors {#sec-vectors}

A **vector** is a column of numbers, which we write as
$$
  \bm{v} = \begin{bmatrix}
             v_1 \\
             v_2 \\
             \vdots \\
             v_n
           \end{bmatrix} .
$$
The number of entries $n$ is its **dimension**.
Geometrically, we can picture a vector as a point in $n$-dimensional space, or equivalently as an arrow from the origin to that point.
This is how we view each observation in a data set: a vector with one entry per variable, and so a point in a space with one axis per variable.

The arrow image gives a vector two attributes: a **length** (how far the point lies from the origin) and a **direction** (the way the arrow points).
The length (also called the magnitude or **norm**), and written $\|\bm{v}\|$, is the straight-line distance from the origin to the point,
$$
  \|\bm{v}\| = \sqrt{v_1^2 + v_2^2 + \cdots + v_n^2}.
$$
For a two-dimensional vector this is just Pythagoras' theorem, and in general it is known as the "Euclidean distance".
Note that the length of a vector (its magnitude) is a different idea from its dimension (its number of entries); we reserve $\|\bm{v}\|$ and the word *length* for the magnitude.
The length of a vector is always non-negative, and is zero if and only if the vector is the zero vector $\bm{0} = (0, \ldots, 0)^\intercal$.

A vector with length one, $\|\bm{v}\| = 1$, is called a **unit vector**.
Any nonzero vector can be turned into a unit vector pointing the same way by dividing it by its length, $\bm{v}/\|\bm{v}\|$, a step called **normalising**.

```{r}
#| label: vector

v <- c(3, 4)
sqrt(sum(v^2)) # length of v
v / sqrt(sum(v^2)) # unit vector in the direction of v
```

## Matrices {#sec-matrices}

A **matrix** is a rectangular array of numbers, denoted by $\bm{A} = (a_{ij})$, where $a_{ij}$ is the entry in row $i$ and column $j$.
The number of rows $m$ and columns $n$ are called the **dimensions** of the matrix, and we say that $\bm{A}$ is an $m \times n$ matrix.
A matrix with the same number of rows and columns, i.e. with dimensions $n \times n$, is called a **square matrix**.
A vector is a special case of a matrix, with one column.
A matrix with one row is called a **row vector**.

Two matrices $\bm{A}$ and $\bm{B}$ can be added together if they are of the same dimensions, and the sum $\bm{A} + \bm{B}$ is the matrix of the same dimensions whose $(i,j)$th entry is $a_{ij} + b_{ij}$.

A matrix can also be multiplied by another matrix if the number of columns of $\bm{A}$ matches the number of rows of $\bm{B}$.
If $\bm{A}$ is an $m \times p$ matrix and $\bm{B}$ is a $p \times n$ matrix, the product $\bm{A}\bm{B}$ is the $m \times n$ matrix with $(i,j)$th element given by
$$
  (\bm{A}\bm{B})_{ij} = \sum_{k=1}^p a_{ik} b_{kj}.
$$

In R, matrix multiplication uses the `%*%` operator; the ordinary `*` operator instead multiplies matrices element by element, which is a different operation.
For example:

```{r}
#| label: matrix

A <- matrix(c(3, 1, 1, 2), 2, 2)
B <- matrix(seq(6), nrow = 3)
A
B
B %*% A
```

Scalar multiplication of a matrix $\bm{A}$ by a number $c$ is defined by multiplying each entry of $\bm{A}$ by $c$, and is denoted by $c\bm{A}$.
Similarly, scalar addition of a matrix $\bm{A}$ by a number $c$ is defined by adding $c$ to each entry of $\bm{A}$, and is denoted by $\bm{A} + c$.

## Transpose {#sec-transpose}

The **transpose** of a matrix $\bm{A}$, written $\bm{A}^\intercal$, is obtained by swapping its rows and columns: the entry in row $i$ and column $j$ of $\bm{A}$ becomes the entry in row $j$ and column $i$ of $\bm{A}^\intercal$.
So an $m \times n$ matrix becomes $n \times m$.
Transposing a single-column vector turns it into a single row, which is why we write a column vector compactly as $\bm{v} = (v_1, \ldots, v_n)^\intercal$ (@sec-vectors).

A square matrix that is unchanged by transposing, so that $\bm{A} = \bm{A}^\intercal$ (equivalently, $a_{ij} = a_{ji}$), is called **symmetric**: its entries are mirror images of each other across the diagonal.

The transpose is computed in R using the `t` function.

```{r}
#| label: transpose
#| dependson: matrix

t(B)
```

## Identity matrix {#sec-identity-matrix}

The identity matrix is like the number 1: multiplying a matrix by the identity matrix leaves the matrix unchanged.
The identity matrix is a square matrix with ones on the diagonal and zeros elsewhere, denoted by $\bm{I}$.
Sometimes we will use subscripts to denote the size of the identity matrix; e.g. $\bm{I}_n$ is an $n \times n$ identity matrix.

```{r}
#| label: identity

I3 <- diag(3)
I3
```

## Matrices and directions {#sec-matrices-directions}

A matrix can be applied to a vector $\bm{v}$ by matrix multiplication, producing a new vector $\bm{A}\bm{v}$.
Geometrically, this transforms $\bm{v}$, generally changing both its length and the direction it points.
Different starting vectors are typically transformed in different ways: a matrix may stretch some directions, shrink or rotate others, or even collapse two distinct directions onto the same output.

```{r}
#| label: matrixA
#| dependson: matrix

v <- c(1, 0)
A %*% v
```

Geometrically, $\bm{v}$ originally pointed along the $x$-axis with length 1; after $\bm{A}$ is applied, it points up and to the right and is more than three times longer.

```{r}
#| label: vector-transform-plot
#| echo: false
#| dependson: matrixA
#| warning: false
#| fig-asp: 0.4
#| fig-width: 7
#| out-width: "80%"

rbind(v, c(A %*% v)) |>
  as_tibble() |>
  ggplot() +
  geom_segment(
    aes(x = 0, y = 0, xend = V1, yend = V2),
    arrow = arrow(length = unit(0.25, "cm")),
    linewidth = 0.8,
    col = discrete_colours[1:2]
  ) +
  annotate(
    "text",
    x = c(1, 3.0),
    y = c(-0.1, 1.1),
    label = c("bolditalic(v)", "bolditalic(Av)"),
    parse = TRUE,
    col = discrete_colours[1:2],
    family = "serif"
  ) +
  labs(x = "x", y = "y") +
  scale_y_continuous(breaks = seq(0, 1, by = 0.5), minor_breaks = NULL) +
  coord_equal()
```

## Diagonal matrices {#sec-diagonal-matrices}

Matrices that are zeros everywhere apart from the diagonal elements are called **diagonal matrices**.
They are denoted by $\text{diag}(d_1, \ldots, d_n)$, where $d_i$ is the value of the $i$th diagonal element.
Diagonal matrices are easy to work with because they only stretch or shrink vectors along the coordinate axes, without rotating them.
The product of a diagonal matrix and a vector is just the elementwise product of the diagonal entries and the vector entries.

In R, a diagonal matrix can be created using the `diag` function.

```{r}
#| label: diagonal

diag(c(2, 3, 4))
```

Notice that we also used the `diag()` function to create a square identity matrix in @sec-identity-matrix.
The `diag` function is overloaded to do different things depending on the type of its argument: when given a single number, it creates an identity matrix of that size; when given a vector, it creates a diagonal matrix with those values on the diagonal.
A third use is when given a matrix, it extracts the diagonal elements as a vector.

```{r}
#| label: diagonal-extract
#| dependson: matrix

diag(A)
```

## Rank {#sec-rank}

The **rank** of a matrix measures how many genuinely independent rows (or equivalently, columns) it contains.
A **rank-deficient** matrix has redundancy --- some rows are duplicates or weighted sums of others; a **full-rank** matrix has no redundancy.

More formally, the rank is the number of linearly independent rows, which is always equal to the number of linearly independent columns.
A set of rows is **linearly independent** when none of them can be written as a linear combination (i.e., a weighted sum) of the others.
For an $m \times n$ matrix, the rank is at most $\min(m, n)$, and a matrix has full rank when this maximum is attained.

The rank of a matrix can be computed in R using `qr(<matrix>)$rank`.

```{r}
#| label: rank
#| dependson: matrix

qr(A)$rank
```

For a rank-deficient example, the matrix $\begin{pmatrix} 1 & 2 \\ 2 & 4 \end{pmatrix}$ has rank 1, since its second row is twice its first row.

```{r}
#| label: rank-deficient

qr(matrix(c(1, 2, 2, 4), 2, 2))$rank
```

## Matrix inverse {#sec-matrix-inverse}

The inverse $\bm{A}^{-1}$ of a matrix $\bm{A}$ is the matrix analogue of $1/x$ for a number: multiplying by it undoes what $\bm{A}$ did.
Just as $1/x$ does not exist when $x = 0$, a matrix inverse does not always exist.
When it does, $\bm{A}$ is called **invertible** (or **non-singular**); otherwise it is called **singular**.
Only square, full-rank matrices have inverses; a rank-deficient matrix has lost information that the inverse would need to recover.

Formally, for a square matrix $\bm{A}$, the inverse $\bm{A}^{-1}$ is defined by $\bm{A}^{-1}\bm{A} = \bm{A}\bm{A}^{-1} = \bm{I}$, where $\bm{I}$ is the identity matrix.

The inverse of a matrix can be computed using the `solve` function in R.

```{r}
#| label: inverse
#| dependson: matrix

Ainv <- solve(A)
Ainv
Ainv %*% A
```

## Determinants {#sec-determinants}

The **determinant** of a square matrix measures how much the matrix stretches or shrinks space when applied to vectors.
A determinant of zero means the matrix has collapsed at least one direction down to nothing --- which is why such a matrix cannot be inverted: the information lost in the flattening cannot be recovered.
So a square matrix is singular (equivalently, rank-deficient) if and only if its determinant is zero.

The determinant of a square matrix $\bm{A}$ is denoted by $|\bm{A}|$ or $\text{det}(\bm{A})$.
The determinant of a $1 \times 1$ matrix is just the value of the element.
For a $2 \times 2$ matrix,
$$
  |\bm{A}| = a_{11}a_{22} - a_{12}a_{21},
$$
where $a_{ij}$ is the element in row $i$ and column $j$.
For larger matrices, the formula is usually computed recursively using the Laplace expansion: for any choice of row $i$,
$$
  |\bm{A}| = \sum_{j=1}^n (-1)^{i+j} a_{ij} |\bm{A}_{ij}|,
$$
where $\bm{A}_{ij}$ is the matrix obtained by deleting the $i$th row and $j$th column of $\bm{A}$.

The determinant of a matrix can be computed using the `det` function in R.

```{r}
#| label: determinant
#| dependson: matrix

det(A)
```

## Eigenvalues and eigenvectors {#sec-eigen}

Some matrices have special directions along which they only stretch or shrink vectors (possibly reversing them), without rotating them to point a different way.
Any nonzero vector along such a direction is called an **eigenvector**, and the corresponding stretching factor is the **eigenvalue**.
Formally, a nonzero vector $\bm{v}$ is an eigenvector of $\bm{A}$, with eigenvalue $\lambda$, if $\bm{A}\bm{v} = \lambda \bm{v}$.
A square matrix has full rank (see @sec-rank) if and only if all of its eigenvalues are nonzero.

When $\bm{A}$ is symmetric, its eigenvectors can be chosen to be **orthonormal**, meaning that each one has length one and every pair is **orthogonal** --- at right angles to one another, so that $\bm{v}_i^\intercal\bm{v}_j = 0$ whenever $i \ne j$.
Collecting them as the columns of a matrix $\bm{V}$, and placing the corresponding eigenvalues on the diagonal of $\bm{\Lambda} = \text{diag}(\lambda_1, \ldots, \lambda_n)$, the relationship $\bm{A}\bm{v} = \lambda\bm{v}$ holds for all the eigenvectors at once as $\bm{A}\bm{V} = \bm{V}\bm{\Lambda}$.
Because the columns of $\bm{V}$ are orthonormal, $\bm{V}^\intercal\bm{V} = \bm{V}\bm{V}^\intercal = \bm{I}$ (so that $\bm{V}^{-1} = \bm{V}^\intercal$), and this rearranges into the **eigendecomposition** (or **spectral decomposition**)
$$
  \bm{A} = \bm{V}\bm{\Lambda}\bm{V}^\intercal,
$$
which expresses $\bm{A}$ entirely in terms of its eigenvalues and eigenvectors.

The `eigen()` function returns $\bm{V}$ in `$vectors` and the diagonal entries of $\bm{\Lambda}$ in `$values`, so the decomposition can be reconstructed directly.

```{r}
#| label: eigen
#| dependson: matrix

e <- eigen(A)
e
e$vectors %*% diag(e$values) %*% t(e$vectors)
```

Eigendecomposition is an essential tool in dimension reduction (@sec-highdim).

## Positive definite matrices {#sec-positive-definite}

Just as some numerical values must be positive (for example, a distance or a variance), or at least non-negative, sometimes we need something similar for matrices.

A **positive-definite** matrix is a special type of symmetric matrix where all the eigenvalues are positive.
A positive-definite matrix is invertible, and its inverse is also positive definite.

A **positive semi-definite** matrix is a symmetric matrix where all the eigenvalues are non-negative.
A positive semi-definite matrix is invertible if and only if it is positive definite.

Covariance matrices are used throughout this book, and they must always be at least positive semi-definite.

## Cholesky decomposition {#sec-cholesky}

A Cholesky decomposition is like a square root for a matrix.
It is a decomposition of a positive semi-definite matrix $\bm{A}$ into a product $\bm{A} = \bm{L}\bm{L}^\intercal$, where $\bm{L}$ is a **lower triangular** matrix (one whose entries above the diagonal are all zero) with non-negative diagonal entries.
If $\bm{A}$ is positive definite, then the diagonal elements of $\bm{L}$ are positive.

Alternatively, it can be expressed using an **upper triangular** matrix (one whose entries below the diagonal are all zero), $\bm{U} = \bm{L}^\intercal$, so that $\bm{A} = \bm{U}^\intercal \bm{U}$.
The `chol` function in R computes the upper triangular version of the Cholesky decomposition (although only for positive-definite matrices).

```{r}
#| label: cholesky
#| dependson: matrix

U <- chol(A)
U
t(U) %*% U
```
